---
title: "Import samples"
author: "Abed Kurdi and David Wu"
---

Set working directory to project directory
```{r setup}
require(knitr)
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

### Optionally convert this Rmd to R script 
```{r eval=FALSE, message=FALSE}
current_file <- rstudioapi::getActiveDocumentContext()$path
output_file <- stringr::str_replace(current_file, '.Rmd', '.R')
knitr::purl(current_file, output = output_file)
file.edit(output_file)
```

### Directories
```{r, message=FALSE}
output_dir <- 'analysis/output/01_import' # analysis file output directory
data_dir <- 'data/derived/pipestance' # data file output directory

dir.create(output_dir, showWarnings = FALSE)
dir.create(data_dir, showWarnings = FALSE)
```

### Loading the libraries
```{r, message=FALSE}
library(Seurat) 
library(tidyverse)
library(scDblFinder) 
```


```{r, message=FALSE}
# library(metap) 
# library(MAST) 
# 
# library(ggplot2) 
# library(cowplot) 
# library(patchwork) 
# library(writexl)
# library(dplyr)
# library(xlsx)
# library(harmony)
# library(openxlsx)
# library(rgl)
# library(ggExtra)
# library(ggplotify)
```

### Specifying the input and output channels
```{r}
samples_dir <- "analysis/input/samples/"
metadata_df <- read_tsv('analysis/output/00_process/metadata_subset.tsv')
stress_genes <- read_csv('analysis/input/coregene_df-FALSE-v3.csv')
dir.create(file.path(output_dir, 'qc_plots', 'scatter_plot_per_sample'), recursive = TRUE)
```

### Making the needed empty lists and functions
```{r}
#setwd(samples_dir)
all_samples <- list()
all_samples_data <- list()
all_samples_sce <- list()

#function to trim ADT:
trim.feature.names <- function(inmat){
  newnames <- sapply(strsplit(rownames(inmat), "_"),
                     function(x) {
                       if(length(x)==1) return(x)
                       else return(paste(x[-length(x)], collapse="_"))
                     })
  rownames(inmat) <- newnames
  return(inmat)
}
```

### read in the samples, make the Seurat objects, remove doublets and plot featurescatter figures
```{r echo=FALSE}
for(f in metadata_df$id){
	print(f)

	# read in the samples and remove doublets
	all_samples_data[[f]] <- Read10X_h5(paste0(samples_dir, f, '.h5'))

	if(!is.null(names(all_samples_data[[f]]))){
		all_samples_data[[f]][["Gene Expression"]] <- CollapseSpeciesExpressionMatrix(all_samples_data[[f]][["Gene Expression"]], prefix = "GRCh38_", controls = "mm10___", ncontrols = 200)
		all_samples_data[[f]][["Antibody Capture"]] <- trim.feature.names(all_samples_data[[f]][["Antibody Capture"]])	
		all_samples[[f]] <- CreateSeuratObject(counts = all_samples_data[[f]][["Gene Expression"]], project = as.character(f))
	} else {
		all_samples[[f]] <- CreateSeuratObject(counts = all_samples_data[[f]])
	}

	all_samples_sce[[f]] <- as.SingleCellExperiment(all_samples[[f]])
	all_samples_sce[[f]] <- scDblFinder(all_samples_sce[[f]])
	print(table(call=all_samples_sce[[f]]$scDblFinder.class))

	#DefaultAssay(all_samples[[f]]) <- "RNA"

	all_samples[[f]] <- as.Seurat(all_samples_sce[[f]], project=f)	
	all_samples[[f]][["percent.mt"]] <- PercentageFeatureSet(all_samples[[f]], pattern="^MT-", assay="RNA")
	all_samples[[f]][["percent.ribo"]] <- PercentageFeatureSet(all_samples[[f]], pattern="^RPS|^RPL|^MRPS|^MRPL", assay="RNA")
	all_samples[[f]][["percent.stress"]] <- PercentageFeatureSet(all_samples[[f]], features=intersect(stress_genes$gene_symbol, rownames(all_samples[[f]])), assay="RNA")
	
	if(!is.null(names(all_samples_data[[f]]))){
		all_samples[[f]][["ADT"]] <- CreateAssayObject(counts = all_samples_data[[f]][["Antibody Capture"]])
	}

	all_samples[[f]] <- subset(all_samples[[f]], scDblFinder.class == "singlet")
	
	# add the metadata
	tmp_df <- subset(metadata_df, id == str_remove(f, '.h5'))
	print(paste0("adding metadata for sample ",f))

	for(m in colnames(tmp_df)){
		print(paste0("adding ",m))
		all_samples[[f]][[m]] <- tmp_df[[m]]
	}
	
	## scatterplot
	pdf(paste0(output_dir,"/qc_plots/scatter_plot_per_sample/",f,".pdf"), width=14)
	options(repr.plot.height = 6, repr.plot.width = 15)
	p1 <- FeatureScatter(all_samples[[f]], feature1 = "nCount_RNA", feature2 = "percent.mt") + ggtitle(f)+ geom_point(shape=16, color="blue", alpha = 0.1) + 
		geom_hline(yintercept=15, linetype="dashed", color = "black", size=1.5) + geom_vline(xintercept = 10000, linetype="dashed", color = "black", size=1.5)

	p2 <- FeatureScatter(all_samples[[f]], feature1 = "nCount_RNA", feature2 = "nFeature_RNA") + ggtitle(f)+ geom_point(shape=16, color="blue", alpha = 0.1)+
		geom_hline(yintercept=2500, linetype="dashed", color = "black", size=1.5) + geom_vline(xintercept = 10000, linetype="dashed", color = "black", size=1.5)

	print(p1+p2)

	### histograph
	options(repr.plot.height = 6, repr.plot.width = 24)
	plot1 <-  hist(all_samples[[f]]$nCount_RNA, breaks = 100, main=f, xlab = "Count depth (Counts/cell)", col='wheat')
	plot1 <-  abline(v=400,col="red")
	plot2 <-  hist(all_samples[[f]]$nFeature_RNA, breaks = 200, main=f, xlab = "Number of genes per cell", xlim = c(0, 4000), col='wheat')
	plot2 <-  abline(v=250,col="red")

	print(plot1 + plot2)

	dev.off()
}
```

### plot the mito/ribo/stress percentages correlation plots
```{r}
pdf(paste0(output_dir,"/qc_plots/mito_ribo_stress_qc_final.pdf"), width=14, height=6)
for(s in names(all_samples)){
  mt_ribo_stress <- FetchData(object = all_samples[[s]], vars = c("percent.mt", "percent.stress","percent.ribo"))
  
  p1 <- ggplot(mt_ribo_stress, aes(x=percent.ribo, y=percent.mt))+geom_point(alpha=0.25, size=0.5)+ylim(0,100)+xlim(0,100)+ggtitle(s)
  p2 <- ggplot(mt_ribo_stress, aes(x=percent.stress, y=percent.mt))+geom_point(alpha=0.25, size=0.5)+ylim(0,100)+xlim(0,100)+ggtitle(s)
  
  # with marginal histogram
  p11 <- ggMarginal(p1, type="histogram")
  p21 <- ggMarginal(p2, type="histogram")
  print(as.ggplot(p11) + as.ggplot(p21))
  
}
dev.off()
```

### according to the upper plots, I set the thresholds
#### applying filters: nFeatures > 100 and < 6000, percent.mt < 20, percent.ribo < 50, and percent.stress < 30
```{r}
for(f in names(all_samples)){
	all_samples[[f]] <- subset(all_samples[[f]], 
		subset = nFeature_RNA > 100 & 
		nFeature_RNA < 6000 & 
		percent.mt < 20 & 
		percent.ribo < 50 & 
		percent.stress < 30)
}
```

### merge all samples in one object
#### renaming the cells
```{r}
for(s in names(all_samples)){
    print(s)
    all_samples[[s]] <- RenameCells(all_samples[[s]], new.names=paste0(colnames(all_samples[[s]]),"_",s))
}
```

#### merge all
```{r}
merged_samples <- all_samples[[1]]
for (i in 2:length(all_samples)) {
  merged_samples <- merge(merged_samples, y = all_samples[[i]], add.cell.ids = NULL)
}
```
____

### Normalize, FindVariableFeatures, ScaleData and RunPCA
```{r}
merged_samples <- Seurat::NormalizeData(merged_samples, assay="RNA") %>% 
  FindVariableFeatures(selection.method = "vst", nfeatures = 2000, assay="RNA") %>%
  ScaleData(verbose = FALSE, assay="RNA") %>%
  RunPCA(pc.genes = merged_samples@var.genes, npcs = 30, verbose = FALSE, assay="RNA")
```

### Run Harmony
```{r}
merged_samples <- merged_samples %>% 
    RunHarmony(c("treatment", "kit"), plot_convergence = TRUE, assay.use='RNA')
```

### Make the elbow plot
```{r}
pdf(paste0(output_dir,"qc_plots/elbow_plot.pdf"))
ElbowPlot(merged_samples, ndims=30)
dev.off()
```

### RunUMAP, FindNeighbors and FindClusters using resolutions starting from 0.1 to 2, by 0.1
```{r}
merged_samples <- merged_samples %>% 
    RunUMAP(reduction = "harmony", dims = 1:20, assay="RNA") %>% 
    FindNeighbors(reduction = "harmony", dims = 1:20, assay="RNA") %>% 
    FindClusters(resolution = seq(0.1,2,by=0.1), graph.name = "RNA_snn") %>% 
    identity()
```

### Save the "merged_samples" object
```{r}
saveRDS(merged_samples, file=paste0(output_dir,"RData/merged_all_samples_filtered_with_metadata_with_findClusters.RDS"))
```
_____

### Get the markers, UMAP and DotPlot for the marker genes
### Resolution of 0.2
```{r}
Idents(merged_samples) <- "RNA_snn_res.0.2"
clusters_identified <- sort(unique(unfactor(Idents(object=merged_samples))))

OUT <- createWorkbook()
for(i in clusters_identified){
    tryCatch({
	    cat(paste0("Finding Markers for cluster ",i,"\n"))
	    markers <- FindMarkers(merged_samples, ident.1 = i, 
	      verbose = FALSE, test.use="MAST", min.pct = 0, logfc.threshold = 0.25) %>% mutate(., gene = rownames(.))

		addWorksheet(OUT, paste0("cluster_",i))
		writeData(OUT, sheet = paste0("cluster_",i), x = markers)

      cat(paste0("Finished Finding Markers for cluster ",i,"\n"))
      }, error=function(e){message(paste0("We have an error here in cluster ", i))})
}

saveWorkbook(OUT, paste0(output_dir,"markers/DEG_per_cluster_res_0.2.xlsx"))
```

#### UMAP
```{r}
pdf(paste0(output_dir,"umap/umap_at_resolution_0.2.pdf"), width=10, height=8)
DimPlot(merged_samples, reduction = "umap", label = TRUE, pt.size = .1)+labs(title="Resolution 0.2")
dev.off()
```

#### Dot Plot
```{r}
pdf(paste0(output_dir,"dot_plot/dot_plot_marker_genes_at_res_0.2_no_limits_defined.pdf"), width=18, height=8)
DotPlot(merged_samples, features=unique(c("CD3D","CCR7","SELL","KLF2","CD69","ITGAE","CXCR6","CD4","TIGIT","FOXP3",
  "IL2RA","CTLA4","CD8A","CD8B","GZMB","PDCD1","LAG3","KLRB1","PRF1","KLRD1","GNLY",
  "TNFRSF18","PRDM1","BATF","TRAT1","RORA","GATA3","PTGDR2","IL7R","HLA-DRA","HLA-DRB1",
  "CD83","IDO1","CD207","EPCAM","CD68","C1QB","C1QC","CD163","CLEC10A","CD1C","CLEC10A",
  "THBD","XCR1","SIRPA","F13A1","IGKC","JCHAIN","CD79A","MS4A1","NR4A1","NR4A2","KLF4",
  "CEBPB","LYZ","MS4A7","SERPINA1","CD14","S100A9","IL23A","TPSB2","TPSAB1","MKI67","TOP2A",
  "KLRB1","IL7R","ITGA4","NCR1","KLRB1C","IL17A","IL17F","IL23R")), cols=c("#5F4B8BFF", "#ED2B33FF"), cluster.idents=TRUE)+
  RotatedAxis()
dev.off()
```

### Resolution of 1
```{r}
Idents(merged_samples) <- "RNA_snn_res.1"
clusters_identified <- sort(unique(unfactor(Idents(object=merged_samples))))

OUT <- createWorkbook()
for(i in clusters_identified){
    tryCatch({
	    cat(paste0("Finding Markers for cluster ",i,"\n"))
	    markers <- FindMarkers(merged_samples, ident.1 = i, 
	      verbose = FALSE, test.use="MAST", min.pct = 0, logfc.threshold = 0.25) %>% mutate(., gene = rownames(.))

		addWorksheet(OUT, paste0("cluster_",i))
		writeData(OUT, sheet = paste0("cluster_",i), x = markers)

      cat(paste0("Finished Finding Markers for cluster ",i,"\n"))
      }, error=function(e){message(paste0("We have an error here in cluster ", i))})
}

saveWorkbook(OUT, paste0(output_dir,"markers/DEG_per_cluster_res_1.xlsx"))
```

#### UMAP
```{r}
pdf(paste0(output_dir,"umap/umap_at_resolution_1.pdf"), width=10, height=8)
DimPlot(merged_samples, reduction = "umap", label = TRUE, pt.size = .1)+labs(title="Resolution 1")
dev.off()
```

#### Dot Plot
```{r}
pdf(paste0(output_dir,"/dot_plot/dot_plot_marker_genes_at_res_1_no_limits_defined.pdf"), width=18, height=8)
DotPlot(merged_samples, features=unique(c("CD3D","CCR7","SELL","KLF2","CD69","ITGAE","CXCR6","CD4","TIGIT","FOXP3",
  "IL2RA","CTLA4","CD8A","CD8B","GZMB","PDCD1","LAG3","KLRB1","PRF1","KLRD1","GNLY",
  "TNFRSF18","PRDM1","BATF","TRAT1","RORA","GATA3","PTGDR2","IL7R","HLA-DRA","HLA-DRB1",
  "CD83","IDO1","CD207","EPCAM","CD68","C1QB","C1QC","CD163","CLEC10A","CD1C","CLEC10A",
  "THBD","XCR1","SIRPA","F13A1","IGKC","JCHAIN","CD79A","MS4A1","NR4A1","NR4A2","KLF4",
  "CEBPB","LYZ","MS4A7","SERPINA1","CD14","S100A9","IL23A","TPSB2","TPSAB1","MKI67","TOP2A",
  "KLRB1","IL7R","ITGA4","NCR1","KLRB1C","IL17A","IL17F","IL23R")), cols=c("#5F4B8BFF", "#ED2B33FF"), cluster.idents=TRUE)+
  RotatedAxis()
dev.off()
```
