---
title: "ADT Analysis"
author: "David Wu"
---

### Use the here package for robust relative paths
```{r}
library(here)
```

### Optionally convert this Rmd to R script 
```{r eval=FALSE, message=FALSE}
current_file <- rstudioapi::getActiveDocumentContext()$path
output_file <- stringr::str_replace(current_file, '.Rmd', '.R')
knitr::purl(current_file, output = output_file)
file.edit(output_file)
```

### Directories
```{r, message=FALSE}
output_dir <- here('analysis/output/03_adt') # analysis file output directory
data_dir <- here('data/derived/tildra') # data file output directory

dir.create(output_dir, showWarnings = FALSE)
dir.create(data_dir, showWarnings = FALSE)
```

### Load libraries
```{r, message=FALSE}
library(Seurat) 
library(tidyverse)
library(ggExtra)
library(harmony)
library(SeuratWrappers)
library(wutilities) # devtools::install_github('symbiologist/wutilities')
library(tictoc)
theme_set(theme_dwu()) # set default theme
```

### Load Seurat object
```{r}
seuratobj <- read_rds(here(data_dir, 'seuratobj_qc.rds'))
metadata <- seuratobj@meta.data %>% rownames_to_column() %>% as_tibble()
seuratobj
```
### Analysis
#### ADT colnames
```{r}
adt_markers <- seuratobj@assays$ADT %>% rownames()
adt_markers
```
### Normalize by CLR method (margin 2 per)
```{r}
seuratobj <- seuratobj %>% NormalizeData(assay = 'ADT', normalization.method = 'CLR', margin = 2)
```

```{r}
VlnPlot(seuratobj, features = c('ADT_CD4'), group.by = 'id', pt.size = 0.1) + theme(legend.position = 'none')
```

### Extract raw ADT data from ADT samples with mouse-spike
```{r}
adt_raw <- FetchData(seuratobj, vars = c('id', 'mouse', 'species', paste0('ADT_', adt_markers)), slot = 'count') %>% 
  rownames_to_column() %>% 
  filter(mouse == 'Yes') %>% 
  set_names(str_remove(colnames(.), 'ADT_')) %>% 
  as_tibble() %>% 
  select(-mouse)

adt_clr <- FetchData(seuratobj, vars = c('id', 'mouse', 'species', paste0('ADT_', adt_markers)), slot = 'data') %>% 
  rownames_to_column() %>% 
  filter(mouse == 'Yes') %>% 
  set_names(str_remove(colnames(.), 'ADT_')) %>% 
  as_tibble() %>% 
  select(-mouse)
  
  
adt_clr_long <- adt_clr %>% 
  pivot_longer(cols = all_of(adt_markers),
               names_to = 'marker',
               values_to = 'expression') 

adt_raw %>% head()
```



#### Which samples have mouse cells to set background?
```{r}
metadata %>% 
  filter(species == 'Mouse' | adt == 'Yes') %>% 
  select(id, sample, mouse, adt) %>% 
  unique()
```
```{r}
adt_samples <- metadata %>% filter(mouse == 'Yes') %>% pull(id) %>% unique()
```


```{r}
plot_dir <- here(output_dir, 'density_clr2')

adt_results_clr2 <- map(adt_markers, function(i) {
  
  sample_subset <- adt_clr_long %>% 
    filter(marker == i)
  
  mouse_summary <- sample_subset %>% 
    filter(species == 'Mouse') %>% 
    group_by(id) %>% 
    summarize(mean = mean(expression),
              sd = sd(expression),
              threshold1 = mean + sd,
              threshold2 = mean + 2*sd)
    
  p <- sample_subset %>% 
    left_join(mouse_summary) %>% 
    ggplot(aes(x = expression,
               fill = species)) +
    geom_density(alpha = 0.5) +
    geom_vline(aes(xintercept = threshold1), linetype = 'dashed', color = 'red') +
    geom_vline(aes(xintercept = threshold2), linetype = 'dashed', color = 'red') +
    rcartocolor::scale_fill_carto_d() +
    facet_wrap(~id, scales = 'free') +
    theme(legend.position = 'top') +
    labs(x = 'Normalized Expression',
         y = '# of Cells',
         title = i)
  
  ggsave(plot = p,
         filename = paste0(i, '.png'),
         path = plot_dir,
         h = 12,
         w = 12)
  
  sample_subset %>% 
    left_join(mouse_summary)
  
}) %>% bind_rows()


```
### CLR by margin 1
```{r}

seuratobj1 <- seuratobj %>% NormalizeData(assay = 'ADT', normalization.method = 'CLR', margin = 1)

adt_clr <- FetchData(seuratobj1, vars = c('id', 'mouse', 'species', paste0('ADT_', adt_markers)), slot = 'data') %>% 
  rownames_to_column() %>% 
  filter(mouse == 'Yes') %>% 
  set_names(str_remove(colnames(.), 'ADT_')) %>% 
  as_tibble() %>% 
  select(-mouse)
  
  
adt_clr_long <- adt_clr %>% 
  pivot_longer(cols = all_of(adt_markers),
               names_to = 'marker',
               values_to = 'expression') 

plot_dir <- here(output_dir, 'density_clr1')

adt_results_clr1 <- map(adt_markers, function(i) {
  
  sample_subset <- adt_clr_long %>% 
    filter(marker == i)
  
  mouse_summary <- sample_subset %>% 
    filter(species == 'Mouse') %>% 
    group_by(id) %>% 
    summarize(mean = mean(expression),
              sd = sd(expression),
              threshold1 = mean + sd,
              threshold2 = mean + 2*sd)
    
  p <- sample_subset %>% 
    left_join(mouse_summary) %>% 
    ggplot(aes(x = expression,
               fill = species)) +
    geom_density(alpha = 0.5) +
    geom_vline(aes(xintercept = threshold1), linetype = 'dashed', color = 'red') +
    geom_vline(aes(xintercept = threshold2), linetype = 'dashed', color = 'red') +
    rcartocolor::scale_fill_carto_d() +
    facet_wrap(~id, scales = 'free') +
    theme(legend.position = 'top') +
    labs(x = 'Normalized Expression',
         y = '# of Cells',
         title = i)
  
  ggsave(plot = p,
         filename = paste0(i, '.png'),
         path = plot_dir,
         h = 12,
         w = 12)
  
  sample_subset %>% 
    left_join(mouse_summary)
  
}) %>% bind_rows()
```

```{r}
robust_markers <- c('CD1c', 
                    'CD3',
                    'CD4', 
                    'CD8',
                    'CD11c',
                    'CD69',
                    'CD127')
```

### ADTNorm Method of CITE-Seq normalization
https://yezhengstat.github.io/ADTnorm/index.html
Not working
```{r eval=FALSE}
library(ADTnorm)

adtnorm_input <- adt_raw #%>% filter(id %in% c('S195','S1981'))


cell_x_adt <- adtnorm_input %>% select(all_of(robust_markers))
cell_x_feature <- adtnorm_input %>% select(sample = id,
                                           batch = id)
# cell_x_feature$sample = factor(cell_x_feature$study_name)
# cell_x_feature$batch = factor(cell_x_feature$study_name)

adt_results_norm <- ADTnorm(
  cell_x_adt = cell_x_adt, 
  cell_x_feature = cell_x_feature,
  save_outpath = output_dir, 
  study_name = 'adt_norm', 
  marker_to_process = robust_markers, 
  save_intermediate_fig = TRUE
)
```

### Transform
```{r}
adt_calls <- adt_results_clr2 %>% 
  filter(marker %in% robust_markers,
         species == 'Human') %>% 
  mutate('result' = ifelse(expression > threshold2, 'Positive', 'Negative')) %>% 
  select(rowname, marker, result) %>% 
  pivot_wider(names_from = 'marker',
              values_from = 'result')

adt_calls  
```

## Export
```{r}
adt_results_clr2 %>% 
  filter(marker %in% robust_markers) %>% 
    write_tsv(here(output_dir, 'adt_results_clr2.tsv.gz'))

adt_results_norm %>% write_rds(here(output_dir, 'adt_results_norm.rds'), compress = 'gz')

adt_calls %>% write_tsv(here(output_dir, 'adt_calls.tsv.gz'))
```

## Session Info
```{r}
sessionInfo()
```



