---
title: "Subclustering"
author: "David Wu"
---

Set working directory to project directory
```{r setup}
require(knitr)
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

### Optionally convert this Rmd to R script 
```{r eval=FALSE, message=FALSE}
current_file <- rstudioapi::getActiveDocumentContext()$path
output_file <- stringr::str_replace(current_file, '.Rmd', '.R')
knitr::purl(current_file, output = output_file)
file.edit(output_file)
```

### Directories
```{r, message=FALSE}
output_dir <- 'analysis/output/03_subcluster' # analysis file output directory
data_dir <- 'data/derived/seurat' # data file output directory

dir.create(output_dir, showWarnings = FALSE)
dir.create(data_dir, showWarnings = FALSE)
```

### Loading the libraries
```{r, message=FALSE}
library(Seurat) 
library(tidyverse)
library(ggExtra)
library(SeuratWrappers)
library(wutilities) # devtools::install_github('symbiologist/wutilities')
library(tictoc)
library(harmony)
library(Nebulosa)
```

```{r}
seuratobj <- read_rds(file.path(data_dir, 'seuratobj_clustered.rds'))
seuratobj
```

### Visualize
```{r fig.height = 4, fig.width=4}
p <- seurat_feature(seuratobj, features = c('RNA_snn_res.0.2'), facet_hide = TRUE, legend_position = 'none', title = 'Resolution: 0.2')

p
```

```{r}
seurat_feature(seuratobj, feature = 'RNA_snn_res.0.2', facets = 'RNA_snn_res.0.2', label = FALSE)
```
```{r fig.height=10, fig.width=12}
seurat_feature(seuratobj, feature = 'RNA_snn_res.0.4', facets = 'RNA_snn_res.0.4', label = FALSE)
```



### Reynolds et al markers

```{r}
markers_reynolds <- c('KRT14', 'KRT5', 
                      'GATA3', 'KRTDAP',
                      'CRYAB', 'TYRP1',
                      'DKK3', 'PLEKHA4',
                      'LUM', 'APOD',
                      'PHLDA2', 'CCL2',
                      'IL6', 'PDGFRA',
                      'COL1A1', 'CXCL12',
                      'CD82', 'CCL19', 'KCNE4', 'CPE', 'TAGLN', 'MYL9', 'HEY1', 'CCL14', 'PECAM1',
                      'ACKR1', 'SELE', 'SNCG', 'HES1', 'MMRN1', 'XCL1' ,'XCL2', 'KLRC1', 'CCL5', 'GZMA', 
                      'CD8A', 'CD8B', 'IL7R', 'FOXP3', 'TIGIT', 'TPSB2', 'TPSAB1', 'CD79A', 'JCHAIN', 'CD163',
                      'C1QB', 'FCGR2A', 'MS4A6A', 'IL23A', 'CLEC9A', 'CLEC10A', 'CD1C', 'CD207', 'CD1A', 
                      'CD14', 'IL1B', 'CCR7', 'IDO1')


```

```{r fig.height=10, fig.width=10}
Idents(seuratobj) <- 'RNA_snn_res.0.2'
p <- DotPlot(seuratobj, features = rev(markers_reynolds), cluster.idents = FALSE) + coord_flip()
p
```




### Keratinocytes
```{r fig.height = 3, fig.width = 12}
markers_keratinocytes <- c('KRT5', 'KRT14', 'GATA3', 'KRTDAP')
seurat_feature(seuratobj, features = markers_keratinocytes, nrow = 1)
```


### Mast Cell
```{r}
seurat_feature(seuratobj, features = c('CD3D','TPSAB1', 'COL6A2'))
```
### Subclusters
Use UMAP as input to clustering for cleanly selecting clusters
```{r}
seuratobj <- seuratobj %>% 
  FindNeighbors(reduction = 'umap', 
                dims = 1:2) %>% 
  FindClusters(resolution = 0.01)

seuratobj %>% seurat_feature(features = 'seurat_clusters')
```

```{r}
apc_clusters <- 2
lymphocyte_clusters <- c(0, 1)
apcs <- subset(seuratobj, subset = seurat_clusters %in% apc_clusters)
lymphocytes <- subset(seuratobj, subset = seurat_clusters %in% lymphocyte_clusters)

p1 <- seurat_feature(apcs, facet_hide = TRUE)
p2 <- seurat_feature(lymphocytes, facet_hide = TRUE)
p1 + p2
```

### Identify variable genes in these subgroups separately, then recombine objects for grouped analysis
```{r}
apcs <- FindVariableFeatures(apcs, nfeatures = 3000)
lymphocytes <- FindVariableFeatures(lymphocytes, nfeatures = 3000)

combined_features <- union(apcs@assays$RNA@var.features, lymphocytes@assays$RNA@var.features)
combined_features %>% length()
```
### Recombine and recluster
```{r}
subcluster <- subset(seuratobj, 
                     subset = seurat_clusters %in% c(apc_clusters, lymphocyte_clusters))

VariableFeatures(subcluster) <- combined_features

subcluster %>% 
  NormalizeData() %>% 
  ScaleData() %>% 
  RunPCA(features = combined_features) %>% 
  RunHarmony(c("treatment", "chemistry"), 
             plot_convergence = TRUE, 
             assay.use = 'RNA')

ElbowPlot(subcluster, ndims = 30, reduction = 'harmony')
```

### RunUMAP, FindNeighbors, and FindClusters 
```{r}
dims <- 1:30
assay <- 'RNA'
resolution <- seq(0.2, 2.4, by = 0.2)
  
subcluster <- subcluster %>% 
  RunUMAP(reduction = "harmony", dims = dims, assay = assay) %>% 
  FindNeighbors(reduction = "harmony", dims = dims, assay = assay) %>% 
  FindClusters(resolution = resolution)

### Fix cluster factor levels
clusterings <- colnames(subcluster@meta.data) %>% str_subset('_res')

for(i in clusterings) {
  clusters <- subcluster@meta.data[[i]]
  subcluster[[i]] <- factor(clusters, levels = levels(clusters) %>% as.numeric() %>% sort())
}

subcluster %>% seurat_feature(facet_hide = TRUE, title = 'Resolution: 2.4', color_package = 'generate', color_palette = 'pimp')
```

```{r}
subcluster %>% write_rds(file.path(data_dir, 'subcluster.rds'))
```


### Pipeline
```{r}
recluster <- function(seuratobj, 
                      assay = 'RNA',
                      batch = c('treatment', 'chemistry'),
                      dims,
                      resolution = seq(0.2, 2, by = 0.2),
                      default_ident = 'RNA_snn_res.0.2') {
  
  seuratobj <- seuratobj %>% 
    FindVariableFeatures() %>% 
    RunPCA() %>% 
    RunHarmony(c("treatment", "chemistry"), 
               plot_convergence = TRUE, 
               assay.use = 'RNA')
  
  ElbowPlot(seuratobj, ndims = 30, reduction = 'harmony')
  
  ### RunUMAP, FindNeighbors, and FindClusters 
  
  seuratobj <- seuratobj %>% 
    RunUMAP(reduction = "harmony", dims = dims, assay = assay) %>% 
    FindNeighbors(reduction = "harmony", dims = dims, assay = assay) %>% 
    FindClusters(resolution = resolution)
  
  ### Fix cluster factor levels
  clusterings <- colnames(seuratobj@meta.data) %>% str_subset('_res')
  
  for(i in clusterings) {
    clusters <- seuratobj@meta.data[[i]]
  seuratobj[[i]] <- factor(clusters, levels = levels(clusters) %>% as.numeric() %>% sort())
}

Idents(seuratobj) <- default_ident

seuratobj
}
```


## Session Info
```{r}
sessionInfo()
```



