---
title: "Annotation"
author: "David Wu"
---

Set working directory to project directory
```{r setup}
require(knitr)
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

### Optionally convert this Rmd to R script 
```{r eval=FALSE, message=FALSE}
current_file <- rstudioapi::getActiveDocumentContext()$path
output_file <- stringr::str_replace(current_file, '.Rmd', '.R')
knitr::purl(current_file, output = output_file)
file.edit(output_file)
```

### Directories
```{r, message=FALSE}
output_dir <- 'analysis/output/05_annotate' # analysis file output directory
data_dir <- 'data/derived/seurat' # data file output directory

dir.create(output_dir, showWarnings = FALSE)
dir.create(data_dir, showWarnings = FALSE)
```

### Loading the libraries
```{r, message=FALSE}
library(Seurat) 
library(tidyverse)
library(ggExtra)
library(SeuratWrappers)
library(wutilities) # devtools::install_github('symbiologist/wutilities')
library(tictoc)
library(googlesheets4)
```

```{r}
seuratobj <- read_rds(file.path(data_dir, 'seuratobj_subcluster.rds'))
seuratobj
```
### Import annotation table
```{r}
annotations <- read_sheet('https://docs.google.com/spreadsheets/d/1pLMOaXkSGq3qSX7bTIw2daWd3ItA5JxqFomQn_km1IU/edit#gid=0') %>% 
  filter(!is.na(cluster)) %>% 
  mutate(cluster = factor(cluster))

annotations$label <- factor(annotations$label, levels = annotations$label)
annotations$merged_label <- factor(annotations$merged_label, levels = unique(annotations$merged_label))
annotations
```
### Ensure factor level ordering
```{r}
levels(annotations$label)
```

#### Get current clusters and integrate with annotations
```{r}
cluster_annotation <- seuratobj@meta.data %>% 
  select(cluster = seurat_clusters) %>% 
  rownames_to_column() %>% 
  left_join(annotations %>% select(cluster, label, merged_label)) %>% 
  column_to_rownames('rowname')
cluster_annotation
```
### Add Metadata
```{r}
seuratobj <- AddMetaData(seuratobj, cluster_annotation)
Idents(seuratobj) <- 'merged_label'
Idents(seuratobj) %>% head()
```

```{r}
p <- seuratobj %>% seurat_feature(features = 'merged_label', facet_hide = TRUE, size = 0.1, alpha = 0.1)

ggsave(plot = p,
       filename = 'umap_annotated.png',
       path = output_dir,
       dpi = 100,
       w = 20, 
       h = 10)
p
```
### Fix metadata
Patient S1 and S2 treatment response was swapped; S1 should be partial and S2 should be full
```{r}
new_metadata <- seuratobj@meta.data %>% 
  transmute(response = case_when(
    str_detect(sample, 'S1') ~ 'partial',
    str_detect(sample, 'S2') ~ 'full'
))

new_metadata
```
```{r}
seuratobj$response <- new_metadata$response
```


### Export
```{r}
seuratobj@meta.data %>% rownames_to_column() %>% write_rds(file.path(output_dir, 'metadata.rds'), compress = 'gz')
seuratobj %>% write_rds(file.path(data_dir, 'seuratobj_annotated.rds'))
```


## Session Info
```{r}
sessionInfo()
```



