---
title: "Annotation"
author: "David Wu"
---

### Use the here package for robust relative paths
```{r}
library(here)
```

### Optionally convert this Rmd to R script 
```{r eval=FALSE, message=FALSE}
current_file <- rstudioapi::getActiveDocumentContext()$path
output_file <- stringr::str_replace(current_file, '.Rmd', '.R')
knitr::purl(current_file, output = output_file)
file.edit(output_file)
```

### Directories
```{r, message=FALSE}
output_dir <- 'analysis/output/07_annotate' # analysis file output directory
data_dir <- 'data/derived/tildra' # data file output directory

dir.create(output_dir, showWarnings = FALSE)
dir.create(data_dir, showWarnings = FALSE)
```

### Loading the libraries
```{r, message=FALSE}
library(Seurat) 
library(tidyverse)
library(ggExtra)
library(SeuratWrappers)
library(wutilities) # devtools::install_github('symbiologist/wutilities')
library(tictoc)
library(ggsankey)
library(googlesheets4)
theme_set(wutilities::theme_dwu())
```

### Import
#### Clustered object
```{r}
seuratobj <- read_rds(here(data_dir, 'seuratobj_subcluster.rds'))
metadata <- seuratobj@meta.data %>% rownames_to_column()
seuratobj
```

#### Import annotations from RashX paper
```{r}
rashx <- read_rds('data/external/rashx_clean.rds')
rashx_clusters <- read_tsv(here('analysis/output/rashx/rashx_clusters.tsv.gz'))
rashx_clusters %>% head()
```
#### Import transferred labels
```{r}
labels <- read_tsv(here('analysis/output/06_label/transfer_pca30_kNA.tsv.gz')) %>% select(rowname, label = predicted.id)

label_transfer <- metadata %>% select(rowname, local, global) %>% left_join(labels)
label_transfer %>% head()
```
### Import marker-based cluster-matching
```{r}
local_top <- read_tsv(here('analysis/output/06_label/local_top.tsv'))
global_top <- read_tsv(here('analysis/output/06_label/global_top.tsv')) %>% mutate(global = factor(global))
```

## Analysis
### Combine all annotations
```{r}
merged_annotations <- metadata %>% 
  select(rowname, id, local, global) %>% 
  left_join(label_transfer %>% select(rowname, label)) %>% 
  separate(rowname, sep = '_', into = c('barcode', NA), remove = FALSE) %>% 
  left_join(rashx_clusters %>% select(barcode, id, rashx = cluster)) %>% 
  as_tibble()

merged_annotations
```
### Agreement metrics
```{r}
clusters <- c('local', 'global') %>% set_names(.)
annotations <- c('label', 'rashx') %>% set_names(.)

all_clusterings <- c(clusters, annotations)

pairwise_table <- crossing('cluster1' = all_clusterings,
                           'cluster2' = all_clusterings) %>% 
  filter(cluster1 != cluster2) %>% 
  rowwise() %>% 
  mutate(combination = paste(sort(c(cluster1, cluster2)), collapse = ':')) %>% 
  group_by(combination) %>% 
  slice_max(cluster1) %>% 
  arrange(cluster1)
  
pairwise_table
```

```{r}
agreement_metrics <- map(1:nrow(pairwise_table), function(i) {
  row_input <- pairwise_table[i, ]
  
  cluster_input <- merged_annotations %>% select(all_of(c(row_input$cluster1, row_input$cluster2))) %>% drop_na()
  
  tibble('combination' = row_input$combination,
         calculate_agreement(cluster_input[[row_input$cluster1]], 
                             cluster_input[[row_input$cluster2]]) %>% 
    mutate(across(everything(), round, 3)))
  
}) %>% 
  bind_rows() %>% 
  rowwise() %>% 
  mutate(mean = mean(c(AMI, ARI))) %>% 
  arrange(-mean) %>% 
  ungroup()
agreement_metrics
```

### Calculate odds and overlap
```{r}
annotation_odds <- map(clusters, function(i) {
  map(annotations, function(j) {
    
    odds_table(merged_annotations %>% select(all_of(c(i, j))) %>% drop_na(), 
               group1 = i,
               group2 = j)
  })
})

annotation_odds$local$label
```


### Sankey plots
```{r}
dpi <- 300

sankey1_loop <- crossing('cluster1' = all_clusterings,
                         'cluster2' = all_clusterings) %>% 
  filter(cluster1 != cluster2) %>% 
  left_join(merged_annotations %>% select(all_of(all_clusterings)) %>% unique() %>% 
    pivot_longer(cols = everything(),
                 names_to = 'cluster1',
                 values_to = 'celltype') %>% 
    unique() %>% 
    drop_na() %>% 
    arrange(cluster1, celltype))

sankey1_loop
  
```

```{r}
library(parallel)
cores <- 20

tic()
mclapply(1:nrow(sankey1_loop), mc.cores = cores, FUN = function(i) {
  
  parameters <- sankey1_loop[i,]
  
  print2(paste(parameters$cluster1, parameters$cluster2, parameters$celltype))
  
  nodes <- c(parameters$cluster1, parameters$cluster2)
    
    print2(paste(parameters$cluster1, parameters$cluster2, parameters$celltype, sep = ': '))
    
    p <- plot_sankey1(merged_annotations %>% filter(.[[parameters$cluster1]] == parameters$celltype),
                      nodes = nodes,
                      names = str_to_sentence(nodes), 
                      drop_na = TRUE,
                      prefix = 'Cells: ')
    
    ggsave(plot = p,
           filename = paste0(str_remove(parameters$celltype, '/'), '.png'),
           path = here(output_dir, parameters$cluster1, paste0('sankey1_', parameters$cluster1, '_vs_', parameters$cluster2)),
           dpi = dpi,
           h = 4,
           w = 6)  
    
})
toc()
```


### Two-way
```{r}
sankey2_loop <- merged_annotations %>% 
  select(all_of(clusters)) %>% 
  unique() %>% 
  pivot_longer(cols = everything(),
               names_to = 'cluster',
               values_to = 'celltype') %>% 
  unique() %>% 
  drop_na() %>% 
  arrange(cluster, celltype) %>% 
  mutate(across(everything(), as.character))

sankey2_loop
```

```{r}
tic()
mclapply(1:nrow(sankey2_loop), mc.cores = cores, FUN = function(i) {

  parameters <- sankey2_loop[i,]
  
  print2(paste(parameters$cluster, parameters$celltype, sep = ': '))
  
  nodes <- c(annotations[1], parameters$cluster, annotations[2])
  
  p <- plot_sankey2(merged_annotations %>% filter(.[[parameters$cluster]] == parameters$celltype),
                    nodes = nodes,
                    names = str_to_sentence(nodes), 
                    drop_na = FALSE,
                    prefix = 'Cells: ')
  
  ggsave(plot = p,
         filename = paste0(str_remove(parameters$celltype, '/'), '.png'),
         path = here(output_dir, parameters$cluster, paste0('sankey2_', parameters$cluster, '_vs_', annotations[1], '_vs_', annotations[2])),
         dpi = dpi,
         h = 4,
         w = 6)
})
toc()
```


#### Annotation centric
```{r}
annotation_loop <- merged_annotations %>% 
  select(all_of(annotations)) %>% 
  unique() %>% 
  pivot_longer(cols = everything(),
               names_to = 'cluster',
               values_to = 'celltype') %>% 
  unique() %>% 
  drop_na() %>% 
  arrange(cluster, celltype) %>% 
  mutate(across(everything(), as.character))

annotation_loop
```

```{r}
mclapply(1:nrow(annotation_loop), mc.cores = cores, FUN = function(i) {

  parameters <- annotation_loop[i,]
  
  print2(paste(parameters$cluster, parameters$celltype, sep = ': '))
  
  # vs clusters
  nodes <- c(clusters[1], parameters$cluster, clusters[2])
  p <- plot_sankey2(merged_annotations %>% filter(.[[parameters$cluster]] == parameters$celltype),
                    nodes = nodes,
                    names = str_to_sentence(nodes), 
                    drop_na = FALSE, 
                    prefix = 'Cells: ')
  
  ggsave(plot = p,
         filename = paste0(str_remove(parameters$celltype, '/'), '.png'),
         path = here(output_dir, parameters$cluster, 'sankey2_vs_clusters'),
         dpi = dpi,
         h = 4,
         w = 6)
  
  # vs other annotation
  other_annotation <- setdiff(annotations, parameters$cluster)
  nodes <- c(parameters$cluster, other_annotation)
  
  p <- plot_sankey1(merged_annotations %>% filter(.[[parameters$cluster]] == parameters$celltype),
                    nodes = nodes,
                    names = str_to_sentence(nodes), 
                    drop_na = FALSE, 
                    top_n = 4,
                    prefix = 'Cells: ')
  
  ggsave(plot = p,
         filename = paste0(str_remove(parameters$celltype, '/'), '.png'),
         path = here(output_dir, parameters$cluster, paste0('sankey1_vs_', other_annotation)), 
         dpi = dpi,
         h = 4,
         w = 6)
  
})
```



```{r}
library(pheatmap)

pheatmap_input <- annotation_odds$local$label %>% 
  select(group1,
         group2,
         metric = overlap_coef) %>% 
  pivot_wider(names_from = group1, values_from = metric, values_fill = 0)

p <- pheatmap(mat = pheatmap_input %>% column_to_rownames('group2'), 
         angle_col = 0, 
         color = c('white', colorRampPalette(RColorBrewer::brewer.pal(n = 7, name = "Blues"))(100)),
         breaks = seq(0, 1, length.out = 102), display_numbers = TRUE)
```

### Annotate
#### Factor level ordering
```{r}
rashx_clusters$supercluster %>% unique() %>% sort()
```

```{r}
supercluster_order <- c('Tcm', 
                        'Trm', 
                        'Tmm', 
                        'CTL',
                        'CTLem',
                        'CTLex',
                        'Tet',
                        'Tn',
                        'Treg',
                        'cmTreg',
                        'eTreg',
                        'ILC',
                        'NK',
                        'ILC/NK',
                        'B',
                        'Plasma',
                        'Mac',
                        'Mono',
                        'InfMono',
                        'DC',
                        'migDC',
                        'moDC',
                        'LC',
                        'Mast')

annotation_order <- rashx_clusters %>% select(cluster, supercluster) %>% 
  unique() %>% 
  mutate(supercluster = factor(supercluster, levels = supercluster_order)) %>% 
  arrange(supercluster, cluster) %>% 
  mutate(order = 1:n(),
         supercluster = as.character(supercluster))

annotation_order
```


#### Prepare addition of local and global annotations
```{r}
annotation_local <- annotation_odds$local$label %>% 
  group_by(group1) %>% 
  slice_max(f_group1) %>% 
  select(local = group1,
         cluster = group2,
         f_in_cluster = f_group1) %>% 
  left_join(local_top) %>% # expression-based cluster-matching
  mutate(match = ifelse(label == cluster, TRUE, FALSE))

annotation_local
```
```{r}
annotation_global <- annotation_odds$global$label %>% 
  group_by(group1) %>% 
  slice_max(f_group1) %>% 
  select(global = group1,
         cluster = group2,
         f_in_cluster = f_group1) %>% 
  left_join(global_top) %>% # expression-based cluster-matching
  mutate(match = ifelse(label == cluster, TRUE, FALSE))

annotation_global
```

#### Manually curate select clusters based on DotPlot
```{r}
rashx_markers <- c("CD3D",
                  "CCR7",
                  "SELL",
                  "KLF2","CD69","ITGAE","CXCR6","CD4","TIGIT","FOXP3",
                  "IL2RA","CTLA4","CD8A","CD8B","GZMB","PDCD1","LAG3","KLRB1","PRF1","KLRD1","GNLY",
                  "TNFRSF18","PRDM1","BATF","TRAT1","RORA","GATA3","PTGDR2",
                  "IL7R", "HLA-DRA","HLA-DRB1",
                  "CD83","IDO1","CD207","EPCAM","CD68","C1QB","C1QC","CD163","CLEC10A","CD1C",
                  "THBD","XCR1","SIRPA","F13A1","IGKC","JCHAIN","CD79A","MS4A1","NR4A1","NR4A2","KLF4",
                  "CEBPB","LYZ","MS4A7","SERPINA1","CD14","S100A9","IL23A","TPSB2","TPSAB1","MKI67","TOP2A",
                  "ITGA4","NCR1","IL17A","IL17F","IL23R")

dotplot_custom <- function(seuratobj,
                           features,
                           group.by,
                           assay = 'RNA',
                           cluster.idents = TRUE) {
  
  DotPlot(object = seuratobj, 
          features = features, 
          group.by = group.by,
          cluster.idents = cluster.idents, 
          assay = assay, 
          col.min = 0.3, 
          col.max = 0.8, 
          dot.min = 0.12, 
          dot.scale = 1) +
    scale_size(range = c(0, 5))+ 
    scale_size_area(max_size = 5)+ 
    theme(text = element_text(family = 'Arial'),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1),
          plot.background = element_rect(fill = 'white')) +
    labs(x = '',
         y = '') +
    scale_color_gradientn(colours = viridis::magma(20), limits = c(0,1), oob = scales::squish)
  
}

p_rashx <- dotplot_custom(rashx,
                          features = rashx_markers,
                          group.by = 'Ident2')

ggsave(plot = p_rashx,
       path = output_dir,
       width = 14,
       height = 8,
       filename = 'dotplot_rashx.png')

p_local <- dotplot_custom(seuratobj,
                          features = rashx_markers,
                          group.by = 'local')

ggsave(plot = p_local,
       path = output_dir,
       width = 14,
       height = 8,
       filename = 'dotplot_local.png')

p_global <- dotplot_custom(seuratobj,
                           features = rashx_markers,
                           group.by = 'global')

ggsave(plot = p_global,
       path = output_dir,
       width = 14,
       height = 8,
       filename = 'dotplot_global.png')

p_local
```
```{r}
df_rashx <- ggplot_build(p_rashx)
data.plot <- df_rashx$plot$data
data.plot
```

#### Custom dot plot function for dual object clustering
```{r}
dotplot_dataprep <- function(seuratobj,
                             features,
                             cells = NULL,
                             group.by,
                             prefix = NULL) {
  
  data.features <- FetchData(seuratobj,
                             vars = features,
                             cells = cells)
  
  if(is.null(prefix)) {
    prefix <- group.by
  }
  
  seuratobj[['group']] <- paste0(prefix, '_', seuratobj[[group.by]][[1]])
  data.features$id <- seuratobj[['group', drop = TRUE]]
  id.levels <- levels(x = data.features$id)
  data.features$id <- as.vector(x = data.features$id)
  
  data.plot <- lapply(
    X = unique(x = data.features$id),
    FUN = function(ident) {
      data.use <- data.features[data.features$id == ident, 1:(ncol(x = data.features) - 1), drop = FALSE]
      avg.exp <- apply(
        X = data.use,
        MARGIN = 2,
        FUN = function(x) {
          return(mean(x = expm1(x = x)))
        }
      )
      pct.exp <- apply(X = data.use, MARGIN = 2, FUN = PercentAbove, threshold = 0)
      return(list(avg.exp = avg.exp, pct.exp = pct.exp))
    }
  )
  
  names(x = data.plot) <- unique(x = data.features$id)
  data.plot
}

dotplot_cluster <- function(data.plot) {
  
  id.levels <- names(data.plot)
  
  mat <- do.call(
    what = rbind,
    args = lapply(X = data.plot, FUN = unlist)
  )
  mat <- scale(x = mat)
  id.levels <- id.levels[hclust(d = dist(x = mat))$order]
  
  data.plot <- lapply(
    X = names(x = data.plot),
    FUN = function(x) {
      data.use <- as.data.frame(x = data.plot[[x]])
      data.use$features.plot <- rownames(x = data.use)
      data.use$id <- x
      return(data.use)
    }
  )
  data.plot <- do.call(what = 'rbind', args = data.plot)
  
  if (!is.null(x = id.levels)) {
    data.plot$id <- factor(x = data.plot$id, levels = id.levels)
  }
  
  data.plot
  
}

dotplot_scale <- function(data.plot, 
                          col.min = 0.3, 
                          col.max = 0.8) {
  avg.exp.scaled <- sapply(
    X = unique(x = data.plot$features.plot),
    FUN = function(x) {
      data.use <- data.plot[data.plot$features.plot == x, 'avg.exp']
        data.use <- scale(x = data.use)
        data.use <- MinMax(data = data.use, min = col.min, max = col.max)

      return(data.use)
    }
  )
  
  avg.exp.scaled <- as.vector(x = t(x = avg.exp.scaled))
  
  data.plot$avg.exp.scaled <- avg.exp.scaled
  
  data.plot
  
  
}

doubledotplot <- function(seuratobj1,
                          seuratobj2,
                          group.by1,
                          group.by2,
                          prefix1,
                          prefix2,
                          features,
                          dot.min = 0.12,
                          dot.scale = 1,
                          scale.min = NA,
                          scale.max = NA,
                          col.min = 0.3,
                          col.max = 0.8,
                          plot = TRUE) {
  
  dataprep1 <- dotplot_dataprep(seuratobj1, features = features, group.by = group.by1, prefix = prefix1)
  dataprep2 <- dotplot_dataprep(seuratobj2, features = features, group.by = group.by2, prefix = prefix2)
  
  data_combined <- dotplot_cluster(c(dataprep1, dataprep2))
  
  data.plot <- data_combined %>% dotplot_scale(col.min = col.min,
                                               col.max = col.max)
  
  data.plot$pct.exp[data.plot$pct.exp < dot.min] <- NA
  data.plot$pct.exp <- data.plot$pct.exp * 100
  
   if (!is.na(x = scale.min)) {
    data.plot[data.plot$pct.exp < scale.min, 'pct.exp'] <- scale.min
  }
  if (!is.na(x = scale.max)) {
    data.plot[data.plot$pct.exp > scale.max, 'pct.exp'] <- scale.max
  }
  
  data.plot$features.plot <- factor(data.plot$features.plot, levels = features)
  
  if (plot) {
    data.plot %>% 
      ggplot(aes(x = features.plot,
                 y = id)) +
      geom_point(aes(size = pct.exp,
                     color = avg.exp.scaled)) +
      scale_radius(range = c(0, dot.scale), limits = c(scale.min, scale.max)) +
      guides(size = guide_legend(title = 'Percent Expressed')) +
      theme_dwu() +
      theme(axis.title.x = element_blank(), 
            axis.title.y = element_blank(),
            axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
      scale_color_gradientn(colours = viridis::magma(20), limits = c(0, 1), oob = scales::squish) +
      scale_size(range = c(0, 5))+ 
      scale_size_area(max_size = 5)
  }
  
  else {
    data.plot
  }
}

dotplot_filtered_plot <- function(data.plot,
                                  id = NULL,
                                  features = NULL,
                                  dot.min = 0.12,
                                  dot.scale = 1,
                                  scale.min = NA,
                                  scale.max = NA,
                                  col.min = 0.3,
                                  col.max = 0.8) {
  
  if(!is.null(id)) {
    data.plot <- data.plot %>% filter(id %in% {{id}})
  }
  if(!is.null(features)) {
    data.plot <- data.plot %>% filter(features.plot %in% {{features}})
  }
    
   data.plot %>% 
      ggplot(aes(x = features.plot,
                 y = id)) +
      geom_point(aes(size = pct.exp,
                     color = avg.exp.scaled)) +
      scale_radius(range = c(0, dot.scale), limits = c(scale.min, scale.max)) +
      guides(size = guide_legend(title = 'Percent Expressed')) +
      theme_dwu() +
      theme(axis.title.x = element_blank(), 
            axis.title.y = element_blank(),
            axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
      scale_color_gradientn(colours = viridis::magma(20), limits = c(0, 1), oob = scales::squish) +
      scale_size(range = c(0, 5))+ 
      scale_size_area(max_size = 5)
}
```

```{r}
p_dotplot_rashx_v_local <- doubledotplot(seuratobj1 = rashx,
                                         seuratobj2 = seuratobj,
                                         group.by1 = 'Ident2',
                                         group.by2 = 'local',
                                         prefix1 <- 'RashX',
                                         prefix2 <- 'Local',
                                         features = rashx_markers)

p_dotplot_rashx_v_global <- doubledotplot(seuratobj1 = rashx,
                                          seuratobj2 = seuratobj,
                                          group.by1 = 'Ident2',
                                          group.by2 = 'global',
                                          prefix1 <- 'RashX',
                                          prefix2 <- 'Global',
                                          features = rashx_markers)

ggsave(plot = p_dotplot_rashx_v_local,
       path = output_dir,
       width = 14,
       height = 14,
       filename = 'dotplot_rashx_v_local.png')

ggsave(plot = p_dotplot_rashx_v_global,
       path = output_dir,
       width = 14,
       height = 14,
       filename = 'dotplot_rashx_v_global.png')


```
```{r}
map(clusters, function(which_cluster) {
  
  
})

```

#### Individual plots 
```{r}
top_n <- 5

map(clusters, function(which_cluster) {
  
  prefix1 <- str_to_sentence(which_cluster)
  prefix2 <- 'RashX'

  df_dotplot <- doubledotplot(seuratobj1 = seuratobj,
                              group.by1 = which_cluster,
                              prefix1 = prefix1,
                              seuratobj2 = rashx,
                              group.by2 = 'Ident2',
                              prefix2 <- prefix2,
                              features = rashx_markers,
                              plot = FALSE)
  
  top_per_cluster <- annotation_odds[[which_cluster]]$label %>% 
    group_by(group1) %>% 
    slice_max(f_group1, n = top_n) %>% 
    select(group1, group2, f_group1)
  
  map(top_per_cluster$group1 %>% unique(), function(which_cluster_id) {
    
    print2(which_cluster_id)
    top_matches <- top_per_cluster %>% filter(group1 == which_cluster_id) %>% pull(group2)
    
    ids <- c(paste0(prefix1, '_', which_cluster_id),
             paste0(prefix2, '_', top_matches))
    
    p_individual_dotplot <- dotplot_filtered_plot(df_dotplot,
                                                 id = ids) +
      ggtitle(paste0('Closest Matches for Cluster ', which_cluster_id))
    
    ggsave(plot = p_individual_dotplot,
           path = here(output_dir, 'dotplot', which_cluster),
           width = 14,
           height = 3,
           filename = paste0(which_cluster_id, '.png'))
    
  })
  
  
})



```


#### Manual curation of transferred labels
Write transferred labels to Google Sheets and use markers to refine labels
```{r eval=FALSE}
library(googledrive)
library(googlesheets4)
options(httr_oob_default=TRUE)

ss <- drive_get('Tildra project')
write_sheet(annotation_local, ss = ss, sheet = 'local')
write_sheet(annotation_global, ss = ss, sheet = 'global')

```
Import curated labels
```{r eval=FALSE}
annotation_local_curated <- read_sheet(ss = ss, sheet = 'local_curated')
annotation_local_curated %>% write_tsv(here(output_dir, 'annotation_local_curated.tsv'))
```

```{r}
annotation_local_curated <- read_tsv(here(output_dir, 'annotation_local_curated.tsv'))
annotation_local_curated
```

```{r}
annotation_local_final <- annotation_local_curated %>% 
  select(local, cluster = curated) %>% 
  left_join(annotation_order) %>% 
  ungroup() %>% 
  select(local, 
         local_cluster = cluster, 
         local_supercluster = supercluster, 
         local_order = order) 

annotation_local_final
```


```{r}
annotation_global <- annotation_odds$global$label %>% 
  group_by(group1) %>% 
  slice_max(f_group1) %>% 
  select(group1,
         cluster = group2,
         f_in_cluster = f_group1) %>% 
  left_join(annotation_order) %>% 
  ungroup() %>% 
  select(global = group1, 
         global_cluster = cluster, 
         global_supercluster = supercluster, 
         global_order = order,
         f_in_global_cluster = f_in_cluster)

annotation_global
```




```{r}
annotation_local_curated <- 
  annotation_local

annotation_local %>% 
```


```{r}
annotation_local_levels <- annotation_local %>% 
  select(local_order, local_cluster, local_supercluster) %>% 
  unique() %>% 
  arrange(local_order)

annotation_global_levels <- annotation_global %>% 
  select(global_order, global_cluster, global_supercluster) %>% 
  unique() %>% 
  arrange(global_order)

annotation_global_levels
```

#### Merged annotation information
```{r}
new_annotations <- seuratobj@meta.data %>% 
  rownames_to_column() %>% 
  select(rowname, local, global) %>% 
  left_join(annotation_local) %>% 
  left_join(annotation_global) %>% 
  mutate(local_cluster = factor(local_cluster, levels = annotation_local_levels$local_cluster),
         local_supercluster = factor(local_supercluster, levels = unique(annotation_local_levels$local_supercluster)),
         global_cluster = factor(global_cluster, levels = annotation_global_levels$global_cluster),
         global_supercluster = factor(global_supercluster, levels = unique(annotation_global_levels$global_supercluster))) %>% 
  column_to_rownames()

new_annotations 
```
### Add annotation
```{r}
seuratobj <- AddMetaData(seuratobj, new_annotations)
```
#### Ensure correct sample factor levels
```{r}
sample_order <- seuratobj@meta.data %>% 
  select(patient, sample, sample_order) %>% 
  unique() %>% 
  arrange(sample_order)
sample_order
```

```{r}
seuratobj$condition <- factor(seuratobj$condition, levels = c('HC', 'AD', 'PV'))
seuratobj$treatment <- factor(seuratobj$treatment, levels = c('None', 'Pre', 'Mid'))
seuratobj$group <- factor(seuratobj$group, levels = c('Responder', 'Non-responder'))
seuratobj$patient <- factor(seuratobj$patient, levels = unique(sample_order$patient))
seuratobj$sample <- factor(seuratobj$sample, levels = sample_order$sample)
```

### Use global or local identities and set as cluster/supercluster in the object
Use this code to select between global or local identities as the default for a given analysis
Here we will set global as the default
```{r}
which_identity <- 'local'
which_cluster <- paste0(which_identity, '_cluster')
which_supercluster <- paste0(which_identity, '_supercluster')

seuratobj$cluster <- seuratobj[[which_cluster]] 
seuratobj$supercluster <- seuratobj[[which_supercluster]] 

seuratobj$annotation <- seuratobj$cluster # set cluster as main annotation
Idents(seuratobj) <- 'annotation'
```
### Export
```{r}
seuratobj@meta.data %>% rownames_to_column() %>% write_rds(here(output_dir, 'metadata.rds'), compress = 'gz')

seuratobj %>% write_rds(here(data_dir, 'seuratobj_annotated.rds'))
```

### Save plots
```{r}
dpi <- 600
w <- 5
h <- 5
```

#### Local
```{r}
p <- seuratobj %>% 
  seurat_feature(features = 'local_cluster', 
                 facet_hide = TRUE, 
                 legend_position = 'none',
                 size = 0.1, 
                 alpha = 0.1, 
                 color_package = 'carto', 
                 color_palette = 'Bold', 
                 rasterize_dpi = dpi)

ggsave(plot = p,
       filename = 'umap_annotated_local_cluster.png',
       dpi = dpi,
       path = output_dir,
       w = w, 
       h = h)

p
```


```{r}
p <- seuratobj %>% 
  seurat_feature(features = 'local_supercluster', 
                 facet_hide = TRUE, 
                 legend_position = 'none', 
                 size = 0.1, alpha = 0.1, 
                 color_package = 'carto', 
                 color_palette = 'Bold', 
                 rasterize_dpi = dpi)

ggsave(plot = p,
       filename = 'umap_annotated_local_supercluster.png',
       path = output_dir,
       dpi = dpi,
       w = w, 
       h = h)

p

```
```{r}
p <- seuratobj %>% 
  seurat_feature(features = 'local', 
                 facets = 'local_cluster', 
                 legend_position = 'none',
                 size = 0.1, 
                 alpha = 0.1, 
                 label = FALSE,
                 color_package = 'carto', 
                 color_palette = 'Bold', 
                 rasterize_dpi = dpi)

ggsave(plot = p,
       filename = 'umap_annotated_local_cluster_facets.png',
       dpi = dpi,
       path = output_dir,
       w = w * 2, 
       h = h * 2)
```

#### Global
```{r}
p <- seuratobj %>% 
  seurat_feature(features = 'global_cluster', 
                 facet_hide = TRUE, 
                 legend_position = 'none',
                 size = 0.1, 
                 alpha = 0.1, 
                 color_package = 'carto', 
                 color_palette = 'Bold', 
                 rasterize_dpi = dpi)

ggsave(plot = p,
       filename = 'umap_annotated_global_cluster.png',
       dpi = dpi,
       path = output_dir,
       w = w, 
       h = h)

p
```
```{r}
p <- seuratobj %>% 
  seurat_feature(features = 'global', 
                 facets = 'global_cluster', 
                 legend_position = 'none',
                 size = 0.1, 
                 alpha = 0.1, 
                 label = FALSE,
                 color_package = 'carto', 
                 color_palette = 'Bold', 
                 rasterize_dpi = dpi)

ggsave(plot = p,
       filename = 'umap_annotated_global_cluster_facets.png',
       dpi = dpi,
       path = output_dir,
       w = w * 2, 
       h = h * 2)
```


```{r}
p <- seuratobj %>% 
  seurat_feature(features = 'global_supercluster', 
                 facet_hide = TRUE, 
                 legend_position = 'none', 
                 size = 0.1, alpha = 0.1, 
                 color_package = 'carto', 
                 color_palette = 'Bold', 
                 rasterize_dpi = dpi)

ggsave(plot = p,
       filename = 'umap_annotated_global_supercluster.png',
       path = output_dir,
       dpi = dpi,
       w = w, 
       h = h)

p

```

## Session Info
```{r}
sessionInfo()
```
