---
title: "DE analysis"
author: "David Wu"
---

## Setup
### Use the here package for robust relative paths
```{r}
library(here)
```

### Optionally convert this Rmd to R script 
```{r eval=FALSE, message=FALSE}
current_file <- rstudioapi::getActiveDocumentContext()$path
output_file <- stringr::str_replace(current_file, '.Rmd', '.R')
knitr::purl(current_file, output = output_file)
file.edit(output_file)
```

### Directories
```{r, message=FALSE}
output_dir <- here('analysis/output/08_de') # analysis file output directory
data_dir <- here('data/derived/tildra') # data file output directory

dir.create(output_dir, showWarnings = FALSE)
dir.create(data_dir, showWarnings = FALSE)
```

### Loading the libraries
```{r, message=FALSE}
library(Seurat) 
library(tidyverse)
library(ggExtra)
library(SeuratWrappers) # devtools::install_github('satijalab/seurat-wrappers')
library(wutilities) # devtools::install_github('symbiologist/wutilities')
library(tictoc)
library(openxlsx)
```

### Custom functions
```{r}
de_pipeline <- function(seuratobj,
                        cluster_col = 'annotation',
                        cluster = 'All',
                        comparison = NULL,
                        group1_id,
                        group2_id,
                        group1_barcodes = NULL,
                        group2_barcodes = NULL,
                        group1_filters, # named list of value pairs
                        group2_filters,
                        min.pct = 0.01,
                        logfc.threshold = log2(1.1),
                        test.use = 'Presto') {

  # Extract metadata, set cluster column, and filter columns
  metadata <- seuratobj@meta.data %>% rownames_to_column() 
  
  if(cluster != 'All') {
    
    metadata <- metadata %>% 
      mutate(cluster_col = .[[cluster_col]]) %>% 
      filter(cluster_col %in% {{cluster}})
    
  }
  
  # If barcodes are NULL, apply filters 
  if(is.null(group1_barcodes) | is.null(group2_barcodes)) {
    
    group1_barcodes <- wutilities::apply_filters(input = metadata,
                                                 filters = group1_filters) %>% pull(rowname)
     
    group2_barcodes <- wutilities::apply_filters(input = metadata,
                                                 filters = group2_filters) %>% pull(rowname)
  } 
  
  # Filter on barcodes
  metadata <- metadata %>% filter(rowname %in% c(group1_barcodes, group2_barcodes))
  
  # Run DE
  if(test.use == 'Presto') {
    
    de_genes <- RunPresto(object = seuratobj, 
                          ident.1 = group1_barcodes,
                          ident.2 = group2_barcodes, 
                          min.pct = min.pct,
                          logfc.threshold = logfc.threshold)
    
  } else {
    
    de_genes <- FindMarkers(object = seuratobj, 
                            ident.1 = group1,
                            ident.2 = group2, 
                            min.pct = min.pct,
                            logfc.threshold = logfc.threshold,
                            test.use = test.use) 
   
  }
  
  if(is.null(comparison)) {
    
    comparison <- paste(group1_id, 'vs', group2_id) 
      
  }
  
  de_genes %>%
    rownames_to_column('gene') %>% 
    mutate(pct.diff = pct.1 - pct.2,
           comparison,
           cluster = cluster,
           group1 = group1_id,
           group2 = group2_id,
           n_group1 = length(group1_barcodes),
           n_group2 = length(group2_barcodes)) %>% 
    select(comparison, cluster, gene, group1, group2, n_group1, n_group2, everything())

}

write_excel_notebook <- function(input_table, 
                                 output_file, 
                                 tabs = 'cluster', 
                                 prefix = 'cluster_') {
  OUT <- createWorkbook()
  
  input_table$tabs <- input_table[[tabs]]
  tabs <- input_table[[tabs]] %>% unique()
  
  
  for(i in tabs){
    tryCatch({
      table_subset <- input_table %>% filter(tabs == i) %>% select(-tabs)
      
      sheet_name <- paste0(prefix, i) %>% str_remove('[\\?/]') %>% str_remove('[\\/]') %>% str_remove('[\\:/]') 
      addWorksheet(OUT, sheet_name)
      writeData(OUT, sheet = sheet_name, x = table_subset)
      
    }, error=function(e){message(paste0("We have an error here in ", i))})
  }
  
  saveWorkbook(OUT, output_file, overwrite = TRUE)
}
```

### Import

```{r}
seuratobj <- read_rds(here(data_dir, 'seuratobj_annotated.rds'))
seuratobj
```
### Set identity
```{r}
which_annotation <- 'global_cluster'
seuratobj$annotation <- seuratobj[[which_annotation]]
Idents(seuratobj) <- 'annotation'

metadata <- seuratobj@meta.data %>% rownames_to_column()

output_dir <- here(output_dir, which_annotation)
dir.create(output_dir)
```

### Prepare metadata subset
```{r}
analyze <- metadata %>% 
  select(rowname, id, sample, patient, condition, treatment, annotation, response, group) %>% 
  as_tibble()

analyze %>% head()
```

```{r}
metadata_patient <- metadata %>% 
  select(id, patient, sample, treatment, condition, response, group) %>% 
  unique()

metadata_patient
```


### Global parameters
```{r}
which_test <- 'Presto'
which_annotation <- 'annotation'
sample_threshold <- 2 # at least 3 samples per comparison group (within a cluster)
responder_threshold <- 1 # fewer patients
cell_threshold <- 10 # at least 10 cells per sample (within a cluster)
```


### AD vs HC (11 v 11)
```{r eval=FALSE}
group1_filters <- list('condition' = 'AD')
group2_filters <- list('condition' = 'HC')

data_subset <- bind_rows(
  analyze %>% apply_filters(group1_filters) %>% mutate(group = 'group1'),
  analyze %>% apply_filters(group2_filters) %>% mutate(group = 'group2')
)

patients_per_cluster <- data_subset %>% 
  group_by(sample, group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  filter(n > cell_threshold) %>% 
  group_by(group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  pivot_wider(names_from = group,
              values_from = n,
              values_fill = 0) %>% 
  filter(group1 > sample_threshold,
         group2 > sample_threshold) 

de_ad_vs_hc <- map(patients_per_cluster[[which_annotation]], function(i) {
  
  print2(i)
  
  de_pipeline(seuratobj = seuratobj,
              cluster_col = which_annotation,
              cluster = i, 
              group1_id = 'AD',
              group2_id = 'HC',
              group1_filters = group1_filters,
              group2_filters = group2_filters)
  
}) %>% bind_rows()

de_ad_vs_hc %>% write_tsv(here(output_dir, 'de_ad_vs_hc.tsv.gz'))

write_excel_notebook(input_table = de_ad_vs_hc,
                     output_file = here(output_dir, 'de_ad_vs_hc.xlsx'))

de_ad_vs_hc %>% head()
```
### PV (13) vs HC (11)
```{r}
group1_filters <- list('condition' = 'PV',
                       'treatment' = c('None', 'Pre'))
group2_filters <- list('condition' = 'HC')

data_subset <- bind_rows(
  analyze %>% apply_filters(group1_filters) %>% mutate(group = 'group1'),
  analyze %>% apply_filters(group2_filters) %>% mutate(group = 'group2')
)

patients_per_cluster <- data_subset %>% 
  group_by(sample, group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  filter(n > cell_threshold) %>% 
  group_by(group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  pivot_wider(names_from = group,
              values_from = n,
              values_fill = 0) %>% 
  filter(group1 > sample_threshold,
         group2 > sample_threshold) 

de_pv_vs_hc <- map(patients_per_cluster[[which_annotation]], function(i) {
  
  print2(i)
  
  de_pipeline(seuratobj = seuratobj,
              cluster_col = which_annotation,
              cluster = i, 
              group1_id = 'PV',
              group2_id = 'HC',
              group1_filters = group1_filters,
              group2_filters = group2_filters)
  
}) %>% bind_rows()

de_pv_vs_hc %>% write_tsv(here(output_dir, 'de_pv_vs_hc.tsv.gz'))

write_excel_notebook(input_table = de_pv_vs_hc,
                     output_file = here(output_dir, 'de_pv_vs_hc.xlsx'))

de_pv_vs_hc %>% head()
```



### AD+PV vs HC (24 v 11)
```{r eval=FALSE}
group1_filters <- list('condition' = c('AD', 'PV'),
                       'treatment' = c('None', 'Pre'))
group2_filters <- list('condition' = 'HC')

data_subset <- bind_rows(
  analyze %>% apply_filters(group1_filters) %>% mutate(group = 'group1'),
  analyze %>% apply_filters(group2_filters) %>% mutate(group = 'group2')
)

patients_per_cluster <- data_subset %>% 
  group_by(sample, group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  filter(n > cell_threshold) %>% 
  group_by(group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  pivot_wider(names_from = group,
              values_from = n,
              values_fill = 0) %>% 
  filter(group1 > sample_threshold,
         group2 > sample_threshold) 

de_adpv_vs_hc <- map(patients_per_cluster[[which_annotation]], function(i) {
  
  print2(i)
  
  de_pipeline(seuratobj = seuratobj,
              cluster_col = which_annotation,
              cluster = i, 
              group1_id = 'AD+PV',
              group2_id = 'HC',
              group1_filters = group1_filters,
              group2_filters = group2_filters)
  
}) %>% bind_rows()

de_adpv_vs_hc %>% write_tsv(here(output_dir, 'de_adpv_vs_hc.tsv.gz'))

write_excel_notebook(input_table = de_adpv_vs_hc,
                     output_file = here(output_dir, 'de_adpv_vs_hc.xlsx'))

de_adpv_vs_hc %>% head()
```

### PV vs AD (13 v 11)
```{r eval=FALSE}
group1_filters <- list('condition' = 'AD')

group2_filters <- list('condition' = 'PV',
                       'treatment' = c('None', 'Pre'))

data_subset <- bind_rows(
  analyze %>% apply_filters(group1_filters) %>% mutate(group = 'group1'),
  analyze %>% apply_filters(group2_filters) %>% mutate(group = 'group2')
)

patients_per_cluster <- data_subset %>% 
  group_by(sample, group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  filter(n > cell_threshold) %>% 
  group_by(group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  pivot_wider(names_from = group,
              values_from = n,
              values_fill = 0) %>% 
  filter(group1 > sample_threshold,
         group2 > sample_threshold) 

de_ad_vs_pv <- map(patients_per_cluster[[which_annotation]], function(i) {
  
  print2(i)
  
  de_pipeline(seuratobj = seuratobj,
              cluster_col = which_annotation,
              cluster = i, 
              group1_id = 'AD',
              group2_id = 'PV',
              group1_filters = group1_filters,
              group2_filters = group2_filters)
  
}) %>% bind_rows()

de_ad_vs_pv %>% write_tsv(here(output_dir, 'de_ad_vs_pv.tsv.gz'))
write_excel_notebook(input_table = de_ad_vs_pv,
                     output_file = here(output_dir, 'de_ad_vs_pv.xlsx'))

de_ad_vs_pv %>% head() 


```


### Pre vs Mid, all treatment patients, all clusters
```{r}
group1_filters <- list('condition' = 'PV',
                       'treatment' = 'Pre')

group2_filters <- list('condition' = 'PV',
                       'treatment' = 'Mid')

data_subset <- bind_rows(
  analyze %>% apply_filters(group1_filters) %>% mutate(group = 'group1'),
  analyze %>% apply_filters(group2_filters) %>% mutate(group = 'group2')
)

patients_per_cluster <- data_subset %>% 
  group_by(sample, group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  filter(n > cell_threshold) %>% 
  group_by(group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  pivot_wider(names_from = group,
              values_from = n,
              values_fill = 0) %>% 
  filter(group1 > sample_threshold,
         group2 > sample_threshold) 

de_pre_vs_mid <- map(patients_per_cluster[[which_annotation]], function(i) {
  
  print2(i)
  
  de_pipeline(seuratobj = seuratobj,
              cluster_col = which_annotation,
              cluster = i, 
              group1_id = 'Pre',
              group2_id = 'Mid',
              group1_filters = group1_filters,
              group2_filters = group2_filters)
  
}) %>% bind_rows()

de_pre_vs_mid %>% write_tsv(here(output_dir, 'de_pre_vs_mid.tsv.gz'))
write_excel_notebook(input_table = de_pre_vs_mid,
                     output_file = here(output_dir, 'de_pre_vs_mid.xlsx'))

de_pre_vs_mid %>% head()
```

### Pre vs Mid, Responders
```{r}
group1_filters <- list('condition' = 'PV',
                       'treatment' = 'Pre',
                       'group' = 'Responder')

group2_filters <- list('condition' = 'PV',
                       'treatment' = 'Mid',
                       'group' = 'Responder')

data_subset <- bind_rows(
  analyze %>% apply_filters(group1_filters) %>% mutate(group = 'group1'),
  analyze %>% apply_filters(group2_filters) %>% mutate(group = 'group2')
)

patients_per_cluster <- data_subset %>% 
  group_by(sample, group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  filter(n > cell_threshold) %>% 
  group_by(group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  pivot_wider(names_from = group,
              values_from = n,
              values_fill = 0) %>% 
  filter(group1 > responder_threshold,
         group2 > responder_threshold) 

de_pre_vs_mid_responder <- map(patients_per_cluster[[which_annotation]], function(i) {
  
  print2(i)
  
  de_pipeline(seuratobj = seuratobj,
              cluster_col = which_annotation,
              cluster = i, 
              group1_id = 'Responder Pre',
              group2_id = 'Responder Mid',
              group1_filters = group1_filters,
              group2_filters = group2_filters)
  
}) %>% bind_rows() %>% 
  mutate(group = 'Responder')

de_pre_vs_mid_responder %>% write_tsv(here(output_dir, 'de_pre_vs_mid_responder.tsv.gz'))
write_excel_notebook(input_table = de_pre_vs_mid_responder,
                     output_file = here(output_dir, 'de_pre_vs_mid_responder.xlsx'))
```

### Pre vs Mid, Non-responders
```{r}
group1_filters <- list('condition' = 'PV',
                       'treatment' = 'Pre',
                       'group' = 'Non-responder')

group2_filters <- list('condition' = 'PV',
                       'treatment' = 'Mid',
                       'group' = 'Non-responder')

data_subset <- bind_rows(
  analyze %>% apply_filters(group1_filters) %>% mutate(group = 'group1'),
  analyze %>% apply_filters(group2_filters) %>% mutate(group = 'group2')
)

patients_per_cluster <- data_subset %>% 
  group_by(sample, group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  filter(n > cell_threshold) %>% 
  group_by(group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  pivot_wider(names_from = group,
              values_from = n,
              values_fill = 0) %>% 
  filter(group1 > responder_threshold,
         group2 > responder_threshold) 

de_pre_vs_mid_nonresponder <- map(patients_per_cluster[[which_annotation]], function(i) {
  
  print2(i)
  
  de_pipeline(seuratobj = seuratobj,
              cluster_col = which_annotation,
              cluster = i, 
              group1_id = 'Non-responder Pre',
              group2_id = 'Non-responder Mid',
              group1_filters = group1_filters,
              group2_filters = group2_filters)
  
}) %>% bind_rows() %>% 
  mutate(group = 'Non-responder')

de_pre_vs_mid_nonresponder %>% write_tsv(here(output_dir, 'de_pre_vs_mid_nonresponder.tsv.gz'))
write_excel_notebook(input_table = de_pre_vs_mid_nonresponder,
                     output_file = here(output_dir, 'de_pre_vs_mid_nonresponder.xlsx'))
```
### Responders vs non-responders in Pre samples (all clusters)
```{r}
group1_filters <- list('condition' = 'PV',
                       'treatment' = 'Pre',
                       'group' = 'Responder')

group2_filters <- list('condition' = 'PV',
                       'treatment' = 'Pre',
                       'group' = 'Non-responder')

data_subset <- bind_rows(
  analyze %>% apply_filters(group1_filters) %>% mutate(group = 'group1'),
  analyze %>% apply_filters(group2_filters) %>% mutate(group = 'group2')
)

patients_per_cluster <- data_subset %>% 
  group_by(sample, group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  filter(n > cell_threshold) %>% 
  group_by(group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  pivot_wider(names_from = group,
              values_from = n,
              values_fill = 0) %>% 
  filter(group1 > responder_threshold,
         group2 > responder_threshold) 

de_responder_vs_nonresponder_pre <- map(patients_per_cluster[[which_annotation]], function(i) {
  
  print2(i)
  
  de_pipeline(seuratobj = seuratobj,
              cluster_col = which_annotation,
              cluster = i, 
              group1_id = 'Responder Pre',
              group2_id = 'Non-responder Pre',
              group1_filters = group1_filters,
              group2_filters = group2_filters)
  
}) %>% bind_rows()

de_responder_vs_nonresponder_pre %>% write_tsv(here(output_dir, 'de_responder_vs_nonresponder_pre.tsv.gz'))
write_excel_notebook(input_table = de_responder_vs_nonresponder_pre,
                     output_file = here(output_dir, 'de_responder_vs_nonresponder_pre.xlsx'))
```

### Responders vs non-responders in mid samples (all clusters)
```{r}
group1_filters <- list('condition' = 'PV',
                       'treatment' = 'Mid',
                       'group' = 'Responder')

group2_filters <- list('condition' = 'PV',
                       'treatment' = 'Mid',
                       'group' = 'Non-responder')

data_subset <- bind_rows(
  analyze %>% apply_filters(group1_filters) %>% mutate(group = 'group1'),
  analyze %>% apply_filters(group2_filters) %>% mutate(group = 'group2')
)

patients_per_cluster <- data_subset %>% 
  group_by(sample, group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  filter(n > cell_threshold) %>% 
  group_by(group, across(all_of(which_annotation))) %>% 
  tally() %>% 
  pivot_wider(names_from = group,
              values_from = n,
              values_fill = 0) %>% 
  filter(group1 > responder_threshold,
         group2 > responder_threshold) 

de_responder_vs_nonresponder_mid <- map(patients_per_cluster[[which_annotation]], function(i) {
  
  print2(i)
  
  de_pipeline(seuratobj = seuratobj,
              cluster_col = which_annotation,
              cluster = i, 
              group1_id = 'Responder Mid',
              group2_id = 'Non-responder Mid',
              group1_filters = group1_filters,
              group2_filters = group2_filters)
  
}) %>% bind_rows()

de_responder_vs_nonresponder_mid %>% write_tsv(here(output_dir, 'de_responder_vs_nonresponder_mid.tsv.gz'))
write_excel_notebook(input_table = de_responder_vs_nonresponder_mid,
                     output_file = here(output_dir, 'de_responder_vs_nonresponder_mid.xlsx'))
```

### Per patient analysis across all clusters
```{r}
patients <- metadata_patient %>% 
  filter(treatment != 'None') %>% 
  select(patient, id, group, treatment) %>% 
  pivot_wider(names_from = treatment, values_from = id)

de_pre_vs_mid_per_patient <- map(patients$patient, function(which_patient) {
  
  single_patient <- patients %>% filter(patient == which_patient)
  
  cells_per_cluster <- analyze %>% 
    filter(patient == which_patient) %>% 
    group_by(annotation, treatment) %>% 
    tally() %>% 
    pivot_wider(names_from = treatment,
                values_from = n) %>% 
    filter(Pre > cell_threshold,
           Mid > cell_threshold) 
  
  cluster_level <- map(cells_per_cluster$annotation, function(which_cluster) {
    
    print2(paste0(which_patient, ': ', which_cluster))
    
    pre_barcodes <- analyze %>% filter(id == single_patient$Pre, .[[which_annotation]] == which_cluster) %>% pull(rowname)
    mid_barcodes <- analyze %>% filter(id == single_patient$Mid, .[[which_annotation]] == which_cluster) %>% pull(rowname)
    
    de_pipeline(seuratobj = seuratobj,
                cluster_col = which_annotation,
                cluster = which_cluster, 
                group1_barcodes = pre_barcodes,
                group2_barcodes = mid_barcodes,
                group1_id = paste0(which_patient, '_Pre'),
                group2_id = paste0(which_patient, '_Mid'), 
                test.use = which_test) 
  }) %>% bind_rows() %>% 
    mutate(patient = which_patient,
           group = single_patient$group)
   
}) %>% bind_rows() 

de_pre_vs_mid_per_patient %>% write_tsv(here(output_dir, 'de_pre_vs_mid_per_patient.tsv.gz'))

write_excel_notebook(input_table = de_pre_vs_mid_per_patient,
                     output_file = here(output_dir, 'de_pre_vs_mid_per_patient.xlsx'),
                     tabs = 'patient',
                     prefix = 'patient_')



```




### Combined DE table
```{r}
list(#de_ad_vs_hc,
     de_pv_vs_hc,
     #de_adpv_vs_hc,
     #de_ad_vs_pv,
     de_pre_vs_mid,
     de_pre_vs_mid_responder,
     de_pre_vs_mid_nonresponder,
     de_responder_vs_nonresponder_pre,
     de_responder_vs_nonresponder_mid) %>% 
  bind_rows() %>% 
  write_tsv(here(output_dir, 'de_combined.tsv.gz'))
```

## Session Info
```{r}
sessionInfo()
```



