---
title: "DE analysis"
author: "David Wu"
---

Set working directory to project directory
```{r setup}
require(knitr)
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

### Optionally convert this Rmd to R script 
```{r eval=FALSE, message=FALSE}
current_file <- rstudioapi::getActiveDocumentContext()$path
output_file <- stringr::str_replace(current_file, '.Rmd', '.R')
knitr::purl(current_file, output = output_file)
file.edit(output_file)
```

### Directories
```{r, message=FALSE}
output_dir <- 'analysis/output/08_de_analyses' # analysis file output directory
data_dir <- 'data/derived/seurat' # data file output directory

dir.create(output_dir, showWarnings = FALSE)
dir.create(data_dir, showWarnings = FALSE)
```

### Loading the libraries
```{r, message=FALSE}
library(Seurat) 
library(tidyverse)
library(ggExtra)
library(SeuratWrappers)
library(wutilities) # devtools::install_github('symbiologist/wutilities')
library(tictoc)
```

### Import
```{r}
seuratobj <- read_rds(file.path(data_dir, 'seuratobj_annotated.rds'))
meta <- seuratobj@meta.data %>% rownames_to_column()
seuratobj
```

```{r}
patient_meta <- meta %>% 
  select(id, sample, treatment, condition) %>% 
  unique()

patient_meta
```
```{r}
analyze <- meta %>% 
  select(rowname, id, sample, condition, treatment, cluster, merged_label) %>% 
  mutate(description = factor(paste(condition, treatment) %>% str_remove(' None'),
                              levels = c('PV Pre',
                                         'PV Post',
                                         'PV',
                                         'AD',
                                         'HC')))

```

```{r}
de_combined <- read_tsv('analysis/output/07_de/de_combined.tsv.gz')
```
### Is the IL17 difference detected in PV vs HC?
```{r}
de_combined %>% 
  filter(comparison == 'PV vs HC',
         gene %in% c('IL17A', 'IL17F', 'IFNG')) %>% 
  filter(p_val_adj < 0.05) %>% 
  select(comparison, cluster, gene, avg_log2FC, pct.1, pct.2, pct.diff)
```

### What genes in what populations are downregulated with tildra blockade?
```{r}
de_resp
```

### Analyze
#### How many DEGs in each cluster per comparison?
```{r}
comparisons <- de_combined$comparison %>% unique()

deg_tally <- de_combined %>% 
  filter(p_val_adj < 0.05) %>% 
  mutate(direction = ifelse(avg_log2FC > 0, 'group1', 'group2')) %>% 
  group_by(comparison, cluster, direction) %>% 
  tally() 

deg_tally
```
```{r}
deg_tally_wide <- deg_tally %>% 
  pivot_wider(names_from = direction,
              values_from = n) %>% 
  mutate(total = group1 + group2)
```

```{r}
deg_tally %>% 
  ggplot(aes(x = cluster,
             y = n,
             fill = direction)) +
  geom_col(position = 'dodge') +
  theme_dwu() +
  coord_flip() 
```

#### Which clusters are enriched for specific conditions?
```{r}

```


```{r}
trm_labels <- analyze %>% filter(str_detect(merged_label, 'Trm')) %>% pull(merged_label) %>% unique() 
condition1 <- 'PV'
condition2 <- 'HC'
treatment1 <- c('None', 'Pre')
treatment2 <- 'None'
comparison <- 'condition'

p <- analyze %>% 
  filter(condition %in% c(condition1, condition2)) %>% 
  filter(merged_label %in% trm_labels) %>% 
  group_by(merged_label, condition, treatment, id, sample, description) %>% 
  tally() %>% 
  filter(condition == 'PV') %>% 
  arrange(-n) %>% 
  ggplot(aes(x = sample,
             y = n,
             fill = description)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  facet_wrap(~merged_label, nrow = 2, scales = 'free_x') +
  theme_dwu(legend.position = 'right')

p
```



### Percent of all T cells
```{r}
t_cell_clusters <- analyze %>% 
  filter(str_detect(merged_label, 'T')) %>% 
  pull(merged_label) %>% 
  unique()

total_t_cells <- analyze %>% filter(merged_label %in% t_cell_clusters) %>% nrow()

total_t_cells_by_sample <- analyze %>% 
  filter(merged_label %in% t_cell_clusters) %>% 
  group_by(sample, id) %>% 
  tally(name = 'n_t_cells_sample')
```

```{r}
p <- analyze %>% 
  filter(condition %in% c(condition1, condition2)) %>% 
  filter(merged_label %in% trm_labels) %>% 
  group_by(merged_label, condition, treatment, id, sample, description) %>% 
  tally() %>% 
  filter(condition == 'PV') %>% 
  mutate(pct = 100 *n / total_t_cells) %>% 
  arrange(-n) %>% 
  ggplot(aes(x = sample,
             y = pct,
             fill = description)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  facet_wrap(~merged_label, nrow = 2, scales = 'free_x') +
  theme_dwu(legend.position = 'right') +
  labs(y = '% of all T cells',
       x = '')

p
```
```{r}
p <- analyze %>% 
  left_join(total_t_cells_by_sample) %>% 
  filter(condition %in% c(condition1, condition2)) %>% 
  filter(merged_label %in% trm_labels) %>% 
  group_by(merged_label, condition, treatment, id, sample, description, n_t_cells_sample) %>% 
  tally() %>% 
  filter(condition == 'PV') %>% 
  mutate(pct = 100 *n / n_t_cells_sample) %>% 
  arrange(-n) %>% 
  ggplot(aes(x = sample,
             y = pct,
             fill = description)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  facet_wrap(~merged_label, nrow = 2, scales = 'free_x') +
  theme_dwu(legend.position = 'right') +
  labs(y = '% of T cells (per sample)',
       x = '')

p
```

```{r}
p <- analyze %>% 
  filter(condition %in% c(condition1, condition2)) %>% 
  filter(merged_label %in% trm_labels) %>% 
  group_by(label, condition, treatment, id, sample, description) %>% 
  tally() %>% 
  filter(condition == 'PV') %>% 
  mutate(pct = 100 *n / total_t_cells) %>% 
  arrange(-n) %>% 
  ggplot(aes(x = sample,
             y = pct,
             fill = description)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  facet_wrap(~label, nrow = 2, scales = 'free_x') +
  theme_dwu(legend.position = 'right') +
  labs(y = '% of all T cells',
       x = '')

p
```
```{r}



p <- analyze %>% 
  left_join(total_t_cells_by_sample) %>% 
  filter(condition %in% c(condition1, condition2)) %>% 
  filter(merged_label %in% trm_labels) %>% 
  group_by(label, condition, treatment, id, sample, description, n_t_cells_sample) %>% 
  tally() %>% 
  filter(condition == 'PV') %>% 
  mutate(pct = 100 *n / n_t_cells_sample) %>% 
  arrange(-n) %>% 
  ggplot(aes(x = sample,
             y = pct,
             fill = description)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  facet_wrap(~label, nrow = 2, scales = 'free_x') +
  theme_dwu(legend.position = 'right') +
  labs(y = '% of T cells (per sample)',
       x = '')

p
```


### Combining 3 Trm clusters: Trm1, Trm2, CTLac:Trm?
```{r}
trm_sublabels <- c('3 Trm1', '13CTLac:Trm?', '17 Trm2')
p <- analyze %>% 
  filter(condition %in% c(condition1, condition2)) %>% 
  filter(merged_label %in% trm_sublabels) %>% 
  group_by(condition, treatment, id, description) %>% 
  tally() %>% 
  filter(condition == 'PV') %>% 
  mutate(pct = 100 *n / total_t_cells) %>% 
  arrange(-n) %>% 
  ggplot(aes(x = id,
             y = pct,
             fill = description)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  #facet_wrap(~merged_label, nrow = 2, scales = 'free_x') +
  theme_dwu(legend.position = 'right') +
  labs(y = '% of all T cells',
       x = '')

p

```


#### PV vs HC

```{r}
de_pv_vs_hc <- read_tsv('analysis/output/07_de/de_pv_vs_hc.tsv.gz')
```
```{r}
pv_vs_hc %>% 
  filter(str_detect(gene, 'IL17'))
```

```{r}
analyze <- meta %>% 
  select(rowname, id, sample, condition, treatment, cluster, label, merged_label) %>% 
  mutate(description = factor(paste(condition, treatment) %>% str_remove(' None'),
                              levels = c('PV Pre',
                                         'PV Post',
                                         'PV',
                                         'AD',
                                         'HC')))

analyze %>% head()
```

```{r}
condition1 <- 'PV'
condition2 <- 'HC'
treatment1 <- c('None', 'Pre')
treatment2 <- 'None'
comparison <- 'condition'

clusters_pv_vs_hc <- analyze %>% 
  filter(condition %in% c(condition1, condition2),
         treatment %in% c(treatment1, treatment2)) %>% 
  group_by(merged_label, condition, treatment, id) %>% 
  tally() %>% 
  filter(n > 10) %>% # at least 10 cells per sample
  group_by(merged_label, condition, treatment) %>% 
  tally() %>% 
  filter(n > 3) %>% # at least 3 samples per comparison group
  select(merged_label, all_of(comparison)) %>% 
  unique() %>% 
  group_by(merged_label) %>% 
  tally() %>% 
  filter(n > 1)

tic()
de_testing <- map(clusters_pv_vs_hc$merged_label[4], function(i) {
  
  print2(i)
  de_pipeline(seuratobj = seuratobj,
              cluster = i, 
              comparison = comparison,
              condition1 = condition1,
              condition2 = condition2,
              treatment1 = treatment1,
              treatment2 = treatment2,
              test.use = 'Presto', 
              min.pct = 0.01,
              logfc.threshold = log(1.1)) 
  
}) %>% bind_rows()
toc()
```

```{r}
de_testing %>% 
  filter(str_detect(gene, 'IL17|IFNG'))
```

```{r}
de_testing %>% 
  filter(#str_detect(gene, 'IL17|IFNG'),
         p_val_adj < 0.05)
```

```{r}
tic()
de_testing_mast <- map(clusters_pv_vs_hc$merged_label[4], function(i) {
  
  print2(i)
  de_pipeline(seuratobj = seuratobj,
              cluster = i, 
              comparison = comparison,
              condition1 = condition1,
              condition2 = condition2,
              treatment1 = treatment1,
              treatment2 = treatment2,
              test.use = 'MAST', 
              min.pct = 0.01,
              logfc.threshold = log2(1.1)) 
  
}) %>% bind_rows()
toc()
```

```{r}
de_testing <- map(clusters_pv_vs_hc$merged_label[4], function(i) {
  
  print2(i)
  de_pipeline(seuratobj = seuratobj,
              cluster = i, 
              comparison = comparison,
              condition2 = condition2,
              treatment1 = treatment1,
              treatment2 = treatment2,
              test.use = 'Presto', 
              min.pct = 0.05,
              logfc.threshold = 0.1) 
  
}) %>% bind_rows()
```

```{r}
p <- VlnPlot(seuratobj,
        features = c('IL17A', 'IL17F', 'IFNG'),
        group.by = 'merged_label',
        split.by = 'condition')
```

```{r}
data_subset <- FetchData(seuratobj, vars = c('merged_label', 'IL17A', 'IL17F', 'IFNG')) %>% rownames_to_column() %>% inner_join(analyze)
data_subset %>% head()
```

```{r}
data_subset %>% 
  #filter(condition == 'PV') %>% 
  filter(treatment == 'None') %>% 
  filter(condition %in% c('PV', 'HC')) %>% 
  group_by(condition) %>% 
  summarize(n_cells = n(),
            il17a_n = sum(IL17A > 0),
            il17f_n = sum(IL17F > 0),
            ifng_n = sum(IFNG > 0)) 
```
```{r}
p <- data_subset %>% 
  filter(treatment == 'None') %>% 
  filter(condition %in% c('PV', 'HC')) %>% 
  group_by(condition, merged_label) %>% 
  summarize(n_cells = n(),
            IL17A = sum(IL17A > 0),
            IL17F = sum(IL17F > 0),
            IFNG = sum(IFNG > 0)) %>% 
  pivot_longer(cols = -c(condition:n_cells),
               names_to = 'gene',
               values_to = 'n') %>% 
  mutate(pct = 100 * n / n_cells) %>% 
  mutate(n_cells = ifelse(gene == 'IFNG', n_cells, '')) %>% 
  ggplot(aes(x = merged_label,
             y = pct,
             fill = gene,
             label = n_cells)) +
  geom_bar(stat = 'identity') +
  geom_text(aes(y = -10), hjust = 1) +
  ylim(c(-10, 100)) + 
  coord_flip() +
  facet_wrap(~condition + gene) +
  labs(x = '',
       y = '% of cluster where gene is detected') +
  theme_dwu()

p
```


## Session Info
```{r}
sessionInfo()
```



