---
title: "Leave One Out Analysis"
author: "David Wu"
---

Set working directory to project directory
```{r}
library(here)
```

### Optionally convert this Rmd to R script 
```{r eval=FALSE, message=FALSE}
current_file <- rstudioapi::getActiveDocumentContext()$path
output_file <- stringr::str_replace(current_file, '.Rmd', '.R')
knitr::purl(current_file, output = output_file)
file.edit(output_file)
```

### Directories
```{r, message=FALSE}
output_dir <- here('analysis/output/leave_one_out') # analysis file output directory
data_dir <- here('data/derived/tildra') # data file output directory

dir.create(output_dir, showWarnings = FALSE)
dir.create(data_dir, showWarnings = FALSE)
```

### Loading the libraries
```{r, message=FALSE}
library(Seurat) 
library(tidyverse)
library(ggExtra)
library(SeuratWrappers)
library(wutilities) # devtools::install_github('symbiologist/wutilities')
library(tictoc)
library(ggforce)
library(ggthemes)
library(ggradar) # devtools::install_github("ricardo-bion/ggradar")
library(extrafont)
library(patchwork)
library(edgeR)

theme_set(theme_dwu()) # set default theme
```


### Custom functions
```{r}
de_pseudobulk_pipeline <- function(pseudobulk_subset,
                                   metadata_subset,
                                   cluster = 'All',
                                   comparison = NULL,
                                   paired = FALSE,
                                   group1_id,
                                   group2_id) {
  
  if(is.null(comparison)) {
    comparison <- paste(group1_id, 'vs', group2_id) 
  }
  

  metadata_table <- metadata_subset %>% 
    filter(id %in% colnames(pseudobulk_subset)) %>% 
    droplevels() %>% 
    arrange(id, colnames(pseudobulk_subset))
  
  if(paired) {
    design <- model.matrix(data = metadata_table, ~patient + group)  
  } else {
    design <- model.matrix(data = metadata_table, ~group)  
  }
  
  
  dge <- DGEList(counts = pseudobulk_subset,
                 group = metadata_table$group)
  
  #keep <- filterByExpr(dge, min.count = 0, min.prop = 0)
  
  dge <- dge %>% calcNormFactors(method = 'TMM')
  dge <- dge %>% estimateDisp(design = design)
  fit <- dge %>% glmFit(design = design)
  res <- fit %>% glmLRT()
  
  de_table <- topTags(res, n = Inf) %>% 
    as.data.frame() %>% 
    rownames_to_column('gene') %>% 
    as_tibble() %>% 
    mutate(logFC = -1 * logFC, # reverse sign such that group 1 is higher
           cluster = cluster,
           comparison = comparison,
           group1 = group1_id,
           group2 = group2_id,
           n_group1 = metadata_table %>% filter(group == 'group1') %>% nrow(),
           n_group2 = metadata_table %>% filter(group == 'group2') %>% nrow()) %>% 
    select(comparison, cluster, gene, group1, group2, n_group1, n_group2,  avg_log2FC = logFC, p_val = PValue, p_val_adj = FDR, everything())
}
```

### Import
```{r}
seuratobj <- read_rds(here(data_dir, 'seuratobj_annotated.rds'))
seuratobj
```
### Set identity

```{r}
which_identity <- 'local'
which_level <- 'supercluster'
which_annotation <- paste0(which_identity, '_', which_level)

which_cluster <- paste0(which_identity, '_cluster')
which_supercluster <- paste0(which_identity, '_supercluster')
which_ultracluster <- paste0(which_identity, '_ultracluster')

seuratobj$cluster <- seuratobj[[which_cluster]] 
seuratobj$supercluster <- seuratobj[[which_supercluster]] 
seuratobj$ultracluster <- seuratobj[[which_ultracluster]] 
seuratobj$annotation <- seuratobj[[which_annotation]]
Idents(seuratobj) <- 'annotation'

metadata <- seuratobj@meta.data %>% rownames_to_column()
```
### Prepare metadata subset
```{r}
analyze <- metadata %>% 
  select(rowname, id, sample, patient, condition, treatment, annotation, response, group) %>% 
  as_tibble()

analyze %>% head()
```

```{r}
metadata_patient <- metadata %>% 
  select(id, patient, sample, treatment, condition, response, group) %>% 
  unique()

metadata_patient
```


### Global parameters
```{r}
which_test <- 'Presto'
which_annotation <- 'annotation'
sample_threshold <- 2 # at least 3 samples per comparison group (within a cluster)
responder_threshold <- 1 # fewer patients
cell_threshold <- 10 # at least 10 cells per sample (within a cluster)
```

```{r}
patients <- c('P1', 'P2', 'P3', 'P4', 'P5', 'P6')

group1_filters <- list('condition' = 'PV',
                       'treatment' = 'Pre',
                       'group' = 'Responder')

group2_filters <- list('condition' = 'PV',
                       'treatment' = 'Mid',
                       'group' = 'Responder')

data_subset <- bind_rows(
  analyze %>% apply_filters(group1_filters) %>% mutate(group = 'group1'),
  analyze %>% apply_filters(group2_filters) %>% mutate(group = 'group2')
)

de_pre_vs_mid_responder <- map(patients, function(i) {
  
  print2(i)
  
  pseudobulk_subset <- pseudobulk_matrix(seuratobj,
                                         cells = data_subset %>%
                                           filter(annotation == 'Trm',
                                                  patient != i) %>%
                                           pull(rowname),
                                         group_by = 'id')
  
  metadata_subset <- data_subset %>% 
    select(id, patient, group) %>% 
    unique()
  
  res <- de_pseudobulk_pipeline(pseudobulk_subset = pseudobulk_subset, 
                         metadata_subset = metadata_subset,
                         cluster = 'Trm',
                         paired = TRUE,
                         group1_id = 'Responders Pre',
                         group2_id = 'Responders Mid',
                         comparison = i)
  
  res
  
}) %>% bind_rows()

de_pre_vs_mid_responder %>% write_tsv(here(output_dir, 'de_pre_vs_mid_responder_loo.tsv.gz'))
```


## Radar

### Functions
```{r}
calculate_signature <- function(seuratobj, 
                                genes,
                                name = 'signature') {
  
  gene_data <- FetchData(seuratobj, genes)
  as.data.frame(rowSums(gene_data)) %>% set_names(name)
  
}
```

### Import data
#### Visium
```{r}
visiumobj <- read_rds(here(data_dir, 'visium_integrated.rds'))

visium_meta <- visiumobj@meta.data %>% rownames_to_column() %>% 
  mutate(cluster = case_when(seurat_clusters == 3 ~ 'Keratinocyte',
                             seurat_clusters == 2 ~ 'Vascular',
                             seurat_clusters == 1 ~ 'Fibroblast',
                             seurat_clusters == 0 ~ 'Stromal'))
```

### CD4 CD8 Trm
#### CD4 vs CD8 status based on ADT and RNA data
```{r}
cd_data <- FetchData(seuratobj, 
                     vars = c('CD4', 'CD8A', 'CD8B', 'ADT.CD4', 'ADT.CD8', 
                              'cluster', 'supercluster', 'celltype', 'treatment')) %>% 
  filter(celltype == 'Lymphocyte') %>% 
  rownames_to_column() %>% 
  mutate(ADT = case_when(
    ADT.CD4 == 'Positive' & ADT.CD8 == 'Positive' ~ 'Double-positive',
    ADT.CD4 == 'Positive' ~ 'CD4',
    ADT.CD8 == 'Positive' ~ 'CD8',
    TRUE ~ 'Double-negative'
  ))
```


#### Classify CD4/8 status based on protein then RNA
```{r}
cd_status <- cd_data %>% 
  mutate(CD = case_when(
    ADT == 'Double-positive' ~ 'Unknown',
    ADT == 'CD4' ~ 'CD4',
    ADT == 'CD8' ~ 'CD8',
    CD4 > 1 & (CD8A < 1 & CD8B < 1) ~ 'CD4',
    (CD8A > 1 | CD8B > 1) & CD4 < 1 ~ 'CD8',
    TRUE ~ 'Unknown'
  ))

cd_status %>% pull(CD) %>% table()
```


#### Add metadata
```{r}
add_cd <- seuratobj@meta.data %>% 
  rownames_to_column() %>% 
  select(rowname) %>% 
  left_join(cd_status) %>% 
  mutate(cd_supercluster = paste(CD, supercluster))
seuratobj$CD <- add_cd$CD
seuratobj$cd_supercluster <- add_cd$cd_supercluster
```

### DE
```{r}
de_combined <- read_tsv(here(output_dir, 'de_pre_vs_mid_responder_loo.tsv.gz'))
```

```{r}
de_patient_cluster <- read_tsv(here('analysis/output/08_de', which_cluster, 'de_pre_vs_mid_per_patient.tsv.gz')) %>% mutate(level = 'cluster')
de_patient_supercluster <- read_tsv(here('analysis/output/08_de', which_supercluster, 'de_pre_vs_mid_per_patient.tsv.gz')) %>% mutate(level = 'supercluster')
de_patient_ultracluster <- read_tsv(here('analysis/output/08_de', which_ultracluster, 'de_pre_vs_mid_per_patient.tsv.gz')) %>% mutate(level = 'ultracluster')

de_patient <- bind_rows(de_patient_cluster,
                        de_patient_supercluster,
                        de_patient_ultracluster)
```

### Prepare metadata
```{r}
metadata <- seuratobj@meta.data %>% rownames_to_column()

metadata_sample <- metadata %>% 
  select(id, patient, sample, condition, treatment, response, group, patient_label, sample_label) %>% 
  unique() %>% 
  as_tibble()

metadata_sample
```

#### Combine visium and scRNA metadata
```{r}
metadata_combined <- metadata %>% 
  bind_rows(visium_meta %>% select(rowname, id, cluster) %>% drop_na()) %>% 
  select(rowname, 
         id, 
         sample, 
         patient,
         condition, 
         treatment, 
         group, 
         cluster, 
         supercluster, 
         celltype, 
         sample_label, 
         patient_label) %>% 
  mutate(cellgroup = case_when(
    str_detect(cluster, 'Trm1') ~ 'Trm1',
    celltype == 'APC' ~ 'APC',
    str_detect(cluster, 'T') ~ 'Tcell',
    TRUE ~ cluster
  ))

metadata_combined
```

### Major clusters
```{r}
clusters_all <- metadata$annotation %>% unique()
clusters_all
```

## Gene sets
### PV50
```{r}
pv50 <- readxl::read_excel(here('analysis/input/PV50.xlsx'))
pv50_genes <- pv50$...1 %>% intersect(rownames(seuratobj))
```

### Tildra-responsive genes
#### Genes that change in responders, LOO
##### Group-level
```{r}
p_val_threshold <- 0.05
```

```{r}
de_pre_vs_mid_response <- de_combined

il23_group_response <- de_pre_vs_mid_response %>% 
  filter(p_val_adj < p_val_threshold,
         avg_log2FC > 0) 

il23_group_response
```

##### Patient-level
```{r}
de_pre_vs_mid_patients <- de_patient %>% filter(level == which_level)

de_patient_tally <- de_pre_vs_mid_patients %>% 
  filter(p_val_adj < p_val_threshold,
         avg_log2FC > 0) %>% 
  filter(cluster == 'Trm') %>% 
  group_by(cluster, gene, group) %>% 
  tally() %>% 
  pivot_wider(names_from = group,
              values_from = n,
              values_fill = 0)

de_patient_tally
```

```{r}
il23_response <- il23_group_response %>% 
  left_join(de_patient_tally)

il23_response
```

### Response genes per cluster
```{r}
il23_genes_all <- map('Trm' %>% set_names(.), function(i) {
  
  il23_response %>% 
    filter(cluster == i, 
           p_val_adj < p_val_threshold) %>% 
    pull(gene)
  
})

il23_genes_strict <- map('Trm' %>% set_names(.), function(i) {
  
  il23_response %>% 
    filter(cluster == i, 
           p_val_adj < p_val_threshold, 
           Responder > `Non-responder`,
           Responder > 2, 
           `Non-responder` < 2) %>% 
    pull(gene) 
  
})

il23_genes_strict$Trm
```

```{r}
il23_genes_all$Trm
```

```{r}
il23_genes_table <- map(names(il23_genes_all), function(i) {
  
  strict <- il23_genes_strict[[i]]
  
  output <- tibble('cluster' = i,
         'gene' = il23_genes_all[[i]]) 
  
  if(!is.null(strict)) {
    output <- output %>% mutate(Status = ifelse(gene %in% strict, 'Strict', 'All'))
  } else {
    output <- output %>% mutate(Status = 'All')
  }
  
  output %>% mutate(across(everything(), as.character))
  
})  %>% bind_rows()

il23_genes_table %>% write_tsv(here(output_dir, 'il23_genes_table.tsv'))
il23_genes_table
```


### IL17 responsive genes from Harirchian et al. JID 2019
https://pubmed.ncbi.nlm.nih.gov/30543901/
```{r}
il17 <- bind_rows(
  readxl::read_xlsx(here('analysis/input/NIHMS1522377-supplement-2 Cho IL17A.xlsx'), sheet = 'Table S1C', skip = 1) %>% mutate(time = 'T_1'),
  readxl::read_xlsx(here('analysis/input/NIHMS1522377-supplement-2 Cho IL17A.xlsx'), sheet = 'Table S1D', skip = 1) %>% mutate(time = 'T_24'))
colnames(il17)[1] <- 'gene'
il17
```


```{r}
il17_responsive <- il17 %>% 
  filter(logFC > 0.4,
         FDR < 0.05) %>% 
  add_count(gene) %>% filter(n > 1)

il17_responsive
```


```{r}
il17_genes <- il17_responsive$gene %>% unique()
```

#### Other signatures of interest
```{r}
secreted_signature <- c('IL17A', 'IL17F', 'IL26', 'CXCl13', 'GZMB', 'GNLY', 'PCSK7')

il23_peripheral <- c('GNLY',
                     'HGPD',
                     'IL17A',
                     'IL17F',
                     'IL26',
                     'PTPN13',
                     'S100A4')
```

### Extract all expression data
```{r}
extract_expression <- function(seuratobj,
                               cells = NULL,
                               genes,
                               normalize = TRUE,
                               log = FALSE,
                               assay = 'RNA',
                               slot = 'counts') {
  
  data_subset <- FetchData(object = seuratobj, 
                           cells = cells,
                           vars = genes,
                           assay = assay,
                           slot = slot) %>% 
    rownames_to_column() %>% 
    pivot_longer(cols = -rowname,
                 names_to = 'gene',
                 values_to = 'counts')
  
  total_count <- colSums(GetAssayData(seuratobj, 
                                        slot = slot, 
                                        assay = assay))
  
  if(!is.null(cells)) {
    total_count <- total_count[cells]
  }
  
  data_subset <- data_subset %>% 
    left_join(data.frame('total' = total_count) %>% rownames_to_column(), by = 'rowname')
  
  if(normalize) {
   data_subset <- data_subset %>% 
      mutate(cpm = 1e6 * counts / total) 
  }
  
  data_subset
  
}

pseudobulk <- function(seuratobj,
                       genes,
                       cells = NULL,
                       group_by = 'ident',
                       pseudocount = 1,
                       rounding = 3,
                       assay = 'RNA',
                       slot = 'count') {
  
  metadata_groups <- FetchData(seuratobj, vars = group_by, cells = cells) %>% rownames_to_column()
  
  expression <- extract_expression(seuratobj,
                                   genes = genes,
                                   cells = cells,
                                   normalize = FALSE,
                                   log = FALSE,
                                   assay = assay,
                                   slot = slot) %>% 
    left_join(metadata_groups, by = 'rowname')
  
  totals <- expression %>% 
    select(rowname, total, all_of(group_by)) %>% 
    unique() %>% 
    group_by(across(all_of(group_by))) %>% 
    summarize(total = sum(total))
  
  merged <- expression %>% 
    group_by(across(all_of(c('gene', group_by)))) %>% 
    summarize(counts = sum(counts)) %>% 
    ungroup() %>% 
    left_join(totals) %>% 
    mutate(cpm = counts/total * 1e6,
           log10cpm = log10(cpm + pseudocount))
  
  merged %>% mutate(across(contains('cpm'), round, rounding))
  
}

plot_radar <- function(df,
                       group_by = NULL,
                       features = NULL,
                       colors = NULL,
                       title = '',
                       legend.position = 'top') {
  
  if(is.null(group_by)) {
    
    group_by <- colnames(df)[1]
    
  }
    
  if(is.null(features)) {
    
    features <- colnames(df) %>% setdiff(group_by)
    
  }
  
  radar_input <- df %>% 
    select(all_of(c(group_by, features))) %>% 
    mutate(across(all_of(features), scales::rescale_max))
  
  radar_input[[group_by]] <- droplevels(radar_input[[group_by]])
  
  radar_input %>% 
    ggradar(legend.text.size = 8,
            grid.label.size = 0,
            axis.label.size = 2,
            group.point.size = 2,
            axis.label.offset = 1.1,
            group.line.width = 1,
            group.colours = colors,
            plot.title = title,
            legend.position = legend.position) +
    theme(plot.title = element_text(size = 9, hjust = 0.5),
          plot.background = element_rect(color = 'black'),
          #legend.box.background = element_rect(color = 'blue'),
          legend.key.height = unit(1, 'mm'),
          legend.spacing = unit(0, 'mm')) 
}

plot_radar_patient <- function(pseudo,
                               cluster,
                               patient,
                               genes,
                               normalize = TRUE,
                               scale = c('within', 'across'),
                               colors = c('dodgerblue4', 'orangered'),
                               test = TRUE,
                               significance_colors = c('orangered', 'red4'),
                               significance_cutoff = 0.01,
                               title = 'patient') {
  
  plot_title <- pseudo %>% filter(patient == {{patient}}) %>% pull({{title}}) %>% unique()
  
  plot_input <- pseudo %>% 
    filter(cluster %in% {{cluster}}, 
           gene %in% {{genes}}) %>% 
    select(patient,
           feature = gene,
           treatment,
           normalized = cpm)
  
  # significance testing
  if(test) {
    res <- test_pre_vs_mid(pseudo = {{pseudo}},
                           cluster = {{cluster}},
                           genes = {{genes}},
                           patient = {{patient}})
    
    if(res$wilcox$p.value < 0.001) {
      subtitle <- paste0('p < 0.001')
    } else {
      subtitle <- paste0('p = ', round(res$wilcox$p.value, 3))  
    }
    
    
    # if(res$t$p.value < significance_cutoff) {
    #   plot_title <- paste0(plot_title, '*')
    # }
    
    if(res$wilcox$p.value < significance_cutoff) {
      plot_title <- paste0(plot_title, '*')
    }
    
    # color title based on significance
    n_significant <- str_count(plot_title, '\\*')
    if(n_significant == 0) {
      title_color <- 'black'
    } else {
      title_color <- significance_colors[n_significant]
    }
    
  } else {
    
    subtitle <- NULL
  
    }
  
  if(scale == 'across') {
    
    plot_input <- plot_input %>% 
      group_by(feature) %>% 
      mutate(feature_max = max(normalized),
             normalized = ifelse(feature_max > 0, normalized/feature_max, 0)) %>% 
      filter(patient == {{patient}})
    
  } else if(scale == 'within') {
    
    plot_input <- plot_input %>% 
      filter(patient == {{patient}}) %>% 
      group_by(feature) %>% 
      mutate(feature_max = max(normalized),
             normalized = ifelse(feature_max > 0, normalized/feature_max, 0))
  }
  
  radar_input <- plot_input %>% 
    mutate(treatment = droplevels(treatment)) %>% 
    select(treatment, feature, normalized) %>% 
    pivot_wider(names_from = feature,
                values_from = normalized)
  
  p_radar <- radar_input %>% 
    ggradar(legend.text.size = 8,
            grid.label.size = 0,
            axis.label.size = 2,
            group.point.size = 2,
            axis.label.offset = 1.1,
            group.line.width = 1,
            group.colours = colors,
            legend.position = 'top', 
            font.radar = 'Arial') +
    theme(text = element_text(family = 'Arial'),
          plot.title = element_text(size = 9, hjust = 0.5, color = title_color),
          plot.subtitle = element_text(size = 8, hjust = 0.5),
          plot.background = element_rect(color = 'black'),
          legend.key.height = unit(1, 'mm'),
          legend.spacing = unit(0, 'mm')) +
    labs(title = plot_title,
         subtitle = subtitle)
  
  p_radar
  
  
}

test_pre_vs_mid <- function(pseudo,
                            patient,
                            cluster,
                            genes,
                            logtransform = TRUE,
                            pseudocount = 1,
                            binarize = TRUE,
                            cutoff = 0.5) {
  
  test_input <- pseudo %>% 
    filter(cluster %in% {{cluster}}, 
           gene %in% {{genes}},
           patient == {{patient}}) %>% 
    select(gene,
           treatment,
           expression = cpm) %>% 
    rowwise() %>% 
    mutate(expression = ifelse(logtransform, log10(expression + 1), expression)) %>% 
    ungroup() %>% 
    pivot_wider(names_from = 'treatment',
                values_from = 'expression')
  
  if(binarize) {
    test_input <- test_input %>% rowwise() %>% 
      mutate(Max = max(Pre, Mid), 
             Pre = as.integer(Pre > cutoff * Max), 
             Mid = as.integer(Mid > cutoff * Max)) %>% 
      ungroup() %>% 
      select(-Max)
    
  }
  
  if((test_input$Pre == test_input$Mid)) {
    t_res <- list(p.value = 1)
    wilcox_res <- list(p.value = 1)
  } else {
    t_res <- t.test(test_input$Pre,
                    test_input$Mid,
                    paired = TRUE)
    
    wilcox_res <- wilcox.test(test_input$Pre,
                              test_input$Mid,
                              paired = TRUE)
  } 
  
  list('t' = t_res,
       'wilcox' = wilcox_res)
}


```

```{r}
tildra_barcodes <- metadata %>% filter(treatment %in% c('Pre', 'Mid'), annotation == 'Trm') %>% pull(rowname)
genes_core <- c('IL17A', 'IL17F', 'IL26', 'IFNG')
gene_set <- c('IL17A', 'IL17F', 'IL26', 'TNF', 'IFNG', 'IL23R', 'IL21', 'IL22', 'IL36A', 'CXCL13')

pseudo <- pseudobulk(seuratobj,
                     cells = tildra_barcodes,
                     genes = c(genes_core, gene_set, il23_genes_all %>% unlist() %>% unique()) %>% unique(),
                     group_by = c(which_level, 'id')) %>% 
  left_join(metadata_sample %>% select(id, patient, treatment, group, sample_label, patient_label)) %>% 
  dplyr::rename(cluster = {{which_level}})

pseudo 
```

### Radar plot

```{r}

# p <- plot_radar_patient(pseudo = pseudo,
#                    cluster = 'Mac',
#                    patient = 'P1',
#                    genes = c(genes_core, il23_genes_strict$Mac),
#                    scale = 'across')
# 
# ggsave(plot = p,
#        h = 10,
#        w = 10,
#        device = cairo_pdf,
#        path = output_dir,
#        filename = 'test.pdf')
# 
# p
```

#### Limited set
```{r}
p <- map(pseudo$patient %>% unique() %>% sort() %>% as.character(), function(which_patient) {
  plot_radar_patient(pseudo = pseudo,
                     cluster = 'Trm',
                     patient = which_patient,
                     genes = gene_set,
                     scale = 'within')
}) %>% wrap_plots(ncol = 3)

ggsave(plot = p,
       h = 10,
       w = 10,
       device = cairo_pdf,
       path = output_dir,
       filename = 'radar_cytokines.pdf')

p
```
```{r}
il23_genes_all$Trm %>% sort() %>% as_tibble() %>% View()
```


#### Loop
```{r}
update_geom_defaults('text', list(fontface = 'italic'))
which_cluster <- 'Trm'
map(patients, function(i) {
  
  # gene_core <- NULL
  # genes_all <- c(genes_core, il23_genes_all[[which_cluster]]) %>% unique()
  # genes_strict <- c(genes_core, il23_genes_strict[[which_cluster]]) %>% unique()
  
  genes_all <- il23_group_response %>% filter(comparison == i) %>% pull(gene) %>% unique()
  
  map(c('within'), function(which_scale) {
    
    if(length(genes_all) > length(genes_core)) {
      
      print2(paste(which_cluster, which_scale))
      
      p <- map(pseudo$patient %>% unique() %>% sort() %>% as.character(), function(which_patient) {
        
        
        if((pseudo %>% 
            filter(patient == which_patient,
                   cluster == which_cluster,
                   gene %in% genes_all,
                   cpm > 0) %>% 
            nrow() > 0)) {
          
          plot_radar_patient(pseudo = pseudo,
                             cluster = which_cluster,
                             patient = which_patient,
                             genes = genes_all,
                             scale = which_scale)
        } else {
          ggplot()
        }
      }) %>% wrap_plots() 
      
      print2('Saving')
      dir.create(here(output_dir, 'radar_all'), showWarnings = FALSE)
      ggsave(plot = p,
             h = 10,
             w = 10,
             device = cairo_pdf,
             path = here(output_dir, 'radar_all'),
             filename = paste0(which_scale, '_', str_remove(which_cluster, '/'), '_leave_out_', i,  '.pdf'))
      
    }
    
    # if(length(genes_strict) > length(genes_core)) {
    #   
    #   p <- map(pseudo$patient %>% unique() %>% sort(), function(which_patient) {
    #     
    #     
    #     if((pseudo %>% filter(patient == which_patient,
    #                           cluster == which_cluster,
    #                           gene %in% genes_strict,
    #                           cpm > 0) %>% 
    #         nrow() > 0)) {
    #       
    #       plot_radar_patient(pseudo = pseudo,
    #                          cluster = which_cluster,
    #                          patient = which_patient,
    #                          genes = genes_strict,
    #                          scale = which_scale)
    #     } else {
    #       ggplot()
    #     }
    #     
    #     
    #   }) %>% wrap_plots()
    #   
    #   dir.create(here(output_dir, 'radar_strict'), showWarnings = FALSE)
    #   
    #   ggsave(plot = p,
    #          h = 10,
    #          w = 10,
    #          device = cairo_pdf,
    #          path = here(output_dir, 'radar_strict'),
    #          filename = paste0(which_scale, '_', str_remove(which_cluster, '/'), '.pdf'))
    # }
  }) %>% invisible()
}) %>% invisible()

update_geom_defaults('text', list(fontface = 1))

```

### Stats
```{r}
which_cluster <- 'Trm'
genes_all <- c(genes_core, il23_genes_all[[which_cluster]]) %>% unique()
genes_strict <- c(genes_core, il23_genes_strict[[which_cluster]]) %>% unique()
  
stats_input <- pseudo %>% 
  filter(cluster == which_cluster,
         gene %in% genes_all) %>% 
  select(gene, patient, treatment, group, cpm) %>% 
  pivot_wider(names_from = treatment,
              values_from = cpm) %>% 
  mutate(ratio = (Pre - Mid)/Pre) %>% 
  arrange(gene, patient) %>% 
  group_by(gene, group) %>% 
  summarize(ratio = median(ratio)) %>% 
  pivot_wider(names_from = group,
              values_from = ratio)

stats_input
```
```{r}
stats_input %>% 
  ungroup() %>% 
  drop_na() %>% 
  summarize(across(where(is.numeric), median))
```

```{r}
wilcox.test(stats_input$Responder,
            stats_input$`Non-responder`,
            paired = TRUE)
```

### patient-specific
```{r}
pseudo %>% 
  filter(cluster == which_cluster,
         gene %in% genes_all) %>% 
  select(gene, patient, treatment, group, cpm) %>% 
  pivot_wider(names_from = treatment,
              values_from = cpm) %>% 
  mutate(ratio = (Pre - Mid)/Pre) %>% 
  arrange(gene, patient) %>% 
  group_by(patient) %>% 
  summarize(mean = round(mean(ratio, na.rm = T), 3),
            median = round(median(ratio, na.rm = T), 3))
```
```{r}
pseudo %>% 
  filter(cluster == which_cluster,
         gene %in% genes_all) %>% 
  select(gene, patient, treatment, group, cpm) %>% 
  pivot_wider(names_from = treatment,
              values_from = cpm) %>% 
  mutate(ratio = (Pre - Mid)/Pre) %>% 
  filter(patient == 'P9') %>% 
  filter(Mid > Pre)
```


```{r}
extract_signature <- function(seuratobj,
                              cells = NULL,
                              genes,
                              name = 'signature',
                              normalize = TRUE,
                              log = FALSE,
                              assay = 'RNA',
                              slot = 'counts') {
  
  seuratobj <- seuratobj %>% add_signature(genes = genes, name = name)
  
}

add_signature <- function(seuratobj, 
                          genes,
                          name = 'signature') {
  
  seuratobj <- seuratobj %>% PercentageFeatureSet(features = genes, col.name = name)
  seuratobj[[name]] <- seuratobj[[name]] * seuratobj[['nCount_RNA']]/100
  
  seuratobj
}
```


### IL23-responsive gene expression in mid-treatment samples 
Strict
```{r}
p <- data_table_long_tx_summarized %>% 
  filter(gene %in% c(il23_genes_strict, 'IL17A', 'IL17F'),
         treatment == 'Mid',
         cellgroup == 'Trm1') %>% 
  ggplot(aes(x = reorder(gene, expression),
             y = expression,
             color = group)) +
  #geom_boxplot(outlier.color = NA) +
  geom_bar(stat = 'summary', position = 'dodge', fill = 'grey90') +
  #facet_wrap(~gene, scales = 'free_x') +
  theme_dwu(legend.position = 'top') +
  scale_color_manual(values = c('orangered', 'dodgerblue4')) +
  geom_jitter(alpha = 0.8, size = 2) +
  coord_flip() +
  labs(x = '',
       y = 'Trm1 Average Expression')

ggsave(plot = p,
       filename = 'bar_mid_separators_strict.pdf',
       path = output_dir,
       h = 14,
       w = 6)

p

```
Full
```{r}
p <- data_table_long_tx_summarized %>% 
  
  filter(gene %in% c(il23_genes_all, 'IL17A', 'IL17F'),
         treatment == 'Mid',
         cellgroup == 'Trm1') %>% 
  ggplot(aes(x = reorder(gene, expression),
             y = expression,
             color = group)) +
  #geom_boxplot(outlier.color = NA) +
  geom_bar(stat = 'summary', position = 'dodge', fill = 'grey90') +
  geom_sina(alpha = 0.8, size = 2) +
  #facet_wrap(~gene, scales = 'free_x') +
  theme_dwu(legend.position = 'top') +
  scale_color_manual(values = c('orangered', 'dodgerblue4')) +
  coord_flip() +
  labs(x = '',
       y = 'Trm1 Average Expression')

ggsave(plot = p,
       filename = 'bar_mid_separators_full.pdf',
       path = output_dir,
       h = 44,
       w = 6)

p
```
### Facets
```{r}
p <- data_table_long_tx_summarized %>% 
  
  filter(gene %in% c(il23_genes_all, 'IL17A', 'IL17F'),
         treatment == 'Mid',
         cellgroup == 'Trm1') %>% 
  ggplot(aes(x = group,
             y = expression,
             color = group)) +
  #geom_boxplot(outlier.color = NA) +
  geom_bar(stat = 'summary', position = 'dodge', fill = 'grey90') +
  geom_point(alpha = 0.8, size = 2, aes(color = patient)) +
  facet_wrap(~gene, scales = 'free_y', ncol = 15) +
  theme_dwu(legend.position = 'top') +
  rcartocolor::scale_color_carto_d() +
  #scale_color_manual(values = c('orangered', 'dodgerblue4')) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(x = '',
       y = 'Trm1 Average Expression')

ggsave(plot = p,
       filename = 'bar_mid_separators_facet_full.pdf',
       path = output_dir,
       h = 30,
       w = 17)
```

```{r}
p <- data_table_long_tx_summarized %>% 
  
  filter(gene %in% c(il23_genes_strict, 'IL17A', 'IL17F'),
         treatment == 'Mid',
         cellgroup == 'Trm1') %>% 
  ggplot(aes(x = group,
             y = expression,
             color = group)) +
  #geom_boxplot(outlier.color = NA) +
  geom_bar(stat = 'summary', position = 'dodge', fill = 'grey90') +
  geom_point(alpha = 0.8, size = 2, aes(color = patient)) +
  facet_wrap(~gene, scales = 'free_y', ncol = 10) +
  theme_dwu(legend.position = 'top') +
  #scale_color_manual(values = c('orangered', 'dodgerblue4')) +
  rcartocolor::scale_color_carto_d(palette = 'Prism') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(x = '',
       y = 'Trm1 Average Expression')

ggsave(plot = p,
       filename = 'bar_mid_separators_facet_strict.pdf',
       path = output_dir,
       h = 14,
       w = 12)
```




#### Secretion
```{r}
p <- data_table_long_tx_summarized %>% 
  
  filter(gene %in% 'Secreted',
         treatment == 'Mid',
         cellgroup == 'Trm1') %>% 
  ggplot(aes(x = reorder(patient, -expression),
             y = expression,
             fill = group)) +
  #geom_boxplot(outlier.color = NA) +
  geom_bar(stat = 'summary', position = 'dodge', color = 'black') +
  theme_dwu(legend.position = 'top') +
  #rcartocolor::scale_color_carto_d(palette = 'Prism') +
  scale_fill_manual(values = c('orangered', 'dodgerblue4')) +
  labs(x = 'Patient',
       y = 'Trm1 Average Expression',
       title = 'Secreted Gene Signature')

p
```

#### Peripheral Th17 IL23
```{r}


p <- data_table_long_tx_summarized %>% 
  
  filter(gene == 'IL23_peripheral',
         treatment == 'Mid',
         cellgroup == 'Trm1') %>% 
  ggplot(aes(x = reorder(patient, -expression),
             y = expression,
             fill = group)) +
  #geom_boxplot(outlier.color = NA) +
  geom_bar(stat = 'summary', position = 'dodge', color = 'black', fun = sum) +
  theme_dwu(legend.position = 'top') +
  #rcartocolor::scale_color_carto_d(palette = 'Prism') +
  scale_fill_manual(values = c('orangered', 'dodgerblue4')) +
  labs(x = 'Patient',
       y = 'Trm1 Average Expression',
       title = 'Peripheral IL23 Gene Signature')

p
```

### Statistics
```{r}

```
### T vs APC
```{r}
p <- data_table_long %>% 
  filter(treatment != 'None',
         gene %in% c('IL17A', 'IL17F'),
         group %in% c('APC', 'Trm1', 'Tcell')) %>% 
  group_by(patient, treatment, group, group, gene) %>% 
  summarize(expression = mean(expression)) %>% 
  ggplot(aes(x = patient,
             y = expression,
             fill = group)) +
  geom_bar(stat = 'identity') +
  facet_wrap(~group + treatment + gene) +
  theme_dwu(legend.position = 'top')
  
p
```


### PV, AD, HC
```{r fig.height=6, fig.width=12}
display_features <- c('APC IL23A',
                      'Trm1 IL23R',
                      paste('Trm1', il23_genes_all),
                      'Trm1 IL17A',
                      'Trm1 IL17F',
                      'APC IL17A',
                      'APC IL17F',
                      'Trm1 ZFP36',
                      'Trm1 ZFP36L2') %>% 
  unique()

plot_input <- data_table_long_tx_summarized %>% 
  mutate(feature = paste(cellgroup, group)) %>% 
  filter(feature %in% display_features) %>% 
  mutate(feature = factor(feature, levels = display_features))

p <- plot_input %>% 
  ggplot(aes(x = feature,
             y = normalized,
             fill = condition)) +
  geom_boxplot()+
  theme_dwu(legend.position = 'top') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(y = 'Normalized Expression',
       x = '') +
  ggthemes::scale_fill_few()

p
```

```{r}

metadata_combined %>% 
  filter(cluster %in% clusters_apc,
         treatment != 'None') %>% 
  group_by(group, cluster, treatment) %>% 
  tally() %>% 
  pivot_wider(names_from = c(group, treatment),
              values_from = n) %>% 
  View()

```

### CD4 vs CD8
```{r}
tildra_barcodes <- metadata %>% filter(treatment %in% c('Pre', 'Mid')) %>% pull(rowname)
genes_core <- c('IL17A', 'IL17F', 'IL26', 'IFNG')
gene_set <- c('IL17A', 'IL17F', 'IL26', 'TNF', 'IFNG', 'IL23R', 'IL21', 'IL22', 'IL36A', 'CXCL13')

pseudo <- pseudobulk(seuratobj,
                     cells = tildra_barcodes,
                     genes = c(genes_core, gene_set, il23_genes_all$Trm %>% unlist()) %>% unique(),
                     group_by = c('cd_supercluster', 'id')) %>% 
  left_join(metadata_sample %>% select(id, patient, treatment, group, sample_label, patient_label)) %>% 
  dplyr::rename(cluster = cd_supercluster)

pseudo 
```

```{r}
update_geom_defaults('text', list(fontface = 'italic'))


map(c('CD4 Trm', 'CD8 Trm'), function(which_cluster) {
  
  #gene_core <- NULL
  genes_all <- c(genes_core, il23_genes_all[['Trm']]) %>% unique()
  
  which_scale <- 'within'
    
    if(length(genes_all) > length(genes_core)) {
      
      print2(paste(which_cluster, which_scale))
      
      p <- map(pseudo$patient %>% unique() %>% sort() %>% as.character(), function(which_patient) {
        
        
        if((pseudo %>% 
            filter(patient == which_patient,
                   cluster == which_cluster,
                   gene %in% genes_all,
                   cpm > 0) %>% 
            nrow() > 0)) {
          
          plot_radar_patient(pseudo = pseudo,
                             cluster = which_cluster,
                             patient = which_patient,
                             genes = genes_all,
                             scale = which_scale)
        } else {
          ggplot()
        }
      }) %>% wrap_plots() 
      
      print2('Saving')
      dir.create(here(output_dir, 'radar_all'), showWarnings = FALSE)
      ggsave(plot = p,
             h = 10,
             w = 10,
             device = cairo_pdf,
             path = here(output_dir, 'radar_all'),
             filename = paste0(which_scale, '_', str_remove(which_cluster, '/'), '.pdf'))
      
    }
    
  
  }) %>% invisible()

update_geom_defaults('text', list(fontface = 1))
```

### Stats
```{r}
which_cluster <- 'CD4 Trm'
  
stats_input <- pseudo %>% 
  filter(cluster == which_cluster,
         gene %in% genes_all) %>% 
  select(gene, patient, treatment, group, cpm) %>% 
  pivot_wider(names_from = treatment,
              values_from = cpm) %>% 
  mutate(ratio = (Pre - Mid)/Pre) %>% 
  arrange(gene, patient) %>% 
  group_by(gene, group) %>% 
  summarize(ratio = median(ratio)) %>% 
  pivot_wider(names_from = group,
              values_from = ratio)

stats_input
```
```{r}
stats_input %>% 
  ungroup() %>% 
  drop_na() %>% 
  summarize(across(where(is.numeric), median))
```

```{r}
wilcox.test(stats_input$Responder,
            stats_input$`Non-responder`,
            paired = TRUE)
```

### patient-specific
```{r}
reduction_summary <- pseudo %>% 
  filter(gene %in% genes_all) %>% 
  filter(cluster %in% c('CD4 Trm', 'CD8 Trm')) %>% 
  select(gene, patient, treatment, group, cpm, cluster) %>% 
  pivot_wider(names_from = treatment,
              values_from = cpm) %>% 
  mutate(ratio = (Pre - Mid)/Pre) %>% 
  arrange(gene, patient, cluster) %>% 
  group_by(patient, cluster) %>% 
  summarize(mean = round(mean(ratio, na.rm = T), 3),
            median = round(median(ratio, na.rm = T), 3)) 
reduction_summary %>% 
  select(-mean) %>% 
  pivot_wider(names_from = cluster,
              values_from = median)
```

```{r}
p <- reduction_summary %>% 
  ggplot(aes(x = -median,
             y = factor(patient, levels = rev(levels(patient))),
             fill = cluster)) +
  geom_col(position = 'dodge',
           color = 'black') +
  scale_fill_few() +
  theme(legend.position = 'right') +
  labs(x = 'Median % Change Pre to Mid',
       y = NULL)
  
ggsave(plot = p,
       device = cairo_pdf,
       path = output_dir,
       h = 4,
       w = 4,
	filename = 'S3A_cd4_cd8.pdf')
  
```

```{r}
pseudo %>% filter(cluster == 'Trm') %>% select(-cluster) %>% 
  write_tsv(here(output_dir,'pseudobulk_table.tsv.gz'))
```


## Session Info
```{r}
sessionInfo()
```



