---
title: "Annotation"
author: "David Wu"
---

### Use the here package for robust relative paths
```{r}
library(here)
```

### Optionally convert this Rmd to R script 
```{r eval=FALSE, message=FALSE}
current_file <- rstudioapi::getActiveDocumentContext()$path
output_file <- stringr::str_replace(current_file, '.Rmd', '.R')
knitr::purl(current_file, output = output_file)
file.edit(output_file)
```

### Directories
```{r, message=FALSE}
output_dir <- 'analysis/output/07_annotate' # analysis file output directory
data_dir <- 'data/derived/tildra' # data file output directory

dir.create(output_dir, showWarnings = FALSE)
dir.create(data_dir, showWarnings = FALSE)
```

### Loading the libraries
```{r, message=FALSE}
library(Seurat) 
library(tidyverse)
library(ggExtra)
library(SeuratWrappers)
library(wutilities) # devtools::install_github('symbiologist/wutilities')
library(tictoc)
library(ggsankey)
library(aricode)
theme_set(wutilities::theme_dwu())
```

### Import
#### Clustered object
```{r}
seuratobj <- read_rds(here(data_dir, 'seuratobj_subcluster.rds'))
metadata <- seuratobj@meta.data %>% rownames_to_column()
seuratobj
```

#### Import annotations from RashX paper
```{r}
rashx_clusters <- read_tsv(here('analysis/output/rashx/rashx_clusters.tsv.gz'))
rashx_clusters %>% head()
```
#### Import transferred labels
```{r}
labels <- read_tsv(here('analysis/output/06_label/transfer.tsv.gz')) %>% select(rowname, label = predicted.id)

label_transfer <- metadata %>% select(rowname, local, global) %>% left_join(labels)
label_transfer %>% head()
```
#### Which label to use?
```{r}
clusters <- c('local', 'global') %>% set_names(.)
#labels <- c('scibet', 'seurat') %>% set_names(.)
labels <- c('seurat200', 'seurat30') %>% set_names(.)

label_odds <- map(clusters, function(i) {
  map(labels, function(j) {
    
    odds_table(label_transfer %>% select(all_of(c(i, j))) %>% drop_na(), 
               group1 = i,
               group2 = j) %>% 
      group_by(group1) %>% 
      slice_max(f_group1) %>% 
      select(group1, group2) %>% 
      mutate('cluster' = i,
             'label' = j)
  }) %>% bind_rows()
}) %>% bind_rows() %>% 
  ungroup()

label_match <- label_odds %>% 
  pivot_wider(names_from = 'label',
              values_from = 'group2') %>% 
  mutate(match = ifelse(seurat200 == seurat30, 'Yes', 'No'))

label_match
```
```{r}
label_match %>% filter(match == 'No')
```

```{r}
label_match %>% 
  group_by(cluster) %>% 
  summarize(n_total = n(),
            n_match = sum(match == 'Yes')) %>% 
  mutate(f_match = n_match / n_total)
```
```{r}
celltype <- 'Trm1'

p <- plot_sankey(label_match %>% filter(cluster == 'local', (seurat %in% celltype) | (scibet %in% celltype)),
                 nodes = c('scibet', 'group1', 'seurat'),
                 reverse = FALSE)

p
```

```{r}
plot_sankey <- function(df,
                        nodes,
                        names = NULL,
                        reverse = TRUE,
                        total = TRUE,
                        prefix = 'Interactions:',
                        arrange = 1,
                        title = NULL,
                        drop_na = FALSE) {
  
  sankey_input <- df %>% select(all_of(nodes)) %>% mutate(across(.fns = as.character)) %>% arrange(across(nodes[arrange]))
  
  if(drop_na) {
    sankey_input <- sankey_input %>% drop_na()
  }
  
  if(is.null(names)) {
    names <- nodes
  } else {
    colnames(sankey_input) <- names
  }
  
  if(reverse) {
    sankey_input <- sankey_input %>% make_long(length(nodes):1)
  } else {
    sankey_input <- sankey_input %>% make_long(1:length(nodes))
  }
  
    
  if(total) {
    total <- paste(prefix, nrow(df))
  } else {
    total <- NULL
  }
  
  # Construct plot
  p <- ggplot(sankey_input %>% mutate(node = factor(node)), 
              aes(x = x, 
                  next_x = next_x, 
                  node = node, 
                  next_node = next_node,
                  fill = node,
                  label = node)) +
    geom_sankey(flow.alpha = 0.6,
                flow.color = 'black',
                node.color = 'black') +
    theme_sankey(base_size = 10, base_family = 'Arial') +
    theme(legend.position = 'none',
          plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.text.x = element_text(size = 10, color = 'black'))+
    geom_sankey_label(size = 3, color = "black", fill = "white", family = 'Arial') + 
    scale_fill_viridis_d() + 
    labs(x = '',
         title = title,
         subtitle = total)
  
}
```



## Analysis
### Combine all annotations
```{r}
merged_annotations <- metadata %>% 
  select(rowname, id, local, global) %>% 
  left_join(label_transfer %>% select(rowname, label = seurat200)) %>% 
  separate(rowname, sep = '_', into = c('barcode', NA), remove = FALSE) %>% 
  left_join(rashx_clusters %>% select(barcode, rashx = cluster)) %>% 
  as_tibble()

merged_annotations
```
### Agreement metrics
```{r}
clusters <- c('local', 'global') %>% set_names(.)
annotations <- c('label', 'rashx') %>% set_names(.)

all_clusterings <- c(clusters, annotations)

pairwise_table <- crossing('cluster1' = all_clusterings,
                           'cluster2' = all_clusterings) %>% 
  filter(cluster1 != cluster2) %>% 
  rowwise() %>% 
  mutate(combination = paste(sort(c(cluster1, cluster2)), collapse = ':')) %>% 
  group_by(combination) %>% 
  slice_max(cluster1) %>% 
  arrange(cluster1)
  
pairwise_table
```

```{r}
agreement_metrics <- map(1:nrow(pairwise_table), function(i) {
  row_input <- pairwise_table[i, ]
  
  cluster_input <- merged_annotations %>% select(all_of(c(row_input$cluster1, row_input$cluster2))) %>% drop_na()
  
  tibble('combination' = row_input$combination,
         calculate_agreement(cluster_input[[row_input$cluster1]], 
                             cluster_input[[row_input$cluster2]]) %>% 
    mutate(across(everything(), round, 3)))
  
}) %>% 
  bind_rows() %>% 
  rowwise() %>% 
  mutate(mean = mean(c(AMI, ARI))) %>% 
  arrange(-mean) %>% 
  ungroup()

agreement_metrics
```

### Calculate odds and overlap
```{r}
annotation_odds <- map(clusters, function(i) {
  map(annotations, function(j) {
    
    odds_table(merged_annotations %>% select(all_of(c(i, j))) %>% drop_na(), 
               group1 = i,
               group2 = j)
  })
})

annotation_odds$local$label
```



```{r}
input_odds <- annotation_odds$local$label

top_jaccard <- input_odds %>% 
  group_by(group1) %>% 
  slice_max(jaccard)

top_overlap <- input_odds %>% 
  group_by(group1) %>% 
  slice_max(overlap_coef)

top_f <- input_odds %>% 
  group_by(group1) %>% 
  slice_max(f_group1)

top_f_score<- input_odds %>% 
  group_by(group1) %>% 
  slice_max(f_score)

top_odds <- input_odds %>% 
  group_by(group1) %>% 
  slice_max(odds)

top_comparison <- top_f %>% select(group1, f_group1, top_f_group1 = group2) %>% 
  left_join(top_overlap %>% select(group1, overlap_coef, top_overlap = group2)) %>% 
  left_join(top_jaccard %>% select(group1, jaccard, top_jaccard = group2)) %>% 
  left_join(top_f_score %>% select(group1, f_score, top_f_score = group2)) %>% 
  left_join(top_odds %>% select(group1, odds, top_odds = group2)) %>% 
  mutate(match = ifelse(top_f_group1 == top_jaccard, 'Yes', 'No')) 

top_comparison %>% add_count(group1) %>% filter(match == 'No')
```

### Sankey plots
```{r}
dpi <- 300

sankey1_loop <- crossing('cluster1' = all_clusterings,
                         'cluster2' = all_clusterings) %>% 
  filter(cluster1 != cluster2) %>% 
  left_join(merged_annotations %>% select(all_of(all_clusterings)) %>% unique() %>% 
    pivot_longer(cols = everything(),
                 names_to = 'cluster1',
                 values_to = 'celltype') %>% 
    unique() %>% 
    drop_na() %>% 
    arrange(cluster1, celltype))

sankey1_loop
  
```

```{r}
library(parallel)
cores <- 20

tic()
mclapply(1:nrow(sankey1_loop), mc.cores = cores, FUN = function(i) {
  
  parameters <- sankey1_loop[i,]
  
  print2(paste(parameters$cluster1, parameters$cluster2, parameters$celltype))
  
  nodes <- c(parameters$cluster1, parameters$cluster2)
    
    print2(paste(parameters$cluster1, parameters$cluster2, parameters$celltype, sep = ': '))
    
    p <- plot_sankey1(merged_annotations %>% filter(.[[parameters$cluster1]] == parameters$celltype),
                      nodes = nodes,
                      names = str_to_sentence(nodes), 
                      drop_na = TRUE,
                      prefix = 'Cells: ')
    
    ggsave(plot = p,
           filename = paste0(str_remove(parameters$celltype, '/'), '.png'),
           path = here(output_dir, parameters$cluster1, paste0('sankey1_', parameters$cluster1, '_vs_', parameters$cluster2)),
           dpi = dpi,
           h = 4,
           w = 6)  
    
})
toc()
```


### Two-way
```{r}
sankey2_loop <- merged_annotations %>% 
  select(all_of(clusters)) %>% 
  unique() %>% 
  pivot_longer(cols = everything(),
               names_to = 'cluster',
               values_to = 'celltype') %>% 
  unique() %>% 
  drop_na() %>% 
  arrange(cluster, celltype) %>% 
  mutate(across(everything(), as.character))

sankey2_loop
```

```{r}
tic()
mclapply(1:nrow(sankey2_loop), mc.cores = cores, FUN = function(i) {

  parameters <- sankey2_loop[i,]
  
  print2(paste(parameters$cluster, parameters$celltype, sep = ': '))
  
  nodes <- c(annotations[1], parameters$cluster, annotations[2])
  
  p <- plot_sankey2(merged_annotations %>% filter(.[[parameters$cluster]] == parameters$celltype),
                    nodes = nodes,
                    names = str_to_sentence(nodes), 
                    drop_na = FALSE,
                    prefix = 'Cells: ')
  
  ggsave(plot = p,
         filename = paste0(str_remove(parameters$celltype, '/'), '.png'),
         path = here(output_dir, parameters$cluster, paste0('sankey2_', parameters$cluster, '_vs_', annotations[1], '_vs_', annotations[2])),
         dpi = dpi,
         h = 4,
         w = 6)
})
toc()
```


#### Annotation centric
```{r}
annotation_loop <- merged_annotations %>% 
  select(all_of(annotations)) %>% 
  unique() %>% 
  pivot_longer(cols = everything(),
               names_to = 'cluster',
               values_to = 'celltype') %>% 
  unique() %>% 
  drop_na() %>% 
  arrange(cluster, celltype) %>% 
  mutate(across(everything(), as.character))

annotation_loop
```

```{r}
mclapply(1:nrow(annotation_loop), mc.cores = cores, FUN = function(i) {

  parameters <- annotation_loop[i,]
  
  print2(paste(parameters$cluster, parameters$celltype, sep = ': '))
  
  # vs clusters
  nodes <- c(clusters[1], parameters$cluster, clusters[2])
  p <- plot_sankey2(merged_annotations %>% filter(.[[parameters$cluster]] == parameters$celltype),
                    nodes = nodes,
                    names = str_to_sentence(nodes), 
                    drop_na = FALSE, 
                    prefix = 'Cells: ')
  
  ggsave(plot = p,
         filename = paste0(str_remove(parameters$celltype, '/'), '.png'),
         path = here(output_dir, parameters$cluster, 'sankey2_vs_clusters'),
         dpi = dpi,
         h = 4,
         w = 6)
  
  # vs other annotation
  other_annotation <- setdiff(annotations, parameters$cluster)
  nodes <- c(parameters$cluster, other_annotation)
  
  p <- plot_sankey1(merged_annotations %>% filter(.[[parameters$cluster]] == parameters$celltype),
                    nodes = nodes,
                    names = str_to_sentence(nodes), 
                    drop_na = FALSE, 
                    top_n = 4,
                    prefix = 'Cells: ')
  
  ggsave(plot = p,
         filename = paste0(str_remove(parameters$celltype, '/'), '.png'),
         path = here(output_dir, parameters$cluster, paste0('sankey1_vs_', other_annotation)), 
         dpi = dpi,
         h = 4,
         w = 6)
  
})
```



```{r}
library(pheatmap)

pheatmap_input <- annotation_odds$local$label %>% 
  select(group1,
         group2,
         metric = overlap_coef) %>% 
  pivot_wider(names_from = group1, values_from = metric, values_fill = 0)

p <- pheatmap(mat = pheatmap_input %>% column_to_rownames('group2'), 
         angle_col = 0, 
         color = c('white', colorRampPalette(RColorBrewer::brewer.pal(n = 7, name = "Blues"))(100)),
         breaks = seq(0, 1, length.out = 102), display_numbers = TRUE)
```

### Annotate
#### Factor level ordering
```{r}
rashx_clusters$supercluster %>% unique() %>% sort()
```

```{r}
supercluster_order <- c('Tcm', 
                        'Trm', 
                        'Tmm', 
                        'CTL',
                        'CTLem',
                        'CTLex',
                        'Tet',
                        'Tn',
                        'Treg',
                        'cmTreg',
                        'eTreg',
                        'ILC',
                        'NK',
                        'ILC/NK',
                        'B',
                        'Plasma',
                        'Mac',
                        'Mono',
                        'InfMono',
                        'DC',
                        'migDC',
                        'moDC',
                        'LC',
                        'Mast')

annotation_order <- rashx_clusters %>% select(cluster, supercluster) %>% 
  unique() %>% 
  mutate(supercluster = factor(supercluster, levels = supercluster_order)) %>% 
  arrange(supercluster, cluster) %>% 
  mutate(order = 1:n(),
         supercluster = as.character(supercluster))

annotation_order
```


#### Prepare addition of local and global annotations
```{r}
annotation_local <- annotation_odds$local$label %>% 
  group_by(group1) %>% 
  slice_max(f_group1) %>% 
  select(group1,
         cluster = group2,
         f_in_cluster = f_group1) %>% 
  left_join(annotation_order) %>% 
  ungroup() %>% 
  select(local = group1, 
         local_cluster = cluster, 
         local_supercluster = supercluster, 
         local_order = order,
         f_in_local_cluster = f_in_cluster)

annotation_local
```


```{r}
annotation_global <- annotation_odds$global$label %>% 
  group_by(group1) %>% 
  slice_max(f_group1) %>% 
  select(group1,
         cluster = group2,
         f_in_cluster = f_group1) %>% 
  left_join(annotation_order) %>% 
  ungroup() %>% 
  select(global = group1, 
         global_cluster = cluster, 
         global_supercluster = supercluster, 
         global_order = order,
         f_in_global_cluster = f_in_cluster)

annotation_global
```

```{r}
annotation_local_levels <- annotation_local %>% 
  select(local_order, local_cluster, local_supercluster) %>% 
  unique() %>% 
  arrange(local_order)

annotation_global_levels <- annotation_global %>% 
  select(global_order, global_cluster, global_supercluster) %>% 
  unique() %>% 
  arrange(global_order)

annotation_global_levels
```

#### Merged annotation information
```{r}
new_annotations <- seuratobj@meta.data %>% 
  rownames_to_column() %>% 
  select(rowname, local, global) %>% 
  left_join(annotation_local) %>% 
  left_join(annotation_global) %>% 
  mutate(local_cluster = factor(local_cluster, levels = annotation_local_levels$local_cluster),
         local_supercluster = factor(local_supercluster, levels = unique(annotation_local_levels$local_supercluster)),
         global_cluster = factor(global_cluster, levels = annotation_global_levels$global_cluster),
         global_supercluster = factor(global_supercluster, levels = unique(annotation_global_levels$global_supercluster))) %>% 
  column_to_rownames()

new_annotations 
```
### Add annotation
```{r}
seuratobj <- AddMetaData(seuratobj, new_annotations)
```
#### Ensure correct sample factor levels
```{r}
sample_order <- seuratobj@meta.data %>% 
  select(patient, sample, sample_order) %>% 
  unique() %>% 
  arrange(sample_order)
sample_order
```

```{r}
seuratobj$condition <- factor(seuratobj$condition, levels = c('HC', 'AD', 'PV'))
seuratobj$treatment <- factor(seuratobj$treatment, levels = c('None', 'Pre', 'Mid'))
seuratobj$group <- factor(seuratobj$group, levels = c('Responder', 'Non-responder'))
seuratobj$patient <- factor(seuratobj$patient, levels = unique(sample_order$patient))
seuratobj$sample <- factor(seuratobj$sample, levels = sample_order$sample)
```

### Use global or local identities and set as cluster/supercluster in the object
Use this code to select between global or local identities as the default for a given analysis
Here we will set global as the default
```{r}
which_identity <- 'local'
which_cluster <- paste0(which_identity, '_cluster')
which_supercluster <- paste0(which_identity, '_supercluster')

seuratobj$cluster <- seuratobj[[which_cluster]] 
seuratobj$supercluster <- seuratobj[[which_supercluster]] 

seuratobj$annotation <- seuratobj$cluster # set cluster as main annotation
Idents(seuratobj) <- 'annotation'
```
### Export
```{r}
seuratobj@meta.data %>% rownames_to_column() %>% write_rds(here(output_dir, 'metadata.rds'), compress = 'gz')

seuratobj %>% write_rds(here(data_dir, 'seuratobj_annotated.rds'))
```

### Save plots
```{r}
dpi <- 600
w <- 5
h <- 5
```

#### Local
```{r}
p <- seuratobj %>% 
  seurat_feature(features = 'local_cluster', 
                 facet_hide = TRUE, 
                 legend_position = 'none',
                 size = 0.1, 
                 alpha = 0.1, 
                 color_package = 'carto', 
                 color_palette = 'Bold', 
                 rasterize_dpi = dpi)

ggsave(plot = p,
       filename = 'umap_annotated_local_cluster.png',
       dpi = dpi,
       path = output_dir,
       w = w, 
       h = h)

p
```


```{r}
p <- seuratobj %>% 
  seurat_feature(features = 'local_supercluster', 
                 facet_hide = TRUE, 
                 legend_position = 'none', 
                 size = 0.1, alpha = 0.1, 
                 color_package = 'carto', 
                 color_palette = 'Bold', 
                 rasterize_dpi = dpi)

ggsave(plot = p,
       filename = 'umap_annotated_local_supercluster.png',
       path = output_dir,
       dpi = dpi,
       w = w, 
       h = h)

p

```
```{r}
p <- seuratobj %>% 
  seurat_feature(features = 'local', 
                 facets = 'local_cluster', 
                 legend_position = 'none',
                 size = 0.1, 
                 alpha = 0.1, 
                 label = FALSE,
                 color_package = 'carto', 
                 color_palette = 'Bold', 
                 rasterize_dpi = dpi)

ggsave(plot = p,
       filename = 'umap_annotated_local_cluster_facets.png',
       dpi = dpi,
       path = output_dir,
       w = w * 2, 
       h = h * 2)
```

#### Global
```{r}
p <- seuratobj %>% 
  seurat_feature(features = 'global_cluster', 
                 facet_hide = TRUE, 
                 legend_position = 'none',
                 size = 0.1, 
                 alpha = 0.1, 
                 color_package = 'carto', 
                 color_palette = 'Bold', 
                 rasterize_dpi = dpi)

ggsave(plot = p,
       filename = 'umap_annotated_global_cluster.png',
       dpi = dpi,
       path = output_dir,
       w = w, 
       h = h)

p
```
```{r}
p <- seuratobj %>% 
  seurat_feature(features = 'global', 
                 facets = 'global_cluster', 
                 legend_position = 'none',
                 size = 0.1, 
                 alpha = 0.1, 
                 label = FALSE,
                 color_package = 'carto', 
                 color_palette = 'Bold', 
                 rasterize_dpi = dpi)

ggsave(plot = p,
       filename = 'umap_annotated_global_cluster_facets.png',
       dpi = dpi,
       path = output_dir,
       w = w * 2, 
       h = h * 2)
```


```{r}
p <- seuratobj %>% 
  seurat_feature(features = 'global_supercluster', 
                 facet_hide = TRUE, 
                 legend_position = 'none', 
                 size = 0.1, alpha = 0.1, 
                 color_package = 'carto', 
                 color_palette = 'Bold', 
                 rasterize_dpi = dpi)

ggsave(plot = p,
       filename = 'umap_annotated_global_supercluster.png',
       path = output_dir,
       dpi = dpi,
       w = w, 
       h = h)

p

```

## Session Info
```{r}
sessionInfo()
```
