---
title: "Figure"
author: "David Wu"
---

### Use the here package for robust relative paths
```{r}
library(here)
```

### Optionally convert this Rmd to R script 
```{r eval=FALSE, message=FALSE}
current_file <- rstudioapi::getActiveDocumentContext()$path
output_file <- stringr::str_replace(current_file, '.Rmd', '.R')
knitr::purl(current_file, output = output_file)
file.edit(output_file)
```

### Directories
```{r, message=FALSE}
output_dir <- here('figures/figure1') # analysis file output directory
data_dir <- here('data/derived/tildra') # data file output directory

dir.create(output_dir, showWarnings = FALSE)
dir.create(data_dir, showWarnings = FALSE)
```

### Loading the libraries
```{r, message=FALSE}
library(Seurat) 
library(tidyverse)
library(ggExtra)
library(ggforce)
library(wutilities) # devtools::install_github('symbiologist/wutilities')
library(tictoc)
library(patchwork)
library(rcartocolor)
library(ggrastr)
library(extrafont)
theme_set(theme_dwu())

source(here('figures/colors.R')) # consistent color palettes to use across figures
```

### Import seurat
```{r}
seuratobj <- read_rds(file.path(data_dir, 'seuratobj_annotated.rds'))
seuratobj
```

### Use global or local identities and set as cluster/supercluster in the object
```{r}
which_identity <- 'local'
which_cluster <- paste0(which_identity, '_cluster')
which_supercluster <- paste0(which_identity, '_supercluster')

seuratobj$cluster <- seuratobj[[which_cluster]] # set cluster as main annotation
seuratobj$supercluster <- seuratobj[[which_supercluster]] # set cluster as main annotation
Idents(seuratobj) <- 'cluster'

output_dir <- here(output_dir, which_identity)
dir.create(output_dir)
```

### Metadata
```{r}
metadata <- seuratobj@meta.data %>% rownames_to_column() %>% as_tibble()

patient_only <- subset(seuratobj, subset = treatment %in% c('Pre', 'Mid'))
patient_only
```

#### Plotting parameters
```{r}
h <- 5
w <- 5
dpi <- 600
```

### Figure 1A
#### Diagram

### Figure 1B
#### UMAP of points only (PNG)
```{r fig.height=4, fig.width=4}
p <- seurat_feature(seuratobj, 
                    features = 'cluster', 
                    facet_hide = TRUE, 
                    legend_position = 'none', 
                    label = FALSE, 
                    color_package = 'carto', 
                    color_palette = 'Bold', 
                    rasterize_dpi = dpi) 

ggsave(plot = p,
       path = output_dir,
       height = h,
       width = w,
       dpi = dpi,
       filename = 'F1B_cluster.png') 
p
```


```{r fig.height=4, fig.width=4}
p <- seurat_feature(seuratobj, 
                    features = 'supercluster', 
                    facet_hide = TRUE, 
                    legend_position = 'none', 
                    label = FALSE, 
                    color_package = 'carto', 
                    color_palette = 'Bold', 
                    rasterize_dpi = dpi) 

ggsave(plot = p,
       path = output_dir,
       height = h,
       width = w,
       dpi = dpi,
       filename = 'F1B_supercluster.png') 

p
```
#### Labels and PDF
```{r fig.height=4, fig.width=4}
p <- seurat_feature(seuratobj, 
                    features = 'cluster', 
                    facet_hide = TRUE, 
                    legend_position = 'none', 
                    label = TRUE, 
                    color_package = 'carto', 
                    color_palette = 'Bold', 
                    label_size = 3, 
                    rasterize_dpi = dpi) 

ggsave(plot = p,
       device = cairo_pdf,
       path = output_dir,
       height = h,
       width = w,
       dpi = dpi,
       filename = 'F1B_cluster.pdf') 

p
```

```{r fig.height=4, fig.width=4}
p <- seurat_feature(seuratobj, 
                    features = 'supercluster', 
                    facet_hide = TRUE, 
                    legend_position = 'none', 
                    label = TRUE, 
                    color_package = 'carto', 
                    color_palette = 'Bold', 
                    label_size = 3, 
                    rasterize_dpi = dpi) 

ggsave(plot = p,
       device = cairo_pdf,
       path = output_dir,
       height = h,
       width = w,
       dpi = dpi,
       filename = 'F1B_supercluster.pdf') 

p
```

#### Legend
```{r fig.height=4, fig.width=6}
p <- seurat_feature(seuratobj, 
                    features = 'cluster', 
                    facet_hide = TRUE, 
                    legend_position = 'right', 
                    label = FALSE, 
                    color_package = 'carto', 
                    color_palette = 'Bold', 
                    rasterize_dpi = dpi) 

ggsave(plot = p,
       device = cairo_pdf,
       path = output_dir,
       height = h,
       width = w * 1.5,
       dpi = dpi,
       filename = 'F1B_cluster_legend.pdf') 

p
```

```{r fig.height=4, fig.width=6}
p <- seurat_feature(seuratobj, 
                    features = 'supercluster', 
                    facet_hide = TRUE, 
                    legend_position = 'right', 
                    label = FALSE, 
                    color_package = 'carto', 
                    color_palette = 'Bold', 
                    rasterize_dpi = dpi) 

ggsave(plot = p,
       device = cairo_pdf,
       path = output_dir,
       height = h,
       width = w * 1.5,
       dpi = dpi,
       filename = 'F1B_supercluster_legend.pdf') 

p
```
#### Supercluster manual colors
```{r fig.height=4, fig.width=6}
annotations <- seuratobj$supercluster %>% levels()
annotations
```


```{r fig.height=4, fig.width=6}
lymphocyte_annotations <- str_subset(annotations, 'T|NK')

lymphocyte_colors1 <- rev(RColorBrewer::brewer.pal(n = ceiling(length(lymphocyte_annotations)/2), name = 'Blues'))
lymphocyte_colors2 <- rev(RColorBrewer::brewer.pal(n = ceiling(length(lymphocyte_annotations)/2), name = 'Greens'))
lymphocyte_colors <- c(rbind(lymphocyte_colors1, lymphocyte_colors2))[1:length(lymphocyte_annotations)]

b_annotations <- str_subset(annotations, 'B|Plasma') 
b_colors <- RColorBrewer::brewer.pal(n = length(b_annotations), name = 'Purples')[-1]
  
apc_annotations <- annotations %>% setdiff(c(lymphocyte_annotations, b_annotations))
apc_colors <- RColorBrewer::brewer.pal(n = length(apc_annotations), name = 'YlOrRd')

#colors_supercluster <- colorRampPalette(rev(RColorBrewer::brewer.pal(n = 7, name = "RdYlBu")))(length(annotations))
colors_supercluster <- c(lymphocyte_colors, b_colors, apc_colors)
```


```{r fig.height=4, fig.width=6}
p <- seurat_feature(seuratobj, 
                    features = 'supercluster', 
                    facet_hide = TRUE, 
                    legend_position = 'none', 
                    label = TRUE, 
                    color_package = 'custom', 
                    color_palette = colors_supercluster, 
                    rasterize_dpi = dpi) 

ggsave(plot = p,
       device = cairo_pdf,
       path = output_dir,
       height = h,
       width = w,
       dpi = dpi,
       filename = 'F1B_supercluster_custom.pdf') 

p
```

```{r fig.height=4, fig.width=6}
p <- seurat_feature(seuratobj, 
                    features = 'supercluster', 
                    facet_hide = TRUE, 
                    legend_position = 'right', 
                    label = FALSE, 
                    color_package = 'custom', 
                    color_palette = colors_supercluster, 
                    rasterize_dpi = dpi) 

ggsave(plot = p,
       device = cairo_pdf,
       path = output_dir,
       height = h,
       width = w * 1.5,
       dpi = dpi,
       filename = 'F1B_supercluster_custom_legend.pdf') 

p
```

### Figure 1C, Marker Dot Plot
#### Reynolds et al markers

```{r}
markers_reynolds <- c('KRT14', 'KRT5', 
                      'GATA3', 'KRTDAP',
                      'CRYAB', 'TYRP1',
                      'DKK3', 'PLEKHA4',
                      'LUM', 'APOD',
                      'PHLDA2', 'CCL2',
                      'IL6', 'PDGFRA',
                      'COL1A1', 'CXCL12',
                      'CD82', 'CCL19', 'KCNE4', 'CPE', 'TAGLN', 'MYL9', 'HEY1', 'CCL14', 'PECAM1',
                      'ACKR1', 'SELE', 'SNCG', 'HES1', 'MMRN1', 'XCL1' ,'XCL2', 'KLRC1', 'CCL5', 'GZMA', 
                      'CD8A', 'CD8B', 'IL7R', 'FOXP3', 'TIGIT', 'TPSB2', 'TPSAB1', 'CD79A', 'JCHAIN', 'CD163',
                      'C1QB', 'FCGR2A', 'MS4A6A', 'IL23A', 'CLEC9A', 'CLEC10A', 'CD1C', 'CD207', 'CD1A', 
                      'CD14', 'IL1B', 'CCR7', 'IDO1')

rashx_markers <- c("CD3D",
                  "CCR7",
                  "SELL",
                  "KLF2","CD69","ITGAE","CXCR6","CD4","TIGIT","FOXP3",
                  "IL2RA","CTLA4","CD8A","CD8B","GZMB","PDCD1","LAG3","KLRB1","PRF1","KLRD1","GNLY",
                  "TNFRSF18","PRDM1","BATF","TRAT1","RORA","GATA3","PTGDR2",
                  "IL7R", "HLA-DRA","HLA-DRB1",
                  "CD83","IDO1","CD207","EPCAM","CD68","C1QB","C1QC","CD163","CLEC10A","CD1C",
                  "THBD","XCR1","SIRPA","F13A1","IGKC","JCHAIN","CD79A","MS4A1","NR4A1","NR4A2","KLF4",
                  "CEBPB","LYZ","MS4A7","SERPINA1","CD14","S100A9","IL23A","TPSB2","TPSAB1","MKI67","TOP2A",
                  "ITGA4","NCR1","IL17A","IL17F","IL23R")

main_markers <- c("CD3D",
                  "CCR7",
                  "SELL",
                  "KLF2",
                  "CD69",
                  "ITGAE",
                  "CXCR6",
                  "CD4",
                  "TIGIT",
                  "FOXP3",
                  "IL2RA",
                  "CTLA4",
                  "CD8A",
                  "CD8B",
                  "GZMB",
                  "PDCD1",
                  #"LAG3",
                  "KLRB1",
                  "PRF1",
                  "KLRD1",
                  "GNLY",
                  "TNFRSF18",
                  "PRDM1",
                  "BATF",
                  "TRAT1",
                  "RORA",
                  "GATA3",
                  "PTGDR2",
                  "IL7R", 
                  "HLA-DRA",
                  "CD83",
                  #"IDO1",
                  "CD207",
                  #"EPCAM",
                  "CD68",
                  "C1QB",
                  #"C1QC",
                  "CD163",
                  #"CLEC10A",
                  "CD1C",
                  "THBD",
                  #"XCR1",
                  "SIRPA",
                  #"F13A1",
                  #"IGKC",
                  "JCHAIN",
                  "CD79A",
                  "MS4A1",
                  "NR4A1",
                  #"NR4A2",
                  #"KLF4",
                  "CEBPB",
                  "LYZ",
                  "MS4A7",
                  #"SERPINA1",
                  "CD14",
                  #"S100A9",
                  #"TPSB2",
                  "TPSAB1",
                  #"ITGA4"
                  "IL23A")
```


#### F1C dot plot horizontal
```{r}
p <- DotPlot(seuratobj, 
             features = main_markers,
             cluster.idents = TRUE,
             cols = RColorBrewer::brewer.pal(n = 7, 'BuPu')[c(1,7)],
             dot.min = 0.1,
             dot.scale = 4, 
             scale = FALSE) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, face = 'italic'))

ggsave(plot = p,
       path = output_dir,
       height = 8,
       width = 8,
       dpi = dpi,
       filename = 'F1C.pdf') 

p
```
```{r}
p <- DotPlot(seuratobj, 
             features = main_markers, 
             cluster.idents = TRUE,
             cols = RColorBrewer::brewer.pal(n = 7, 'BuPu')[c(1,7)],
             dot.min = 0.1,
             dot.scale = 4, 
             scale = FALSE) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_text(face = 'italic')) +
  coord_flip()

ggsave(plot = p,
       path = output_dir,
       height = 8,
       width = 8,
       dpi = dpi,
       filename = 'F1C_vertical.pdf') 

p
```

### Figure 1D; UMAP facets
```{r}
unit_height <- 2
unit_width <- 2

color_palette6 <- rcartocolor::carto_pal(n = 6, name = 'Bold')
view_colors(color_palette6)
```

#### By condition
```{r}
p <- seurat_feature(subset(seuratobj, subset = condition != 'AD'), 
                    features = 'condition', 
                    facets = 'condition', 
                    facet_hide = TRUE,
                    label = FALSE, color_package = 'custom', 
                    color_palette = color_palette6[c(6, 1)], 
                    legend_position = 'none', 
                    axis_title_position = 'none', 
                    rasterize_dpi = dpi)

ggsave(plot = p,
       path = output_dir,
       height = unit_height,
       width = unit_width * 2,
       dpi = dpi,
       filename = 'F1D_condition.png') 

p
```

```{r}
p <- seurat_feature(subset(seuratobj, subset = condition != 'AD'), 
                    features = 'condition', 
                    facets = 'condition', 
                    label = FALSE, 
                    facet_hide = TRUE,
                    color_package = 'custom', 
                    color_palette = color_palette6[c(6, 1)], 
                    legend_position = 'none', 
                    axis_title_position = 'none', 
                    nrow = 2,
                    rasterize_dpi = dpi)

ggsave(plot = p,
       path = output_dir,
       height = unit_height * 1.9,
       width = unit_width,
       dpi = dpi,
       filename = 'F1D_condition_vertical.png') 

p
```

#### Treatment time-point
```{r}
p <- seurat_feature(patient_only, 
                    features = 'treatment', 
                    facets = 'treatment', 
                    facet_hide = TRUE,
                    label = FALSE, 
                    nrow = 1, 
                    color_package = 'custom', 
                    color_palette = color_palette6[c(5, 3)], 
                    legend_position = 'none', 
                    axis_title_position = 'none',
                    rasterize_dpi = dpi)

ggsave(plot = p,
       path = output_dir,
       height = unit_height,
       width = unit_width * 2,
       dpi = dpi,
       filename = 'F1D_treatment.png') 

p
```
#### By CD status
#### CD4 vs CD8 status based on ADT and RNA data
```{r}
cd_data <- FetchData(seuratobj, 
                     vars = c('CD4', 'CD8A', 'CD8B', 'ADT.CD4', 'ADT.CD8', 
                              'cluster', 'supercluster', 'celltype', 'treatment')) %>% 
  filter(celltype == 'Lymphocyte') %>% 
  rownames_to_column() %>% 
  mutate(ADT = case_when(
    ADT.CD4 == 'Positive' & ADT.CD8 == 'Positive' ~ 'Double-positive',
    ADT.CD4 == 'Positive' ~ 'CD4',
    ADT.CD8 == 'Positive' ~ 'CD8',
    TRUE ~ 'Double-negative'
  ))
```

#### CD8
```{r}
p <- cd_data %>% 
  ggplot(aes(x = CD8A,
             y = CD8B,
             color = ADT)) +
  geom_point(alpha = 0.1,
             size = 0.5) +
  facet_wrap(~ADT)

p
```
#### CD4
```{r}
p <- cd_data %>% 
  ggplot(aes(x = CD4,
             y = CD8A,
             color = ADT)) +
  geom_point(alpha = 0.1,
             size = 0.5) +
  facet_wrap(~ADT)

p
```
#### Treg vs other T cells
```{r}
p <- cd_data %>% 
  filter(celltype == 'Lymphocyte',
         str_detect(supercluster, 'T')) %>% 
  ggplot(aes(x = CD8A,
             y = CD8B,
             color = ADT)) +
  geom_point(alpha = 0.1,
             size = 0.5) +
  facet_wrap(~supercluster)

p
```
### Lymphocytes vs APCs
```{r}
p <- cd_data %>% 
  ggplot(aes(x = CD8A,
             y = CD8B,
             color = ADT)) +
  geom_point(alpha = 0.1,
             size = 0.5) +
  facet_wrap(~celltype)

p
```

#### Classify CD4/8 status based on protein then RNA
```{r}
cd_status <- cd_data %>% 
  mutate(CD = case_when(
    ADT == 'Double-positive' ~ 'Unknown',
    ADT == 'CD4' ~ 'CD4',
    ADT == 'CD8' ~ 'CD8',
    CD4 > 1 & (CD8A < 1 & CD8B < 1) ~ 'CD4',
    (CD8A > 1 | CD8B > 1) & CD4 < 1 ~ 'CD8',
    TRUE ~ 'Unknown'
  ))

cd_status %>% pull(CD) %>% table()
```
```{r}
cd_data %>% 
  ggplot(aes(x = CD4)) +
  geom_histogram() +
  facet_wrap(~celltype, scales = 'free_y')
```
```{r}
cd_data %>% 
  ggplot(aes(x = CD8A)) +
  geom_histogram() +
  facet_wrap(~celltype, scales = 'free_y')
```


### Assign each cluster to closest match
```{r}
cluster_cd_tally <- cd_status %>% 
  group_by(cluster, CD) %>% 
  tally() %>% 
  pivot_wider(names_from = CD,
              values_from = n) %>% 
  rowwise() %>% 
  mutate(CD = ifelse(CD4 > CD8, 'CD4', 'CD8'))

cluster_cd_tally 
```

```{r}
cd_status <- cd_status #%>% select(-CD) %>% left_join(cluster_cd_tally %>% select(cluster, CD))
```

#### Add metadata
```{r}
add_cd <- seuratobj@meta.data %>% 
  rownames_to_column() %>% 
  select(rowname) %>% 
  left_join(cd_status)
seuratobj$CD <- add_cd$CD

lymphocytes <- subset(seuratobj, str_detect(seuratobj$cluster, 'T') & CD != 'Unknown')
```

```{r}
p <- seurat_feature(lymphocytes, 
                    features = 'CD', 
                    facets = 'CD', 
                    facet_hide = TRUE,
                    label = FALSE, 
                    nrow = 1, 
                    color_package = 'custom', 
                    color_palette = c(color_palette6[c(2, 4)], 'transparent'), 
                    legend_position = 'none', 
                    axis_title_position = 'none',
                    rasterize_dpi = dpi)

ggsave(plot = p,
       path = output_dir,
       height = unit_height,
       width = unit_width * 2,
       dpi = dpi,
       filename = 'F1D_cd.png') 

p
```
```{r}
p <- seurat_feature(lymphocytes, 
                    features = 'CD', 
                    facets = 'CD', 
                    facet_hide = TRUE,
                    label = FALSE, 
                    nrow = 2, 
                    color_package = 'custom', 
                    color_palette = c(color_palette6[c(2, 4)], 'transparent'), 
                    legend_position = 'none', 
                    axis_title_position = 'none',
                    rasterize_dpi = dpi)

ggsave(plot = p,
       path = output_dir,
       height = unit_height * 2,
       width = unit_width,
       dpi = dpi,
       filename = 'F1D_cd_vertical.png') 

p
```

#### By patient, mixed colors
```{r}
patient_colors <- carto_pal(n = 9, name = 'Prism')
view_colors(patient_colors)
```


```{r}
p1 <- seurat_feature(patient_only, 
                     features = 'patient', 
                     label = FALSE, 
                     legend_position = 'none', 
                     facet_hide = TRUE, 
                     axis_title_position = 'none', 
                     color_package = 'custom',
                     color_palette = patient_colors,
                     rasterize_dpi = dpi)

ggsave(plot = p1,
       path = output_dir,
       height = unit_height,
       width = unit_width,
       dpi = dpi,
       filename = 'F1D_patient_umap.png') 

p1
```

```{r}
p1 <- seurat_feature(patient_only, 
                     features = 'patient', 
                     label = FALSE, 
                     facet_hide = TRUE, 
                     legend_position = 'right',  
                     axis_title_position = 'none', 
                     color_package = 'custom',
                     color_palette = patient_colors,
                     rasterize_dpi = dpi)

ggsave(plot = p1,
       device = cairo_pdf,
       path = output_dir,
       height = unit_height,
       width = unit_width,
       dpi = dpi,
       filename = 'F1D_patient_umap_legend.pdf') 

p1
```

#### Patients faceted
```{r}
p2 <- seurat_feature(patient_only, 
                     features = 'patient', 
                     facets = 'patient', 
                     label = FALSE, facet_hide = TRUE, alpha = 0.5, size = 0.2, 
                     nrow = 3,
                     color_package = 'custom',
                     color_palette = patient_colors,
                     rasterize_dpi = dpi)

ggsave(plot = p2,
       device = cairo_pdf,
       path = output_dir,
       height = unit_height * 3,
       width = unit_width * 3.1,
       dpi = dpi,
       filename = 'F1D_patient_matrix.pdf') 

p2
```

```{r}
p <- seurat_feature(patient_only, 
                    features = 'patient', 
                    facets = 'patient', 
                    label = FALSE, 
                    facet_hide = FALSE,
                    legend_position = 'none',
                    alpha = 0.5, 
                    size = 0.2, 
                    nrow = 1,
                    color_package = 'custom',
                    color_palette = patient_colors,
                    rasterize_dpi = dpi)

ggsave(plot = p,
       path = output_dir,
       height = unit_height,
       width = unit_width * 9,
       dpi = dpi,
       filename = 'F1D_patient_linear.png') 

p
```


### Figure 1E
#### Plotting order
```{r}
# patient_order <- metadata %>% select(patient, sample, condition, treatment, response, group, ends_with('label')) %>% unique()
# 
# 
# tildra_order <- patient_order %>% 
#   #filter(treatment != 'None') %>% 
#   arrange(patient, treatment)
# 
# tildra_order
```

#### How many patients are represented in each cell cluster?
```{r}
metadata %>% filter(treatment %in% c('Pre', 'Mid')) %>% select(cluster, patient) %>% unique() %>% group_by(cluster) %>% tally() 
```
```{r}
metadata %>% filter(treatment %in% c('Pre', 'Mid')) %>% select(cluster, patient, treatment) %>% unique() %>% group_by(cluster, treatment) %>% tally() %>% 
  pivot_wider(names_from = treatment, values_from = n)
```

