---
title: "Figrue 4"
author: "David Wu"
---

### Use the here package for robust relative paths
```{r}
library(here)
```

### Optionally convert this Rmd to R script 
```{r eval=FALSE, message=FALSE}
current_file <- rstudioapi::getActiveDocumentContext()$path
output_file <- stringr::str_replace(current_file, '.Rmd', '.R')
knitr::purl(current_file, output = output_file)
file.edit(output_file)
```

### Directories
```{r, message=FALSE}
figures_dir <- here('figures/figure4') # analysis file output directory
data_dir <- here('data/derived/tildra') # data file output directory

dir.create(figures_dir, showWarnings = FALSE)
dir.create(data_dir, showWarnings = FALSE)
```

### Loading the libraries
```{r, message=FALSE}
library(Seurat) 
library(tidyverse)
library(ggExtra)
library(patchwork)
library(ggthemes)
library(ggforce)
library(wutilities) # devtools::install_github('symbiologist/wutilities')
library(tictoc)
library(rcartocolor)
library(ggrepel)
library(extrafont)
loadfonts()
theme_set(theme_dwu()) # set default theme
source(here('figures/colors.R')) # consistent color palettes

```

### Functions
```{r}
library(openxlsx)
write_excel_notebook <- function(input_table, 
                                 output_file, 
                                 tabs = 'cluster', 
                                 prefix = 'cluster_') {
  OUT <- createWorkbook()
  
  input_table$tabs <- input_table[[tabs]]
  tabs <- input_table[[tabs]] %>% unique()
  
  
  for(i in tabs){
    tryCatch({
      table_subset <- input_table %>% filter(tabs == i) %>% select(-tabs)
      
      sheet_name <- paste0(prefix, i) %>% str_remove('[\\?/]') %>% str_remove('[\\/]') %>% str_remove('[\\:/]') 
      addWorksheet(OUT, sheet_name)
      writeData(OUT, sheet = sheet_name, x = table_subset)
      
    }, error=function(e){message(paste0("We have an error here in ", i))})
  }
  
  saveWorkbook(OUT, output_file, overwrite = TRUE)
}
```

### Import
```{r}
seuratobj <- read_rds(here(data_dir, 'seuratobj_annotated.rds'))
metadata <- seuratobj@meta.data %>% rownames_to_column()
seuratobj
```
### Use global or local identities and set as cluster/supercluster in the object
```{r}
which_identity <- 'local'
which_cluster <- paste0(which_identity, '_cluster')
which_supercluster <- paste0(which_identity, '_supercluster')
which_ultracluster <- paste0(which_identity, '_ultracluster')
which_annotation <- which_supercluster

seuratobj$cluster <- seuratobj[[which_cluster]] 
seuratobj$supercluster <- seuratobj[[which_supercluster]] 
seuratobj$ultracluster <- seuratobj[[which_ultracluster]] 
Idents(seuratobj) <- 'cluster' # set cluster as main annotation

output_dir <- here(figures_dir, which_identity)
dir.create(output_dir)
```
### Patient metadata
```{r}
metadata_patient <- metadata %>% 
  select(id, sample, patient, treatment, group) %>% 
  filter(treatment != 'None') %>% 
  mutate(category = case_when(
    patient == 'P7' ~ 'Non-responder (IL-23i-sensitive)',
    group == 'Responder' ~ 'Responder',
    group == 'Non-responder' ~ 'Non-responder (IL-23i-insensitive)')) %>% 
  mutate(category = factor(category, 
                           levels = c('Responder', 'Non-responder (IL-23i-sensitive)', 'Non-responder (IL-23i-insensitive)'))) %>% 
  arrange(sample) %>% 
  unique() %>% 
  droplevels()

metadata_patient
```
### DE genes per patient
```{r}
de_patient_vs_hc <- read_tsv(here('analysis/output/08_de', which_annotation, 'de_patient_vs_hc.tsv.gz'))
```



### Analysis


### Cluster size and % 
```{r}
analyze <- metadata %>% 
  select(rowname, id, patient, sample, condition, treatment, cluster, supercluster, ultracluster, group, celltype) %>% 
  left_join(metadata_patient %>% select(id, category))

analyze %>% head()
```
```{r}
cluster_tallies <- analyze %>% 
  group_by(sample, patient, condition, cluster, celltype, treatment, group, category) %>% 
  tally() %>% 
  group_by(sample) %>% 
  mutate(total = sum(n),
         f = n/total,
         pct = round(100 * f, 2))

supercluster_tallies <- analyze %>% 
  group_by(sample, patient, condition, supercluster, celltype, treatment, group, category) %>% 
  tally() %>% 
  group_by(sample) %>% 
  mutate(total = sum(n),
         f = n/total,
         pct = round(100 * f, 2))

ultracluster_tallies <- analyze %>% 
  group_by(sample, patient, condition, ultracluster, celltype, treatment, group, category) %>% 
  tally() %>% 
  group_by(sample) %>% 
  mutate(total = sum(n),
         f = n/total,
         pct = round(100 * f, 2))

cluster_tallies
```
#### DE tally
```{r}
de_sample_tally <- de_patient_vs_hc %>% 
  filter(p_val_adj < 0.01) %>% 
  group_by(sample, cluster) %>% 
  tally()

de_sample_tally
```


#### Major celltypes

```{r}
four_colors <- carto_pal(n = 5, name = 'Bold')
  
conditions <- c('APC', 'Lymphocyte')

```

```{r}
celltype_input <- cluster_tallies %>% 
    filter(treatment %in% c('Pre', 'Mid')) %>% 
    group_by(patient, sample, treatment, total, celltype, group) %>% 
    summarize(n = sum(n)) %>% 
  rowwise() %>% 
  mutate(f = n/total, 
         pct = 100 * f)
  
```

#### Combined
```{r}
treatment_groups <- celltype_input %>% 
  ungroup() %>% 
  select(treatment, group) %>% 
  unique() %>% 
  arrange(group, treatment) %>% 
  mutate(treatment_group = paste(treatment, group, sep = ':'),
         treatment_group = factor(treatment_group, levels = treatment_group))
treatment_groups
```


```{r}
p <- celltype_input %>% 
  filter(celltype %in% c('APC', 'Lymphocyte')) %>% 
  bind_rows(cluster_tallies %>% 
              filter(treatment %in% c('Pre', 'Mid')) %>% 
              filter(str_detect(cluster, 'Trm')) %>% 
              mutate(celltype = cluster)) %>% 
  bind_rows(supercluster_tallies %>% 
              filter(treatment %in% c('Pre', 'Mid')) %>% 
              filter(str_detect(supercluster, 'Trm')) %>% 
              mutate(celltype = 'Trm all')) %>%
  select(sample, treatment, group, celltype, n, pct) %>% 
  pivot_longer(cols = c(n, pct),
               names_to = 'metric',
               values_to = 'value') %>% 
  left_join(treatment_groups) %>% 
ggplot(aes(x = treatment_group,
             y = value,
             fill = group)) +
  geom_boxplot() +
  labs(x = '',
       y = '# of Cells') +
  facet_wrap(~metric + celltype, scales = 'free_y', nrow = 2) +
  scale_fill_manual(values = colors_response2) +
  theme_dwu(legend.position = 'top') +
  scale_x_discrete(labels = c('Pre', 'Mid', 'Pre', 'Mid'))

ggsave(plot = p,
       device = cairo_pdf,
       filename = 'F2E_box_combined.pdf',
       path = output_dir,
       h = 5,
       w = 10)

p
```
#### All clusters, n
```{r}
p <- cluster_tallies %>% 
  filter(treatment %in% c('Pre', 'Mid')) %>% 
  select(sample, treatment, group, cluster, n, pct) %>% 
  pivot_longer(cols = c(n, pct),
               names_to = 'metric',
               values_to = 'value') %>% 
  left_join(treatment_groups) %>% 
  filter(metric == 'n') %>% 
ggplot(aes(x = treatment_group,
             y = value,
             fill = group)) +
  geom_boxplot() +
  labs(x = '',
       y = '# of Cells') +
  facet_wrap(~cluster, scales = 'free_y') +
  scale_fill_manual(values = colors_response2) +
  theme_dwu(legend.position = 'top') +
  scale_x_discrete(labels = c('Pre', 'Mid', 'Pre', 'Mid'))

ggsave(plot = p,
       device = cairo_pdf,
       filename = 'F2E_box_all_clusters_n.pdf',
       path = output_dir,
       h = 8,
       w = 8)

p
```
#### All clusters, f
```{r}
p <- cluster_tallies %>% 
  filter(treatment %in% c('Pre', 'Mid')) %>% 
  select(sample, treatment, group, cluster, n, pct) %>% 
  pivot_longer(cols = c(n, pct),
               names_to = 'metric',
               values_to = 'value') %>% 
  left_join(treatment_groups) %>% 
  filter(metric == 'pct') %>% 
ggplot(aes(x = treatment_group,
             y = value,
             fill = group)) +
  geom_boxplot() +
  labs(x = '',
       y = '% of Cells') +
  facet_wrap(~cluster, scales = 'free_y') +
  scale_fill_manual(values = colors_response2) +
  theme_dwu(legend.position = 'top') +
  scale_x_discrete(labels = c('Pre', 'Mid', 'Pre', 'Mid'))

ggsave(plot = p,
       device = cairo_pdf,
       filename = 'F2E_box_all_clusters_f.pdf',
       path = output_dir,
       h = 8,
       w = 8)

p
```

#### All superclusters, n
```{r}
p <- supercluster_tallies %>% 
  filter(treatment %in% c('Pre', 'Mid')) %>% 
  select(sample, treatment, group, supercluster, n, pct) %>% 
  pivot_longer(cols = c(n, pct),
               names_to = 'metric',
               values_to = 'value') %>% 
  left_join(treatment_groups) %>% 
  filter(metric == 'n') %>% 
ggplot(aes(x = treatment_group,
             y = value,
             fill = group)) +
  geom_boxplot() +
  labs(x = '',
       y = '# of Cells') +
  facet_wrap(~supercluster, scales = 'free_y') +
  scale_fill_manual(values = colors_response2) +
  theme_dwu(legend.position = 'top') +
  scale_x_discrete(labels = c('Pre', 'Mid', 'Pre', 'Mid'))

ggsave(plot = p,
       device = cairo_pdf,
       filename = 'F2E_box_all_superclusters_n.pdf',
       path = output_dir,
       h = 8,
       w = 8)

p
```
#### All clusters, f
```{r}
p <- supercluster_tallies %>% 
  filter(treatment %in% c('Pre', 'Mid')) %>% 
  select(sample, treatment, group, supercluster, n, pct) %>% 
  pivot_longer(cols = c(n, pct),
               names_to = 'metric',
               values_to = 'value') %>% 
  left_join(treatment_groups) %>% 
  filter(metric == 'pct') %>% 
ggplot(aes(x = treatment_group,
             y = value,
             fill = group)) +
  geom_boxplot() +
  labs(x = '',
       y = '% of Cells') +
  facet_wrap(~supercluster, scales = 'free_y') +
  scale_fill_manual(values = colors_response2) +
  theme_dwu(legend.position = 'top') +
  scale_x_discrete(labels = c('Pre', 'Mid', 'Pre', 'Mid'))

ggsave(plot = p,
       device = cairo_pdf,
       filename = 'F2E_box_all_superclusters_f.pdf',
       path = output_dir,
       h = 8,
       w = 8)

p
```


### Figure 2F
#### n
```{r}
p <- cluster_tallies %>% 
  filter(str_detect(cluster, 'Trm')) %>% 
  filter(!is.na(group)) %>% 
  select(sample, treatment, group, celltype, n, pct) %>% 
  pivot_longer(cols = c(n, pct),
               names_to = 'metric',
               values_to = 'value') %>% 
  ggplot(aes(x = sample,
             y = value,
             fill = group)) +
  geom_bar(stat = 'identity', color = 'black', alpha = 0.8) +
  labs(x = '',
       y = '') +
  theme_dwu(legend.position = 'top') +
  coord_flip() +
  facet_wrap(~metric, nrow = 1, scales = 'free_x') +
  scale_fill_manual(values = colors_response2) +
  scale_x_discrete(limits = rev)

ggsave(plot = p,
       device = cairo_pdf,
       filename = 'F2F_bar_patient_trm_combined.pdf',
       path = output_dir,
       h = 4,
       w = 4)

p
```
```{r}
#remotes::install_github("coolbutuseless/ggpattern")
library(ggpattern)
patient_labels <- metadata_patient %>% 
  filter(!is.na(group)) %>% mutate(label = ifelse(treatment == 'Pre', as.character(patient), '')) %>% 
  pull(label)

p <- cluster_tallies %>% 
  filter(str_detect(cluster, 'Trm1')) %>% 
  filter(!is.na(group)) %>% 
  select(sample, treatment, group, celltype, n, pct) %>% 
  pivot_longer(cols = c(n, pct),
               names_to = 'metric',
               values_to = 'value') %>% 
  ggplot(aes(x = value,
             y = sample,
             fill = group,
             pattern = treatment)) +
  geom_bar_pattern(stat = 'identity', 
                   color = 'black', 
                   alpha = 0.8,
                   pattern_density = 0.1,
                   pattern_angle = 45,
                   pattern_key_scale_factor = 0.6) +
  labs(x = '',
       y = '') +
  theme(legend.position = 'top',
        axis.ticks.y = element_blank(),
        axis.text.y = element_text(vjust = 1)) +
  facet_wrap(~metric, nrow = 1, scales = 'free_x') +
  scale_fill_manual(values = colors_response2) +
  scale_pattern_manual(values = c(Pre = 'stripe', Mid = 'none')) +
  scale_y_discrete(limits = rev,
                   labels = rev(patient_labels)) +
  guides(pattern = guide_legend(override.aes = list(fill = 'white')),
         fill = guide_legend(override.aes = list(pattern = 'none')))

p
```

### Loop
```{r}
dir.create(here(output_dir, 'patient_bar_cluster'))

p_cluster <- map(levels(cluster_tallies$cluster), function(i) {
  
  
  p <- cluster_tallies %>% 
    mutate(sample = factor(sample, levels = str_subset(levels(sample), '-'))) %>% 
    filter(cluster == i) %>% 
    filter(!is.na(group)) %>% 
    select(sample, treatment, group, celltype, n, pct) %>% 
    pivot_longer(cols = c(n, pct),
                 names_to = 'metric',
                 values_to = 'value') %>% 
   ggplot(aes(x = value,
             y = sample,
             fill = group,
             pattern = treatment)) +
  geom_bar_pattern(stat = 'identity', 
                   color = 'black', 
                   alpha = 0.8,
                   pattern_density = 0.1,
                   pattern_angle = 45,
                   pattern_key_scale_factor = 0.6) +
  labs(x = '',
       y = '',
       title = i) +
  theme(legend.position = 'top',
        axis.ticks.y = element_blank(),
        axis.text.y = element_text(vjust = 1)) +
  facet_wrap(~metric, nrow = 1, scales = 'free_x') +
  scale_fill_manual(values = colors_response2) +
  scale_pattern_manual(values = c(Pre = 'stripe', Mid = 'none')) +
  scale_y_discrete(limits = rev,
                   labels = rev(patient_labels)) +
  guides(pattern = guide_legend(override.aes = list(fill = 'white')),
         fill = guide_legend(override.aes = list(pattern = 'none')))
  
  
  ggsave(plot = p,
         device = cairo_pdf,
         filename = paste0('F2F_bar_patient_', str_replace(i, '/','+'),'.pdf'),
         path = here(output_dir, 'patient_bar_cluster'),
         h = 4,
         w = 4)
  
  p
  
}) %>% wrap_plots()

p_cluster
```

```{r eval=FALSE}
dir.create(here(output_dir, 'patient_bar_supercluster_de'))

### with DE genes
p_supercluster <- map(levels(supercluster_tallies$supercluster), function(i) {
  
  print2(i)
  
  p <- supercluster_tallies %>% 
    filter(supercluster == i) %>% 
    filter(!is.na(group)) %>% 
    left_join(de_sample_tally %>% select(sample, supercluster = cluster, DEGs = n)) %>% 
    select(sample, treatment, group, Abundance = n, Proportion = pct, DEGs) %>% 
    pivot_longer(cols = c(Abundance, Proportion, DEGs),
                 names_to = 'metric',
                 values_to = 'value') %>% 
    mutate(metric = factor(metric, levels = c('Abundance', 'Proportion', 'DEGs')),
           sample = factor(sample, levels = metadata_patient$sample)) %>% 
    ggplot(aes(x = sample,
               y = value,
               fill = group)) +
    geom_bar(stat = 'identity', color = 'black', alpha = 0.8) +
    labs(x = '',
         y = '') +
    theme_dwu(legend.position = 'top') +
    coord_flip() +
    facet_wrap(~metric, nrow = 1, scales = 'free_x') +
    scale_fill_manual(values = colors_response2) +
    scale_x_discrete(limits = rev, drop = FALSE) +
    ggtitle(i)
  
  
  ggsave(plot = p,
         device = cairo_pdf,
         filename = paste0(str_replace(i, '/','+'),'.pdf'),
         path = here(output_dir, 'patient_bar_supercluster_de'),
         h = 4,
         w = 6)
  
  p
  
}) %>% wrap_plots()
```


```{r}
### without DE genes
dir.create(here(output_dir, 'patient_bar_supercluster'))


patient_labels_minus_p6 <- metadata_patient %>% 
  filter(!is.na(group)) %>% 
  filter(patient != 'P6') %>% 
  mutate(label = ifelse(treatment == 'Pre', as.character(patient), '')) %>% 
  pull(label)

p <- map(levels(supercluster_tallies$supercluster), function(i) {
  
  print2(i)
  
  p <- supercluster_tallies %>% 
    filter(supercluster == i) %>% 
    filter(!is.na(group)) %>% 
    filter(patient != 'P6') %>% # exclude due to Mast cell issue
    select(sample, treatment, category, Abundance = n, Proportion = pct) %>% 
    pivot_longer(cols = c(Abundance, Proportion),
                 names_to = 'metric',
                 values_to = 'value') %>% 
    mutate(metric = factor(metric, levels = c('Abundance', 'Proportion')),
           sample = factor(sample, levels = (metadata_patient %>% filter(patient != 'P6') %>% pull(sample)))) %>% 
    drop_na() %>% 
    ggplot(aes(x = value,
             y = sample,
             fill = category,
             pattern = treatment)) +
  geom_bar_pattern(stat = 'identity', 
                   color = 'black', 
                   alpha = 0.8,
                   pattern_density = 0.1,
                   pattern_angle = 45,
                   pattern_key_scale_factor = 0.6) +
  labs(x = '',
       y = '',
       title = i) +
  theme(legend.position = 'top',
        axis.ticks.y = element_blank(),
        axis.text.y = element_text(vjust = 1)) +
  facet_wrap(~metric, nrow = 1, scales = 'free_x') +
  scale_fill_manual(values = colors_response3) +
  scale_pattern_manual(values = c(Pre = 'stripe', Mid = 'none')) +
  scale_y_discrete(limits = rev,
                   drop = FALSE,
                   labels = rev(patient_labels_minus_p6)) +
  guides(pattern = guide_legend(override.aes = list(fill = 'white')),
         fill = guide_legend(override.aes = list(pattern = 'none')))
  
  
  ggsave(plot = p,
         device = cairo_pdf,
         filename = paste0(str_replace(i, '/','+'),'.pdf'),
         path = here(output_dir, 'patient_bar_supercluster'),
         h = 4,
         w = 4)
  
  p
  
}) 


p_subset <- p[[2]] + p[[3]] + p[[8]] + p[[13]] + p[[18]] + p[[19]] 
  
ggsave(plot = p_subset,
         device = cairo_pdf,
         filename = 'F4A_patient_bar_supercluster_combined.pdf',
         path = here(output_dir),
         h = 7,
         w = 12)



```
### Truncated
```{r}
p <- map(levels(supercluster_tallies$supercluster), function(i) {
  
  print2(i)
  
  plot_input <- supercluster_tallies %>% 
    filter(supercluster == i) %>% 
    filter(!is.na(group)) %>% 
    filter(patient != 'P6') %>% # exclude due to Mast cell issue
    select(sample, treatment, category, Abundance = n, Proportion = pct) %>% 
    pivot_longer(cols = c(Abundance, Proportion),
                 names_to = 'metric',
                 values_to = 'value') %>% 
    mutate(metric = factor(metric, levels = c('Abundance', 'Proportion')),
           sample = factor(sample, levels = (metadata_patient %>% filter(patient != 'P6') %>% pull(sample)))) %>% 
    drop_na() 
  
  cutoff <- plot_input %>% 
    group_by(metric) %>% 
    summarize(cutoff = quantile(value, probs = 0.9))
  
  p <- plot_input %>% 
    left_join(cutoff) %>% 
    mutate(value = ifelse(value >= cutoff, 0, value)) %>% 
    ggplot(aes(x = value,
               y = sample,
               fill = category,
               pattern = treatment)) +
    geom_bar_pattern(stat = 'identity', 
                     color = 'black', 
                     alpha = 0.8,
                     pattern_density = 0.1,
                     pattern_angle = 45,
                     pattern_key_scale_factor = 0.6) +
    labs(x = '',
         y = '',
         title = i) +
    theme(legend.position = 'top',
          axis.ticks.y = element_blank(),
          axis.text.y = element_text(vjust = 1)) +
    facet_wrap(~metric, nrow = 1, scales = 'free_x') +
    scale_fill_manual(values = colors_response3) +
    scale_pattern_manual(values = c(Pre = 'stripe', Mid = 'none')) +
    scale_y_discrete(limits = rev,
                     drop = FALSE,
                     labels = rev(patient_labels_minus_p6)) +
    guides(pattern = guide_legend(override.aes = list(fill = 'white')),
           fill = guide_legend(override.aes = list(pattern = 'none')))
  
  
  p
  
}) 


p_subset <- p[[2]] + p[[3]] + p[[8]] + p[[13]] + p[[18]] + p[[19]] 
  
ggsave(plot = p_subset,
         device = cairo_pdf,
         filename = 'F4A_patient_bar_supercluster_combined_truncated.pdf',
         path = here(output_dir),
         h = 7,
         w = 12)

```

```{r}
dir.create(here(output_dir, 'patient_bar_ultracluster'))

p_ultracluster <- map(levels(ultracluster_tallies$ultracluster), function(i) {
  
  print2(i)
  
  p <- ultracluster_tallies %>% 
    mutate(sample = factor(sample, levels = str_subset(levels(sample), '-'))) %>% 
    filter(ultracluster == i) %>% 
    filter(!is.na(group)) %>% 
    select(sample, treatment, group, celltype, n, pct) %>% 
    pivot_longer(cols = c(n, pct),
                 names_to = 'metric',
                 values_to = 'value') %>% 
    ggplot(aes(x = value,
             y = sample,
             fill = group,
             pattern = treatment)) +
  geom_bar_pattern(stat = 'identity', 
                   color = 'black', 
                   alpha = 0.8,
                   pattern_density = 0.1,
                   pattern_angle = 45,
                   pattern_key_scale_factor = 0.6) +
  labs(x = '',
       y = '',
       title = i) +
  theme(legend.position = 'top',
        axis.ticks.y = element_blank(),
        axis.text.y = element_text(vjust = 1)) +
  facet_wrap(~metric, nrow = 1, scales = 'free_x') +
  scale_fill_manual(values = colors_response2) +
  scale_pattern_manual(values = c(Pre = 'stripe', Mid = 'none')) +
  scale_y_discrete(limits = rev,
                   labels = rev(patient_labels)) +
  guides(pattern = guide_legend(override.aes = list(fill = 'white')),
         fill = guide_legend(override.aes = list(pattern = 'none')))
  
  
  ggsave(plot = p,
         device = cairo_pdf,
         filename = paste0('F2F_bar_patient_', str_replace(i, '/','+'),'.pdf'),
         path = here(output_dir, 'patient_bar_ultracluster'),
         h = 4,
         w = 4)
  
  p
  
}) %>% wrap_plots()
```
### Numbers


### IL17 secreting cells
```{r}
il17 <- pseudobulk(seuratobj,
                   genes = c('IL17A', 'IL17F', 'IL26', 'IFNG'))

il17 %>% arrange(-cpm)
```
```{r}
il17 %>% 
  ggplot(aes(y = reorder(ident, cpm),
             x = cpm )) +
  geom_col() +
  facet_wrap(~gene) + 
  labs(x = 'Counts per Million',
       y = '') 
```

```{r}
il17 <- FetchData(seuratobj, vars = c('IL17A', 'IL17F', 'IL26', 'IFNG', 'id', 'supercluster')) %>% 
  rownames_to_column() %>% left_join(metadata_patient)

il17
```
```{r}
p <- il17 %>% 
  filter(treatment != 'Mid',
         supercluster == 'Trm') %>% 
  ggplot(aes(x = IL17A,
             y = IL17F)) +
  geom_point(alpha = 0.5) 

ggExtra::ggMarginal(p)
```

```{r}
p <- il17 %>% 
  filter(treatment != 'Mid',
         supercluster == 'Trm') %>% 
  ggplot(aes(x = IL26,
             y = IFNG)) +
  geom_point(alpha = 0.5) 

ggExtra::ggMarginal(p)
```


```{r}
il17 %>% 
  filter(treatment != 'None',
         supercluster == 'Trm',
         IL17A + IL17F + IL26 + IFNG > 0) 
```

```{r}
#dir.create(here(output_dir, 'patient_bar_t17'))

totals <- seuratobj@meta.data %>% group_by(sample) %>% tally(name = 'total')
p <- il17 %>% 
  filter(treatment != 'None',
         supercluster == 'Trm',
         IL17A + IL17F + IL26 + IFNG > 0)  %>% 
  left_join(totals) %>% 
  mutate(sample = factor(sample, levels = str_subset(levels(sample), '-'))) %>% 
  group_by(sample, treatment, group, total) %>% 
  tally() %>% 
  mutate(pct = 100 * n/total) %>% 
  select(sample, treatment, group, Abundance = n, Proportion = pct) %>% 
    pivot_longer(cols = c(Abundance, Proportion),
                 names_to = 'metric',
                 values_to = 'value') %>% 
    mutate(metric = factor(metric, levels = c('Abundance', 'Proportion')),
           sample = factor(sample, levels = metadata_patient$sample)) %>% 
ggplot(aes(x = value,
             y = sample,
             fill = group,
             pattern = treatment)) +
  geom_bar_pattern(stat = 'identity', 
                   color = 'black', 
                   alpha = 0.8,
                   pattern_density = 0.1,
                   pattern_angle = 45,
                   pattern_key_scale_factor = 0.6) +
  labs(x = '',
       y = '') +
  theme(legend.position = 'top',
        axis.ticks.y = element_blank(),
        axis.text.y = element_text(vjust = 1)) +
  facet_wrap(~metric, nrow = 1, scales = 'free_x') +
  scale_fill_manual(values = colors_response2) +
  scale_pattern_manual(values = c(Pre = 'stripe', Mid = 'none')) +
  scale_y_discrete(limits = rev,
                   drop = FALSE,
                   labels = rev(patient_labels)) +
  guides(pattern = guide_legend(override.aes = list(fill = 'white')),
         fill = guide_legend(override.aes = list(pattern = 'none'))) +
    ggtitle('IL17 Secreting Trm Cells')
  
  
  ggsave(plot = p,
         device = cairo_pdf,
         path = here(output_dir, 'patient_bar_supercluster'),
         filename = 'il17.pdf',
         h = 4,
         w = 4)
  
  p
  

```

### Stats
```{r}
which_subset <- 'Trm'
stats_input <- supercluster_tallies %>% 
  filter(treatment != 'None') %>% 
  ungroup() %>% 
  select(patient, supercluster, treatment, group, n, f) %>% 
  pivot_wider(names_from = treatment,
              values_from = c(n, f))

stats_input
```

### Abundance
```{r}
map(stats_input$supercluster %>% unique(), function(which_subset) {
  
  stats_subset <- stats_input %>% filter(supercluster == which_subset)
  
  group1 <- stats_subset %>% filter(group == 'Responder')
  group2 <- stats_subset %>% filter(group == 'Non-responder')
  
  res1 <- wilcox.test(group1$n_Pre,
                     group1$n_Mid)
  
  res2 <- wilcox.test(group2$n_Pre,
                      group2$n_Mid)
  
  tibble('cluster' = which_subset,
         'Responder_p' = res1$p.value,
         'Nonresponder_p' = res2$p.value)
  
}) %>% bind_rows() 
```

```{r}
map(stats_input$supercluster %>% unique(), function(which_subset) {
  
  stats_subset <- stats_input %>% filter(supercluster == which_subset)
  
  group1 <- stats_subset %>% filter(group == 'Responder')
  group2 <- stats_subset %>% filter(group == 'Non-responder')
  
  res1 <- wilcox.test(group1$f_Pre,
                     group1$f_Mid)
  
  res2 <- wilcox.test(group2$f_Pre,
                      group2$f_Mid)
  
  tibble('cluster' = which_subset,
         'Responder_p' = res1$p.value,
         'Nonresponder_p' = res2$p.value)
  
}) %>% bind_rows() 
```

#### Summary stats
```{r}
stats_input %>% 
  mutate(n_drop = 100 * (n_Mid - n_Pre) / (n_Pre)) %>% 
  group_by(supercluster, group) %>% 
  summarize(mean = mean(n_drop),
            median = median(n_drop),
            range_min = min(n_drop),
            range_max = max(n_drop))
```
```{r}
stats_input %>% 
  mutate(n_drop = 100 * (f_Mid - f_Pre) / (f_Pre)) %>% 
  group_by(supercluster, group) %>% 
  summarize(mean = mean(n_drop),
            median = median(n_drop),
            range_min = min(n_drop),
            range_max = max(n_drop))
```

```{r}


map(stats_input$supercluster %>% unique(), function(which_subset) {
  
  stats_subset <- stats_input %>% filter(supercluster == which_subset)
  
  group1 <- stats_subset %>% filter(group == 'Responder')
  group2 <- stats_subset %>% filter(group == 'Non-responder')
  
  res1 <- wilcox.test(group1$f_Pre,
                     group1$f_Mid)
  
  res2 <- wilcox.test(group2$f_Pre,
                      group2$f_Mid)
  
  tibble('cluster' = which_subset,
         'Responder_p' = res1$p.value,
         'Nonresponder_p' = res2$p.value)
  
}) %>% bind_rows() 
```


### UMAP
```{r}
p <- seuratobj %>% seurat_feature(features = c('IL17A', 'IL17F', 'IFNG', 'IL26'), nrow = 1)
p
```

```{r}
p <- seuratobj %>% 
  seurat_feature(features = 'IL17A',
                 facets = 'group')
p
```

### UMAP facets
```{r}
cytokines <- c('IL17A', 'IL17F', 'IFNG', 'IL26')
facet_input <- FetchData(seuratobj,
                         vars = c('UMAP_1', 'UMAP_2', 'id',cytokines), 
                         slot = 'data') %>% 
  inner_join(metadata_patient) %>% 
  mutate(description = paste(group, treatment)) 
  

facet_input
```
  
```{r}
which_gene <- 'IL17A'
tallies <- facet_input %>% group_by(group) %>% tally() %>% pull(n) %>% min()

set.seed(1234)

library(ggrastr)

p <- map(cytokines, function(which_gene) {
  
  plot_input <- facet_input %>% sample_n(size = tallies)
  plot_input$expression <- plot_input[[which_gene]]
  zero_input <- plot_input %>% filter(expression == 0)
  nonzero_input <- plot_input %>% filter(expression > 0) %>% mutate(expression = log10(2^expression))
  
  total_cpm <- nonzero_input %>% group_by(group, treatment) %>% summarize(total = round(sum(10^expression), 2))
  
  p <- nonzero_input %>% 
    ggplot(aes(x = UMAP_1,
               y = UMAP_2,
               color = expression)) +
    rasterize(geom_point(data = zero_input,
               color = 'grey90',
               size = 1), dpi = 300) +
    rasterize(geom_point(alpha = 0.5,
               size = 1), dpi = 300) +
    facet_wrap(~group + treatment, nrow = 1) +
    labs(x = 'UMAP1',
         y = 'UMAP2',
         title = which_gene) +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          legend.title = element_text(),
          legend.position = 'right',
          plot.title = element_text(face = 'italic'),
          axis.title.x = element_text(hjust = 0),
          axis.title.y = element_text(hjust = 0),
          strip.background = element_rect(color = 'black', fill = 'grey50')) +
    geom_text(data = total_cpm,
              x = 0,
              y = 8, 
              inherit.aes = FALSE,
              size = 3,
              color = 'black',
              aes(label = paste0('Total CPM = ', total))) +
    scale_color_gradientn(colors = colorRampPalette(rev(RColorBrewer::brewer.pal(n = 7, name = "OrRd")))(100),
                          limits = c(0, 2),
                          breaks = seq(0, 2, length.out = 3),
                          labels = c(1, 10, 100),
                          oob = scales::squish,
                          guide = guide_colorbar(frame.colour = 'black', 
                                                 ticks = FALSE, 
                                                 nbin = 100,
                                                 barwidth = 0.75,
                                                 title = 'CPM',
                                                 barheight = 3))
  
  p
  
}) %>% wrap_plots(ncol = 1)

# ggsave(plot = p,
# 	device = cairo_pdf,
# 	path = output_dir,
#         h = 12,
#         w = 8,
# 	filename = 'F4B_umap.pdf')
p

```




### Per cell level analysis
```{r}
set.seed(1234)

per_cell_df <- map(cytokines, function(which_gene) {
  
  plot_input <- facet_input %>% sample_n(size = tallies)
  plot_input$expression <- plot_input[[which_gene]]
  zero_input <- plot_input %>% filter(expression == 0)
  
  nonzero_input <- plot_input %>% filter(expression > 0) %>% mutate(expression = log10(2^expression))
  total_cpm <- nonzero_input %>% group_by(group, treatment) %>% summarize(total_expression = round(sum(10^expression), 2),
                                                                          cells = n(),
                                                                          per_cell_expression = round(mean(10^expression), 2)) 
  total_cpm %>% mutate(gene = which_gene)
  
  

  }) %>% bind_rows()

per_cell_df %>% write_tsv(here(output_dir, 'per_cell_table.tsv'))
per_cell_df
```

### Th2
```{r}
genes_th1th2 <- c("IL4R",
                  "CD40LG",
                  "IFNGR1",
                  "IL2RA",
                  "IL4R",
                  "IL2",
                  "IL4",
                  "IFNG",
                  "IL12A",
                  "CD86",
                  "CD28",
                  "IL18",
                  "HLA-DRB1",
                  "CD40",
                  "IL4R",
                  "IL12RB2",
                  "IL18",
                  "HLA-DRB1",
                  "HLA-DRB5",
                  "IL12B",
                  "IL18R1",
                  "NR1I2",
                  "IFNGR2",
                  "IL12RB1",
                  "CD28",
                  "CD86",
                  "HLA-DRA",
                  "HLA-DRB4",
                  "NR1I2",
                  "HLA-DRB3",
                  "NR1I2",
                  "CD40",
                  "IL12RB1",
                  "IL4",
                  "CD86")
```

### UMAP facets
```{r}
cytokines <- genes_th1th2 %>% unique()
facet_input <- FetchData(seuratobj,
                         vars = c('UMAP_1', 'UMAP_2', 'id',cytokines), 
                         slot = 'data') %>% 
  inner_join(metadata_patient) %>% 
  mutate(description = paste(group, treatment)) 
  

facet_input
```
  
```{r}
which_gene <- 'IL17A'
tallies <- facet_input %>% group_by(group) %>% tally() %>% pull(n) %>% min()

set.seed(1234)

library(ggrastr)

p <- map(cytokines %>% intersect(colnames(facet_input)), function(which_gene) {
  
  print2(which_gene)
  plot_input <- facet_input %>% sample_n(size = tallies)
  plot_input$expression <- plot_input[[which_gene]]
  zero_input <- plot_input %>% filter(expression == 0)
  nonzero_input <- plot_input %>% filter(expression > 0) %>% mutate(expression = log10(2^expression))
  
  total_cpm <- nonzero_input %>% group_by(group, treatment) %>% summarize(total = round(sum(10^expression), 2))
  
  p <- nonzero_input %>% 
    ggplot(aes(x = UMAP_1,
               y = UMAP_2,
               color = expression)) +
    rasterize(geom_point(data = zero_input,
               color = 'grey90',
               size = 1), dpi = 300) +
    rasterize(geom_point(alpha = 0.5,
               size = 1), dpi = 300) +
    facet_wrap(~group + treatment, nrow = 1) +
    labs(x = 'UMAP1',
         y = 'UMAP2',
         title = which_gene) +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          legend.title = element_text(),
          legend.position = 'right',
          plot.title = element_text(face = 'italic'),
          axis.title.x = element_text(hjust = 0),
          axis.title.y = element_text(hjust = 0),
          strip.background = element_rect(color = 'black', fill = 'grey50')) +
    geom_text(data = total_cpm,
              x = 0,
              y = 8, 
              inherit.aes = FALSE,
              size = 3,
              color = 'black',
              aes(label = paste0('Total CPM = ', total))) +
    scale_color_gradientn(colors = colorRampPalette(rev(RColorBrewer::brewer.pal(n = 7, name = "OrRd")))(100),
                          limits = c(0, 2),
                          breaks = seq(0, 2, length.out = 3),
                          labels = c(1, 10, 100),
                          oob = scales::squish,
                          guide = guide_colorbar(frame.colour = 'black', 
                                                 ticks = FALSE, 
                                                 nbin = 100,
                                                 barwidth = 0.75,
                                                 title = 'CPM',
                                                 barheight = 3))
  
  
  ggsave(plot = p,
         device = cairo_pdf,
         path = output_dir,
         h = 4,
         w = 8,
         filename = paste0('F4B_umap_th1th2_', which_gene, '.pdf'))
  
  p
}) %>% wrap_plots(ncol = 1)

# ggsave(plot = p,
# 	device = cairo_pdf,
# 	path = output_dir,
#         h = 12,
#         w = 8,
# 	filename = 'F4B_umap_th1th2.pdf')

p

```
