---
title: "Figrue 2"
author: "David Wu"
---

Set working directory to project directory
```{r setup}
require(knitr)
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

### Optionally convert this Rmd to R script 
```{r eval=FALSE, message=FALSE}
current_file <- rstudioapi::getActiveDocumentContext()$path
output_file <- stringr::str_replace(current_file, '.Rmd', '.R')
knitr::purl(current_file, output = output_file)
file.edit(output_file)
```

### Directories
```{r, message=FALSE}
output_dir <- 'figures' # analysis file output directory
data_dir <- 'data/derived/seurat' # data file output directory

dir.create(output_dir, showWarnings = FALSE)
dir.create(data_dir, showWarnings = FALSE)
```

### Loading the libraries
```{r, message=FALSE}
library(Seurat) 
library(tidyverse)
library(ggExtra)
library(ggthemes)
library(ggforce)
#library(SeuratWrappers) # devtools::install_github('satijalab/seurat-wrappers')
library(wutilities) # devtools::install_github('symbiologist/wutilities')
library(tictoc)
library(rcartocolor)
library(ggrepel)
theme_set(theme_dwu()) # set default theme

```

### Import
```{r}
seuratobj <- read_rds(file.path(data_dir, 'seuratobj_annotated.rds'))
meta <- seuratobj@meta.data %>% rownames_to_column() 
de_combined <- read_tsv('analysis/output/07_de/de_combined.tsv.gz')
seuratobj
```


```{r}
patient_meta <- meta %>% 
  select(id, patient, treatment, condition, response, response_group) %>% 
  unique() %>% 
  pivot_wider(names_from = treatment,
              values_from = id) %>% 
  mutate(patient_display = paste0(response_group, ': ', patient, ' (', Pre,', ', Mid, ')') %>% str_replace('Responder', ' Responder'),
         patient_display = factor(patient_display, levels = sort(patient_display))) %>% 
  arrange(patient_display)

patient_meta
```
```{r}
analyze <- meta %>% 
  select(rowname, id, sample, patient, condition, treatment, cluster, merged_label, response_group) %>% 
  mutate(description = factor(paste(condition, treatment) %>% str_remove(' None'),
                              levels = c('PV Pre',
                                         'PV Mid',
                                         'PV',
                                         'AD',
                                         'HC')),
         sample_display = paste0(sample, ' (', id, '), ', response_group)) %>% 
  left_join(patient_meta %>% select(patient, patient_display))

analyze

```


```{r}
tallies <- analyze %>% 
  group_by(patient, condition, treatment, merged_label) %>% 
  tally()

tallies %>%
  filter(condition == 'PV',
         treatment != 'None',
         merged_label == '18 moDC2') %>% 
  pivot_wider(names_from = treatment,
              values_from = n)
```


#### Global
```{r}
h <- 3
w <- 3
dpi <- 600

which_comparison <- c('PV vs HC',
                      'Pre vs Mid',
                      'Responder Pre vs Responder Mid',
                      'Non-responder Pre vs Non-responder Mid')

p_adj_cutoff <- 0.05

de_subset <- de_combined %>% 
  filter(comparison %in% which_comparison) %>% 
  mutate(comparison = factor(comparison, levels = which_comparison),
         sig = ifelse(p_val_adj < 0.05, TRUE, FALSE))
```

### Figure 2A
#### How many DE genes?
#### Facet

```{r}
p <- de_subset %>% 
  filter(p_val_adj < p_adj_cutoff) %>% 
  mutate(comparison = factor(comparison, levels = which_comparison)) %>% 
  group_by(cluster, comparison) %>% 
  tally() %>% 
  ggplot(aes(x = cluster,
             y = n,
             fill = comparison)) +
  geom_col(color = 'black') +
  facet_wrap(~comparison, scales = 'free_y') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  #coord_flip() +
  labs(x = '',
       y = '# of DEGs')

p
```


### Figure 2B
#### Volcano PV vs HC
```{r}
which_cluster <- 'Trm_merge'

de_subset %>% 
  filter(cluster == which_cluster) %>% 
  ggplot(aes(x = avg_log2FC,
             y = -log10(p_val_adj),
             color = sig,
             label = gene)) +
  geom_point(alpha = 0.3) +
  facet_wrap(~comparison, scales = 'free_y') +
  scale_color_manual(values = c('grey50', 'darkorchid4')) +
  geom_text_repel(data = de_subset %>% filter(cluster == which_cluster) %>% group_by(comparison) %>% top_n(10, abs(avg_log2FC)), box.padding = 0.5) +
  labs(x = 'log2 fold-change',
       y = '-log10 adj. p-value')
  
```
### Figure 2C
What is corrected by treatment?

Import gene sets
```{r}
genesets <- read_tsv('analysis/output/msigdb/il23_il17_gene_sets.tsv')
```

```{r}
response_genes <- de_combined %>% 
  filter(comparison == 'Responder Pre vs Responder Mid') %>% 
  filter(p_val_adj < p_adj_cutoff,
         avg_log2FC > 0,
         str_detect(gene, 'mm10', negate = TRUE)) %>% 
  top_n(10, wt = abs(avg_log2FC)) %>% 
  pull(gene) %>% 
  c('IL17A', 'IL17F', 'CXCL13')

response_genes
```
### Extract expression
```{r}

genes_to_extract <- genesets$gene_symbol %>% unique()

metadata_subset <- meta %>% 
  filter(response_group == 'Responder' | condition == 'HC',
         merged_label == 'Trm_merge') %>% 
  mutate(treatment = ifelse(treatment == 'None', 'Healthy', treatment),
         treatment = factor(treatment, levels = c('Pre', 'Mid', 'Healthy')))

barcodes <- metadata_subset$rowname
exprs <- FetchData(seuratobj, 
                   cells = barcodes,
                   vars = genes_to_extract) %>% 
  rownames_to_column() %>% 
  as_tibble()

exprs

```

```{r}
merged_table <- exprs %>% 
  inner_join(metadata_subset %>% select(rowname, patient, id, condition, treatment, response, response_group, merged_label)) %>% 
  pivot_longer(cols = genes_to_extract,
               names_to = 'gene',
               values_to = 'expression')

merged_table
```
```{r}
p <- merged_table %>% 
  ggplot(aes(x = treatment,
             y = expression,
             color = treatment)) +
  geom_boxplot(outlier.shape = NA) +
  #geom_sina(alpha = 0.1, size = 0.5) 
  facet_wrap(~gene, scales = 'free_y') +
  scale_color_carto_d() +
  theme(legend.position = 'top')

p
```

```{r}
summarized_table <- merged_table %>% 
  group_by(treatment, response_group, merged_label, gene) %>% 
  summarize(mean = mean(expression),
            sd = sd(expression))
```



```{r}
p <- summarized_table %>% 
   ggplot(aes(x = treatment,
              y = mean,
              fill = treatment)) +
  geom_col(color = 'black') +
  facet_wrap(~gene, scales = 'free_y') +
  scale_fill_carto_d() +
  #theme(legend.position = 'top') +
  labs(x = '',
       y = 'Average Expression in Trm Cells')

p
```
### Biocarta
```{r}
p <- summarized_table %>% 
  filter(gene %in% (genesets %>% filter(gs_name == 'BIOCARTA_IL17_PATHWAY') %>% pull(gene_symbol))) %>% 
   ggplot(aes(x = treatment,
              y = mean,
              fill = treatment)) +
  geom_col(color = 'black') +
  facet_wrap(~gene, scales = 'free_y') +
  scale_fill_carto_d() +
  #theme(legend.position = 'top') +
  labs(x = '',
       y = 'Average Expression in Trm Cells',
       title = 'Biocarta IL17 Pathway')

p
```
### Wikipathway
```{r}
p <- summarized_table %>% 
  filter(gene %in% (genesets %>% filter(gs_name == 'WP_IL17_SIGNALING_PATHWAY') %>% pull(gene_symbol))) %>% 
   ggplot(aes(x = treatment,
              y = mean,
              fill = treatment)) +
  geom_col(color = 'black') +
  facet_wrap(~gene, scales = 'free_y') +
  scale_fill_carto_d() +
  #theme(legend.position = 'top') +
  labs(x = '',
       y = 'Average Expression in Trm Cells',
       title = 'Wikipathway IL17 Signaling Pathway')

p
```
### PID
```{r}
p <- summarized_table %>% 
  filter(gene %in% (genesets %>% filter(gs_name == 'PID_IL23_PATHWAY') %>% pull(gene_symbol))) %>% 
   ggplot(aes(x = treatment,
              y = mean,
              fill = treatment)) +
  geom_col(color = 'black') +
  facet_wrap(~gene, scales = 'free_y') +
  scale_fill_carto_d() +
  #theme(legend.position = 'top') +
  labs(x = '',
       y = 'Average Expression in Trm Cells',
       title = 'Pathway Interaction Database IL23 Pathway')

p
```

```{r}
shared_pathway_genes <- genesets %>% 
  select(gs_name, gene_symbol) %>% 
  unique() %>% 
  group_by(gene_symbol) %>% 
  tally() 
```

```{r}
p <- summarized_table %>% 
  filter(gene %in% (shared_pathway_genes %>% filter(n > 1) %>% pull(gene_symbol))) %>% 
   ggplot(aes(x = treatment,
              y = mean,
              fill = treatment)) +
  geom_col(color = 'black') +
  facet_wrap(~gene, scales = 'free_y') +
  scale_fill_carto_d() +
  #theme(legend.position = 'top') +
  labs(x = '',
       y = 'Average Expression in Trm Cells',
       title = 'Il23/IL17 Robust Gene Set')

p
```

### clusterProfiler
```{r}

```

```{r}
summarized_table %>% 
  filter(treatment != 'Healthy') %>% 
  ggplot(aes(x = reorder(gene, -mean),
             y = mean,
             color = treatment,
             group = gene)) +
  geom_point(size = 3) +
  geom_line(color = 'grey50', arrow = arrow(length = unit(0.2, "cm"), type = 'closed')) +
  theme_dwu(legend.position = 'right') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_color_manual(values = c('orangered', 'dodgerblue4')) +
  #geom_point(size = 2, data = summarized_table %>% filter(treatment == 'Healthy'), inherit.aes = FALSE, aes(x = reorder(gene, -mean), y = mean)) +
  labs(x = '',
       y = 'Average Trm Expression')
```
### Waterfall

```{r}
ray_genes <- c("IL17F", "GNLY", "LRRN3", "MGAT4A", "CHN1", "CTSH", "PTPN13", "GZMB", "ADGRG1", "SOX4", "GABPB1-AS1", "SNX9", "CD7", "CPM", "CTLA4", "ODF2L", "CBLB", "LAYN", "TNFAIP3", "ARHGEF12", "H1FX", "HIST1H1E", "YPEL2", "GBP5", "DAPK2", "RAP1B", "CXCL13", "FYN")

genes_to_extract <- ray_genes

metadata_subset <- analyze %>% 
  filter(response_group %in% c('Responder', 'Non-responder') | condition == 'HC',
         merged_label == 'Trm_merge') %>% 
  mutate(treatment = ifelse(treatment == 'None', 'Healthy', treatment),
         treatment = factor(treatment, levels = c('Pre', 'Mid', 'Healthy')))

barcodes <- metadata_subset$rowname
exprs <- FetchData(seuratobj, 
                   cells = barcodes,
                   vars = genes_to_extract) %>% 
  rownames_to_column() %>% 
  as_tibble()

exprs




```

```{r}
merged_table <- exprs %>% 
  inner_join(metadata_subset %>% select(rowname, patient, id, condition, treatment, response_group, merged_label, patient_display)) %>% 
  pivot_longer(cols = genes_to_extract,
               names_to = 'gene',
               values_to = 'expression')

merged_table
```


```{r}
p <- merged_table %>% 
  filter(response_group == 'Responder' | treatment == 'Healthy') %>% 
  ggplot(aes(x = treatment,
             y = expression,
             fill = treatment)) +
  geom_boxplot(outlier.shape = NA, color = 'black', alpha = 0.8) +
  #geom_sina(alpha = 0.1, size = 0.5) 
  facet_wrap(~gene, scales = 'free_y') +
  #scale_color_carto_d() +
  scale_fill_manual(values = c('orangered', 'dodgerblue4', 'grey80')) +
  theme(legend.position = 'top') +
  labs(x = '',
       y = 'Average Trm Expression')

p
```
```{r}
summarized_table <- merged_table %>% 
  filter(response_group != 'Non-responder') %>% 
  group_by(treatment, response_group, merged_label, gene) %>% 
  summarize(mean = mean(expression),
            median = median(expression),
            sd = sd(expression))
```

```{r}
plot_input <- merged_table %>% 
  group_by(patient, treatment, response_group, merged_label, gene, patient_display) %>% 
  summarize(mean = mean(expression)) 

p <- plot_input %>% 
  filter(treatment != 'Healthy') %>% 
  ggplot(aes(x = reorder(gene, -mean),
             y = mean,
             color = treatment,
             group = gene)) +
  geom_point(size = 3) +
  geom_line(color = 'grey50', arrow = arrow(length = unit(0.2, "cm"), type = 'closed')) +
  theme_dwu(legend.position = 'right') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  facet_wrap(~patient_display, scales = 'free_y', ncol = 4) +
  scale_color_manual(values = c('orangered', 'dodgerblue4')) +
  # geom_point(size = 2,
  #            data = plot_input %>% filter(treatment == 'Healthy') %>% 
  #              group_by(gene) %>% 
  #              summarize(mean = mean(mean)),
  #            inherit.aes = FALSE,
  #            aes(x = reorder(gene, -mean), y = mean),
  #            shape = 1) +
  labs(x = '',
       y = 'Average Trm Expression')


p
```


```{r}
summarized_table %>% 
  filter(treatment != 'Healthy') %>% 
  ggplot(aes(x = reorder(gene, -mean),
             y = mean,
             color = treatment,
             group = gene)) +
  geom_point(size = 3) +
  geom_line(color = 'grey50', arrow = arrow(length = unit(0.2, "cm"), type = 'closed')) +
  theme_dwu(legend.position = 'right') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_color_manual(values = c('orangered', 'dodgerblue4')) +
  geom_point(size = 2, 
             data = summarized_table %>% filter(treatment == 'Healthy'), 
             inherit.aes = FALSE, 
             aes(x = reorder(gene, -mean), y = mean),
             shape = 1
             ) +
  labs(x = '',
       y = 'Average Trm Expression')
```


```{r}
p <- summarized_table %>% 
  ggplot(aes(x = treatment,
              y = mean,
              fill = treatment)) +
  geom_bar(stat = 'identity', color = 'black', alpha = 0.8) + 
  facet_wrap(~gene, scales = 'free_y') +
  scale_color_carto_d() +
  #theme(legend.position = 'top') +
  scale_fill_manual(values = c('orangered', 'dodgerblue4', 'grey80')) +
  labs(x = '',
       y = 'Average Expression in Trm Cells')

p
  
```


### Figure
Top genes from PV vs HC
```{r}
pv_vs_hc <- de_combined %>% 
  filter(comparison == 'PV vs HC')

pv_vs_hc
```

```{r}
top_pv_vs_hc <- pv_vs_hc %>% 
  filter(avg_log2FC > 0,
         cluster == which_cluster) %>% 
  top_n(10, pct.diff) %>%
  pull(gene)

top_pv_vs_hc
```


#### Which genes normalize?
```{r}
de_normalize <- de_combined %>% 
  filter(comparison %in% c('PV vs HC', 'Pre vs Mid'),
         cluster == which_cluster) %>% 
  select(comparison, gene, avg_log2FC, pct.1, pct.2, pct.diff, p_val_adj) %>% 
  pivot_wider(names_from = comparison,
              values_from = avg_log2FC:p_val_adj) %>% 
  mutate(avg_log2FC = ifelse(`avg_log2FC_PV vs HC` * `avg_log2FC_Pre vs Mid` > 0, TRUE, FALSE))

de_normalize
```

```{r}
de_normalize %>% 
  filter(avg_log2FC)
```



### Extract expression
```{r}
genes_to_extract <- de_normalize$gene

metadata_subset <- meta %>% 
  filter(response_group == 'Responder' | condition == 'HC',
         merged_label == 'Trm_merge') %>% 
  mutate(treatment = ifelse(treatment == 'None', 'Healthy', treatment),
         treatment = factor(treatment, levels = c('Pre', 'Mid', 'Healthy')))

barcodes <- metadata_subset$rowname
exprs <- FetchData(seuratobj, 
                   cells = barcodes,
                   vars = genes_to_extract) %>% 
  rownames_to_column() %>% 
  as_tibble()

exprs

```

```{r}
merged_table <- exprs %>% 
  inner_join(metadata_subset %>% select(rowname, patient, id, condition, treatment, response, response_group, merged_label)) %>% 
  pivot_longer(cols = genes_to_extract,
               names_to = 'gene',
               values_to = 'expression')

merged_table
```
```{r}
summarized_table <- merged_table %>% 
  group_by(patient, treatment, response_group, merged_label, gene) %>% 
  summarize(median = median(expression),
            mean = mean(expression),
            sd = sd(expression))

summarized_table
```

```{r}
summarized_table %>% 
  filter(gene %in% genes_to_extract[1:20]) %>% 
   ggplot(aes(x = treatment,
              y = mean,
              color = treatment)) +
  geom_bar(stat = 'summary', color = 'black', fill = 'white', fun = 'mean') + 
  geom_sina() +
  facet_wrap(~gene, scales = 'free_y') +
  scale_color_carto_d() +
  #theme(legend.position = 'top') +
  labs(x = '',
       y = 'Average Expression in Trm Cells')
```
```{r}

```


## Session Info
```{r}
sessionInfo()
```



