---
title: "Figrue 1"
author: "David Wu"
---

### Use the here package for robust relative paths
```{r}
library(here)
```

### Optionally convert this Rmd to R script 
```{r eval=FALSE, message=FALSE}
current_file <- rstudioapi::getActiveDocumentContext()$path
output_file <- stringr::str_replace(current_file, '.Rmd', '.R')
knitr::purl(current_file, output = output_file)
file.edit(output_file)
```

### Directories
```{r, message=FALSE}
output_dir <- here('figures') # analysis file output directory
data_dir <- here('data/derived/tildra') # data file output directory

dir.create(output_dir, showWarnings = FALSE)
dir.create(data_dir, showWarnings = FALSE)
```

### Loading the libraries
```{r, message=FALSE}
library(Seurat) 
library(tidyverse)
library(ggExtra)
library(wutilities) # devtools::install_github('symbiologist/wutilities')
library(tictoc)
library(patchwork)
library(rcartocolor)
library(ggrastr)
library(extrafont)
theme_set(theme_dwu())
```

### Import seurat
```{r}
seuratobj <- read_rds(file.path(data_dir, 'seuratobj_annotated.rds'))
seuratobj
```

### Select identity
```{r}
which_annotation <- 'global_supercluster'
seuratobj$annotation <- seuratobj[[which_annotation]]
Idents(seuratobj) <- 'annotation'

output_dir <- here(output_dir, which_annotation)
dir.create(output_dir)
```

```{r}
seuratobj$condition <- factor(seuratobj$condition, levels = c('HC', 'PV'))
seuratobj$treatment <- factor(seuratobj$treatment, levels = c('None', 'Pre', 'Mid'))
seuratobj$group <- factor(seuratobj$group, levels = c('Responder', 'Non-responder'))
metadata <- seuratobj@meta.data %>% rownames_to_column() %>% as_tibble()
```

### Figure 1A
#### Global
```{r}
h <- 3
w <- 3
dpi <- 600
```

#### UMAP of points only (PNG)
```{r fig.height=4, fig.width=4, dpi=300}
p <- seurat_feature(seuratobj, features = 'annotation', facet_hide = TRUE, legend_position = 'none', label = FALSE, color_package = 'carto', color_palette = 'Bold') +
  theme(text = element_text(family = 'Arial'))

ggsave(plot = p,
       path = output_dir,
       height = h,
       width = w,
       dpi = dpi,
       filename = 'F1A_points.png') 

p
```
```{r}
p <- seurat_feature(seuratobj, features = 'annotation', facet_hide = TRUE, legend_position = 'none', label = TRUE, color_package = 'carto', color_palette = 'Bold', label_size = 3) +
  theme(text = element_text(family = 'Arial'))

ggsave(plot = p,
       path = output_dir,
       height = h,
       width = w,
       dpi = dpi,
       filename = 'F1A_labeled.png') 

p
```

#### UMAP of labels only
```{r}
p <- seurat_feature(seuratobj, features = 'annotation', facet_hide = TRUE, legend_position = 'none', label = TRUE, drop_points = TRUE, label_color = c('black', NA)) +
  theme(text = element_text(family = 'Arial'))

ggsave(plot = p,
       path = output_dir,
       height = h,
       width = w,
       dpi = dpi,
       filename = 'F1A_labels.pdf') 

p
```
```{r}
DimPlot(seuratobj, group.by = 'annotation', label = TRUE)
```

#### F1A legend
```{r fig.height = 4, fig.width=4}
p <- seurat_feature(seuratobj, features = 'annotation', facet_hide = TRUE, legend_position = 'right', label = FALSE) +
  theme(text = element_text(family = 'Arial'))

ggsave(plot = p,
       path = output_dir,
       height = h,
       width = w,
       dpi = dpi,
       filename = 'F1A_legend.pdf') 

p
```
### Figure 1B, Dot Plot
#### Reynolds et al markers

```{r}
markers_reynolds <- c('KRT14', 'KRT5', 
                      'GATA3', 'KRTDAP',
                      'CRYAB', 'TYRP1',
                      'DKK3', 'PLEKHA4',
                      'LUM', 'APOD',
                      'PHLDA2', 'CCL2',
                      'IL6', 'PDGFRA',
                      'COL1A1', 'CXCL12',
                      'CD82', 'CCL19', 'KCNE4', 'CPE', 'TAGLN', 'MYL9', 'HEY1', 'CCL14', 'PECAM1',
                      'ACKR1', 'SELE', 'SNCG', 'HES1', 'MMRN1', 'XCL1' ,'XCL2', 'KLRC1', 'CCL5', 'GZMA', 
                      'CD8A', 'CD8B', 'IL7R', 'FOXP3', 'TIGIT', 'TPSB2', 'TPSAB1', 'CD79A', 'JCHAIN', 'CD163',
                      'C1QB', 'FCGR2A', 'MS4A6A', 'IL23A', 'CLEC9A', 'CLEC10A', 'CD1C', 'CD207', 'CD1A', 
                      'CD14', 'IL1B', 'CCR7', 'IDO1')

main_markers <- c("CD3D",
                  "CCR7",
                  "SELL",
                  "KLF2","CD69","ITGAE","CXCR6","CD4","TIGIT","FOXP3",
                  "IL2RA","CTLA4","CD8A","CD8B","GZMB","PDCD1","LAG3","KLRB1","PRF1","KLRD1","GNLY",
                  "TNFRSF18","PRDM1","BATF","TRAT1","RORA","GATA3","PTGDR2",
                  "IL7R", "HLA-DRA","HLA-DRB1",
                  "CD83","IDO1","CD207","EPCAM","CD68","C1QB","C1QC","CD163","CLEC10A","CD1C",
                  "THBD","XCR1","SIRPA","F13A1","IGKC","JCHAIN","CD79A","MS4A1","NR4A1","NR4A2","KLF4",
                  "CEBPB","LYZ","MS4A7","SERPINA1","CD14","S100A9","IL23A","TPSB2","TPSAB1","MKI67","TOP2A",
                  "ITGA4","NCR1","IL17A","IL17F","IL23R")

main_markers <- c("CD3D",
                  "CCR7",
                  "SELL",
                  "KLF2",
                  "CD69",
                  "ITGAE",
                  "CXCR6",
                  "CD4",
                  "TIGIT",
                  "FOXP3",
                  "IL2RA",
                  "CTLA4",
                  "CD8A",
                  "CD8B",
                  "GZMB",
                  "PDCD1",
                  #"LAG3",
                  "KLRB1",
                  "PRF1",
                  "KLRD1",
                  "GNLY",
                  "TNFRSF18",
                  "PRDM1",
                  "BATF",
                  "TRAT1",
                  "RORA",
                  "GATA3",
                  "PTGDR2",
                  "IL7R", 
                  "HLA-DRA",
                  "CD83",
                  #"IDO1",
                  "CD207",
                  #"EPCAM",
                  "CD68",
                  "C1QB",
                  #"C1QC",
                  "CD163",
                  #"CLEC10A",
                  "CD1C",
                  "THBD",
                  #"XCR1",
                  "SIRPA",
                  #"F13A1",
                  #"IGKC",
                  "JCHAIN",
                  "CD79A",
                  "MS4A1",
                  "NR4A1",
                  #"NR4A2",
                  #"KLF4",
                  "CEBPB",
                  "LYZ",
                  "MS4A7",
                  #"SERPINA1",
                  "CD14",
                  #"S100A9",
                  
                  #"TPSB2",
                  #"TPSAB1",
                  #"ITGA4"
                  "IL23A"
                  )
```



#### F1B dot plot horizontal
```{r}
p <- DotPlot(seuratobj, 
             features = main_markers,
             cluster.idents = TRUE,
             #cols = c('dodgerblue3', 'orangered'), 
             #cols = viridis::viridis(n = 2, option = 'C'),
             cols = RColorBrewer::brewer.pal(n = 7, 'BuPu')[c(1,7)],
             dot.min = 0.1,
             dot.scale = 4, 
             scale = FALSE) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, face = 'italic'))

ggsave(plot = p,
       path = output_dir,
       height = 5,
       width = 10,
       dpi = 300,
       filename = 'F1B.pdf') 

p
```
```{r}
p <- DotPlot(seuratobj, 
             features = main_markers, 
             cluster.idents = TRUE,
             #cols = c('dodgerblue3', 'orangered'), 
             #cols = viridis::viridis(n = 2, option = 'B'),
             cols = RColorBrewer::brewer.pal(n = 7, 'BuPu')[c(1,7)],
             dot.min = 0.1,
             dot.scale = 4, 
             scale = FALSE) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.text.y = element_text(face = 'italic')) +
  coord_flip()

ggsave(plot = p,
       path = output_dir,
       height = 8,
       width = 8,
       dpi = 600,
       filename = 'F1B_vertical.pdf') 

p
```

### Figure 1C


```{r}
color_palette5 <- rcartocolor::carto_pal(n = 5, name = 'Bold')
unit_height <- 2.2
unit_width <- 2
```


```{r}
p <- seurat_feature(seuratobj, features = 'condition', facets = 'condition', label = FALSE, color_package = 'custom', color_palette = color_palette5[5:3], legend_position = 'none', axis_title_position = 'none')

ggsave(plot = p,
       path = output_dir,
       height = unit_height,
       width = unit_width * 3,
       dpi = 600,
       filename = 'F1C_condition.png') 

p
```
```{r}
p <- seurat_feature(seuratobj, features = 'treatment', facets = 'treatment', label = FALSE, nrow = 1, color_package = 'custom', color_palette = color_palette5[c(5,1,2)], legend_position = 'none', axis_title_position = 'none')

ggsave(plot = p,
       path = output_dir,
       height = unit_height,
       width = unit_width * 3,
       dpi = 600,
       filename = 'F1C_treatment.png') 

p
```


### Figure 1D
```{r}
patient_order <- metadata %>% select(patient, sample, condition, treatment, response, group, ends_with('label')) %>% unique()

tildra_order <- patient_order %>% filter(treatment != 'None') %>% 
  arrange(patient, treatment)

tildra_order
```
### Cluster size and % 
```{r}
tallies <- metadata %>% 
  group_by(annotation) %>% 
  tally() %>% 
  mutate(total = sum(n),
         f = n/total,
         pct = round(100 * f, 2))
  
tallies
```

### Proportion of cells in each cluster
```{r}
analyze <- metadata %>% 
  select(rowname, id, patient, sample, condition, treatment, annotation, celltype, group) %>% 
  mutate(description = factor(paste(condition, treatment) %>% str_remove(' None'),
                              levels = c('PV Pre',
                                         'PV Mid',
                                         'PV',
                                         'AD',
                                         'HC')))

analyze %>% head()
```

```{r}
plot_input <- analyze %>% 
  group_by(description, annotation) %>% 
  tally() %>% 
  group_by(annotation) %>% 
  mutate(total = sum(n),
         fraction = n/total)

plot_input
```

```{r}
p <- plot_input %>% 
  ggplot(aes(x = annotation,
             y = fraction,
             fill = description)) +
  geom_bar(stat = 'identity',
           position = 'stack') +
  coord_flip() +
  rcartocolor::scale_fill_carto_d(palette = 'Bold') +
  labs(x = '',
       y = 'Proportion') +
  theme_dwu(legend.position = 'top')

ggsave(plot = p,
       filename = 'F1D_bar_percentages.pdf',
       path = output_dir,
       h = 6,
       w = 8)

p
```
### Wide
```{r}
p <- plot_input %>% 
  ggplot(aes(x = annotation,
             y = fraction * 100,
             fill = description)) +
  geom_bar(stat = 'identity',
           position = 'stack') +
  rcartocolor::scale_fill_carto_d(palette = 'Bold') +
  labs(x = '',
       y = '% of Cluster') +
  theme_dwu(legend.position = 'top') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

ggsave(plot = p,
       filename = 'F1D_bar_percentages_wide.pdf',
       path = output_dir,
       h = 4,
       w = 8)

p
```



#### Patient-centric versions
```{r}
plot_input <- analyze %>% 
  filter(description %in% c('PV Pre', 'PV Mid')) %>% 
  group_by(sample, patient, description, annotation, treatment, group) %>% 
  tally() %>% 
  group_by(sample) %>% 
  mutate(total = sum(n),
         fraction = n/total) %>% 
  ungroup() %>% 
  select(-sample, -n, -description, -total) %>% 
  pivot_wider(names_from = treatment,
              values_from = fraction) %>% 
  mutate(ratio = Mid/Pre,
         log2ratio = log2(ratio))

plot_input

```
### Boxplots
```{r}
p <- plot_input %>% 
  ggplot(aes(x = group,
             y = ratio,
             fill = annotation)) +
  geom_boxplot() +
  facet_wrap(~annotation, scales = 'free_y') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(plot = p,
       path = output_dir,
       filename = 'box_f_ratios.pdf',
       h = 8,
       w = 8)

p
```
```{r}
plot_input <- analyze %>% 
  filter(description %in% c('PV Pre', 'PV Mid')) %>% 
  group_by(sample, patient, description, annotation, treatment, group) %>% 
  tally() %>% 
  group_by(sample) %>% 
  mutate(total = sum(n),
         fraction = n/total) %>% 
  ungroup() %>% 
  select(-sample, -fraction, -description, -total) %>% 
  pivot_wider(names_from = treatment,
              values_from = n) %>% 
  mutate(ratio = Mid/Pre,
         log2ratio = log2(ratio))

plot_input

```
### Boxplots
```{r}
p <- plot_input %>% 
  ggplot(aes(x = group,
             y = ratio,
             fill = annotation)) +
  geom_boxplot() +
  facet_wrap(~annotation, scales = 'free_y') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(plot = p,
       path = output_dir,
       filename = 'box_n_ratios.pdf',
       h = 8,
       w = 8)

p
```


#### Individual
```{r}
plot_input <- analyze %>% 
  filter(description %in% c('PV Pre', 'PV Mid')) %>% 
  group_by(sample, patient, description, annotation, celltype, treatment, group) %>% 
  tally() %>% 
  group_by(sample) %>% 
  mutate(total = sum(n),
         fraction = n/total,
         sample = factor(sample, levels = rev(tildra_order$sample)))

plot_input

```

```{r}
patient_info <- plot_input %>% select(sample, patient, description, group) %>% unique() %>% arrange(patient, description)
text_color <- ifelse(patient_info$group == 'Responder', 'dodgerblue4', 'orangered')
```


```{r}
p <- plot_input %>% 
  #mutate(sample = factor(sample, levels = rev(tildra_order$sample))) %>% 
  ggplot(aes(x = sample,
             y = fraction * 100,
             fill = annotation)) +
  geom_bar(stat = 'identity',
           position = 'stack') +
  #rcartocolor::scale_fill_carto_d(palette = 'Bold') +
  labs(x = '',
       y = '% of Cells') +
  theme_dwu(legend.position = 'top') +
  coord_flip() +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1, color = text_color))
  theme(axis.text.y = element_text(angle = 0, hjust = 1, color = rev(text_color)))

ggsave(plot = p,
       filename = 'F1D_bar_patient_centric_f.pdf',
       path = output_dir,
       h = 4,
       w = 8)

p
```
```{r}
p <- plot_input %>% 
  
    ggplot(aes(x = sample,
             y = n,
             fill = annotation)) +
  geom_bar(stat = 'identity',
           position = 'stack') +
  #rcartocolor::scale_fill_carto_d(palette = 'Bold') +
  labs(x = '',
       y = '# of Cells') +
  theme_dwu(legend.position = 'top') +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1, color = rev(text_color)))

ggsave(plot = p,
       filename = 'F1D_bar_patient_centric_n.pdf',
       path = output_dir,
       h = 4,
       w = 8)

p
```
### All facets
```{r}
p <- plot_input %>% 
  ggplot(aes(x = sample,
             y = fraction * 100,
             fill = annotation)) +
  geom_bar(stat = 'identity') +
  labs(x = '',
       y = '% of Cells') +
  theme_dwu(legend.position = 'top') +
  coord_flip() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1, color = rev(text_color))) +
  facet_wrap(~annotation, scales = 'free_x')

ggsave(plot = p,
       filename = 'S1X_bar_patient_centric_f_facet.pdf',
       path = output_dir,
       h = 12,
       w = 12)

p
```



```{r}
p <- plot_input %>% 
  
    ggplot(aes(x = sample,
             y = n,
             fill = annotation)) +
  geom_bar(stat = 'identity') +
  #rcartocolor::scale_fill_carto_d(palette = 'Bold') +
  labs(x = '',
       y = '# of Cells') +
  theme_dwu(legend.position = 'top') +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1, color = rev(text_color))) +
  facet_wrap(~annotation, scales = 'free_x')

ggsave(plot = p,
       filename = 'S1X_bar_patient_centric_n_facet.pdf',
       path = output_dir,
       h = 12,
       w = 12)

p
```
### Trm focused
```{r}
which_clusters <- str_subset(plot_input$annotation, 'Trm') %>% unique()
```

```{r}
p <- plot_input %>% 
  filter(annotation %in% which_clusters) %>% 
  ggplot(aes(x = sample,
             y = fraction * 100,
             fill = annotation)) +
  geom_bar(stat = 'identity') +
  labs(x = '',
       y = '% of Cells') +
  theme_dwu(legend.position = 'top') +
  coord_flip() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1, color = rev(text_color))) + 
  facet_wrap(~annotation, scales = 'free_x')

ggsave(plot = p,
       filename = 'F1D_bar_patient_centric_f_trm.pdf',
       path = output_dir,
       h = 4,
       w = 4)

p
```

```{r}
p <- plot_input %>% 
  filter(annotation %in% which_clusters) %>% 
    ggplot(aes(x = sample,
             y = n,
             fill = annotation)) +
  geom_bar(stat = 'identity') +
  #rcartocolor::scale_fill_carto_d(palette = 'Bold') +
  labs(x = '',
       y = '# of Cells') +
  theme_dwu(legend.position = 'top') +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip() +
  theme(axis.text.y = element_text(angle = 0, hjust = 1, color = rev(text_color))) +
  facet_wrap(~annotation, scales = 'free_x')

ggsave(plot = p,
       filename = 'F1D_bar_patient_centric_n_trm.pdf',
       path = output_dir,
       h = 4,
       w = 4)

p
```
```{r}
p <- plot_input %>% 
  group_by(patient, treatment,celltype) %>% 
  summarize(n = sum(n)) %>% 
  #filter(annotation %in% which_clusters) %>% 
    ggplot(aes(x = treatment,
             y = n,
             fill = celltype)) +
  geom_boxplot() +
  #rcartocolor::scale_fill_carto_d(palette = 'Bold') +
  labs(x = '',
       y = '# of Cells') +
  facet_wrap(~celltype, scales = 'free_y') +
  #scale_y_log10() +
  theme_dwu(legend.position = 'top') +
  ggthemes::scale_fill_few()

ggsave(plot = p,
       filename = 'F1E_box_patient_centric_n_celltype.pdf',
       path = output_dir,
       h = 4,
       w = 4)

p
```
```{r}

```

```{r}
p <- plot_input %>% 
  filter(celltype == 'Lymphocyte') %>% 
    ggplot(aes(x = treatment,
             y = n,
             fill = annotation)) +
  geom_boxplot() +
  #rcartocolor::scale_fill_carto_d(palette = 'Bold') +
  labs(x = '',
       y = '# of Cells') +
  
  facet_wrap(~annotation, scales = 'free_y') +
theme_dwu(legend.position = 'top') 

ggsave(plot = p,
       filename = 'F1E_box_patient_centric_n_trm.pdf',
       path = output_dir,
       h = 4,
       w = 4)

p
```

## Session Info
```{r}
sessionInfo()
```



